---
title: Benchmarking mixeff
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking mixeff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-package}
library(mixeff)
```

Speed matters most when a workflow repeats the same fit many times.
Parametric bootstrap is the canonical case: a 1,000-replicate
bootstrap means roughly 1,000 refits, plus the original fit and
summary work. The same logic applies to cluster bootstrap,
simulation studies, multi-seed reanalysis, and anything whose inner
loop is "fit the same model again with new data". A small per-refit
speedup compounds into the difference between an overnight run and
a coffee break.

`mixeff` is roughly two to five times faster than `lme4` on the
parity benchmark — enough to make bootstrap a routine option rather
than a special-occasion calculation. This vignette shows the figures
from a single run of that benchmark and then documents the harness
used to make them. The numbers themselves are machine- and
model-dependent, so the scripts also write CSV files that you can
re-plot and rerun on the designs you actually care about.

## What does the benchmark show?

The benchmark sweeps five mixed-model designs — varying rows,
varying subject levels, adding random slopes, crossing subject and
item, and combining crossed grouping with a random slope — across a
log-spaced grid of scale values. At each cell, `mixeff::lmm()` and
`lme4::lmer()` fit the same model three times. The summary table
records median elapsed seconds and speedup over `lme4` per cell.

```{r benchmark-data}
benchmark_path <- system.file(
  "extdata", "lme4-scaling-summary.csv",
  package = "mixeff"
)
benchmark <- read.csv(benchmark_path, stringsAsFactors = FALSE)

benchmark$scenario_label <- factor(
  benchmark$scenario,
  levels = c("rows", "groups", "slopes", "crossed", "crossed_slope"),
  labels = c(
    "vary rows (1 | subject)",
    "vary subject levels (1 | subject)",
    "add random slope (1 + x | subject)",
    "crossed (1 | subject) + (1 | item)",
    "crossed + slope (1 + x | subject) + (1 | item)"
  )
)

head(benchmark[, c("scenario", "scale_value", "engine",
                   "median_sec", "speedup_vs_lme4")])
```

```{r benchmark-checks, include = FALSE}
stopifnot(nrow(benchmark) > 0L)
stopifnot(all(c("rows", "groups", "slopes", "crossed",
                "crossed_slope") %in% benchmark$scenario))
stopifnot(all(c("lme4", "mixeff") %in% benchmark$engine))
```

The first figure plots the median fit time against design scale, on
log-log axes, with one panel per scenario. `mixeff` runs below
`lme4` everywhere on the grid; the gap widens as design complexity
grows.

```{r fit-time-plot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 4.5, fig.alt = "Median fit time vs design scale, log-log, by scenario, for mixeff and lme4."}
library(ggplot2)

ggplot(benchmark,
       aes(scale_value, median_sec * 1000, colour = engine,
           group = engine)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ scenario_label, scales = "free_x", ncol = 2) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = c(lme4 = "#888888", mixeff = "#3366cc")) +
  labs(
    x = "design scale (rows, subject levels, etc.)",
    y = "median fit time (ms, log scale)",
    colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top",
        panel.grid.minor = element_blank())
```

The second figure folds the same data into one panel: the speedup
of `mixeff` over `lme4` at each cell. The dashed reference line at
1× would mean the two engines are equally fast.

```{r speedup-plot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 4, fig.alt = "Speedup of mixeff over lme4 vs design scale, by scenario."}
mixeff_rows <- subset(benchmark, engine == "mixeff")

ggplot(mixeff_rows,
       aes(scale_value, speedup_vs_lme4,
           colour = scenario_label, group = scenario_label)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey50") +
  geom_line() +
  geom_point() +
  scale_x_log10() +
  scale_y_continuous(breaks = seq(0, 8, by = 1)) +
  labs(
    x = "design scale",
    y = "speedup over lme4 (×)",
    colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        legend.direction = "vertical",
        panel.grid.minor = element_blank()) +
  guides(colour = guide_legend(ncol = 1))
```

Read the speedup figure with the caveats it deserves. Three
replications per cell give a stable median but a wide envelope; the
absolute timings are sub-15 ms throughout, so an extra millisecond
of system noise moves the ratio visibly. The qualitative pattern is
robust: `mixeff` is faster everywhere, and the speedup grows with
the design's complexity — exactly the regime where bootstrap
workflows spend most of their time.

## What should you measure?

There are two separate speed questions.

First, measure model fitting. This asks how quickly each engine solves the same
mixed model as the number of rows, groups, slopes, or crossed random effects
changes.

Second, measure bootstrap routes. This asks how long the user-facing inference
verbs take when they repeatedly refit models.

```{r refit-counts}
data.frame(
  bootstrap_replicates = c(50L, 200L, 1000L),
  approximate_refits = c(50L, 200L, 1000L)
)
```

The arithmetic is simple, but it is the core product point: small per-refit
speed differences compound quickly when `nsim` is large.

## How do you benchmark fitting?

Use the scaling benchmark when you want to compare `mixeff::lmm()` with
`lme4::lmer()` over several synthetic LMM designs.

```{r fit-benchmark-command, eval = FALSE}
system2(
  "Rscript",
  c(
    "inst/benchmarks/lme4-scaling.R",
    "--reps=5",
    "--scenarios=rows,groups,slopes,crossed",
    "--no-plots"
  )
)
```

That script writes one row per engine, design, scale value, and timing
repetition, plus a summary table with median elapsed seconds and speed relative
to `lme4`.

```{r fit-benchmark-files}
data.frame(
  file = c(
    "benchmarks/lme4-scaling/lme4-scaling-raw.csv",
    "benchmarks/lme4-scaling/lme4-scaling-summary.csv"
  ),
  contents = c(
    "one timing row per engine/design/repetition",
    "median seconds, fits per second, and relative speed"
  )
)
```

## How do you benchmark bootstrap?

Use the bootstrap benchmark when you want to time the inference routes that make
the audit-then-bootstrap story practical.

```{r bootstrap-benchmark-command, eval = FALSE}
system2(
  "Rscript",
  c(
    "inst/benchmarks/bootstrap-inference.R",
    "--nsim=200",
    "--reps=3"
  )
)
```

The benchmark includes `test_effect(method = "bootstrap")`,
`confint(method = "bootstrap")`, `compare(method = "bootstrap")`, and
`lme4::bootMer()` as an R-engine baseline. If `pbkrtest` is installed, it also
times `pbkrtest::PBmodcomp()`.

```{r bootstrap-routes}
data.frame(
  route = c(
    "mixeff_test_effect_bootstrap",
    "mixeff_confint_bootstrap",
    "mixeff_compare_bootstrap_lrt",
    "lme4_bootMer_fixef_distribution",
    "pbkrtest_PBmodcomp"
  ),
  target = c(
    "term-level fixed-effect p-value",
    "fixed-effect confidence interval",
    "model comparison p-value",
    "fixed-effect bootstrap distribution",
    "parametric bootstrap model comparison"
  )
)
```

## How do you read the result?

For a local speed claim, read the summary CSV and compute ratios from elapsed
seconds. Values above 1 mean the baseline route took longer than the `mixeff`
route.

```{r speedup-example, eval = FALSE}
summary <- read.csv("benchmarks/bootstrap-inference/bootstrap-inference-summary.csv")
mixeff_ci <- subset(summary, route == "mixeff_confint_bootstrap")
lme4_boot <- subset(summary, route == "lme4_bootMer_fixef_distribution")

data.frame(
  comparison = "lme4 bootMer / mixeff bootstrap CI",
  speedup = lme4_boot$median_sec / mixeff_ci$median_sec
)
```

Treat that ratio as local evidence, not a universal constant. It depends on the
formula, optimizer, random-effects structure, `nsim`, and machine. The point of
the benchmark scripts is to make the claim reproducible for the model class you
are actually fitting.

## What should go in a report?

When you report benchmark results, include the model formula, number of rows,
number of grouping levels, random-effects structure, `nsim`, repetitions, and
the package versions. The raw CSV has the timing rows; the summary CSV is for
tables and figures.

```{r reporting-fields}
data.frame(
  field = c("formula", "nobs", "groups", "random_terms", "nsim", "reps", "versions"),
  reason = c(
    "defines the fixed and random structure being refit",
    "controls the amount of data processed per fit",
    "controls the random-effect dimension",
    "controls covariance structure and optimizer work",
    "controls bootstrap cost",
    "shows timing stability",
    "makes the run reproducible"
  )
)
```
