---
title: "Convergence, Boundaries, and Singularity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Convergence, Boundaries, and Singularity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-package}
library(mixeff)
```

Every fit returned by `lmm()` or `glmm()` carries a labeled convergence
status, and the label is a claim with a precise scope. This vignette is
the reference for that vocabulary: what `converged_interior` does and
does not establish, how boundary and reduced-rank outcomes are recorded,
how singularity is actually decided, how to check an optimum with
`verify_convergence()`, and what happens when a requested GLMM estimator
does not certify. The other vignettes cross-reference these semantics
rather than restating them.

## What `converged_interior` establishes — and what it does not

Start with a fit that goes well: a small longitudinal study with twelve
clinics, a week effect, and a treatment contrast.

```{r clinic-data, include = FALSE}
set.seed(42)
n_clinic <- 12L
n_per_clinic <- 6L
clinic <- factor(rep(seq_len(n_clinic), each = n_per_clinic))
week <- rep(0:(n_per_clinic - 1L), n_clinic)
treatment <- factor(
  rep(rep(c("usual", "coached"), each = n_per_clinic), length.out = length(week)),
  levels = c("usual", "coached")
)
clinic_shift <- rnorm(n_clinic, sd = 0.45)
score <- 7.2 - 0.28 * week - 0.55 * (treatment == "coached") +
  clinic_shift[as.integer(clinic)] + rnorm(length(week), sd = 0.35)
clinic_visits <- data.frame(score, week, treatment, clinic)
```

```{r interior-fit}
fit <- lmm(
  score ~ week + treatment + (1 | clinic),
  clinic_visits,
  control = mm_control(verbose = -1)
)
fit_status(fit)
```

`converged_interior` means the optimizer returned a finite interior
solution that passed mixeff's declared numerical checks, with every
variance component away from zero. It does not establish global
optimality, correct model specification, or that the data carry enough
information — no numerical status can — but it is the status every
other block of a `summary()` assumes.

The status changes when something needs your attention. The vocabulary
has three values you will meet in practice:

- `converged_interior` — a finite interior optimum, as above.
- `converged_reduced_rank` — a genuine optimum at which the
  random-effect covariance has lower rank than the model requested;
  the next section walks through one.
- `not_optimized` — the conservative label for a run that stopped in a
  flat or boundary region *without* certifying a stationary point.
  Estimates are present but the engine does not call them an optimum,
  and `summary()` repeats that state in a `Notes:` line rather than
  relying on the header alone.

Optimization is silent and runs within one native call under bounded
evaluation budgets; the `lmm()`/`glmm()` help pages and `?mm_control`
document the budgets and how to cap them.

What the fit ran is recorded on the object. `optimizer_certificate()`
reports the optimizer, the stopping condition, and the certificate-time
facts the status label summarizes:

```{r interior-certificate}
optimizer_certificate(fit)
```

```{r interior-checks, include = FALSE}
stopifnot(inherits(fit, "mm_lmm"))
stopifnot(identical(fit_status(fit), "converged_interior"))
stopifnot(isFALSE(is_singular(fit)))
stopifnot(nrow(optimizer_certificate(fit)$table) > 0L)
```

## Boundary and reduced-rank fits are labeled facts

Now a fit that does not go well — deliberately. Eighteen subjects, ten
daily reaction-time measurements each, with subject intercepts and
slopes that are nearly perfectly correlated by construction. This
design produces a singular fit in `lme4` and in `mixeff` alike.

```{r singular-data, include = FALSE}
set.seed(3)
n_subj <- 18L
days <- 0:9
b0 <- rnorm(n_subj, sd = 30)
b1 <- 0.5 * b0 / 30 * 10 + rnorm(n_subj, sd = 0.5)

sleep_like <- do.call(rbind, lapply(seq_len(n_subj), function(i) {
  data.frame(
    subj = factor(i),
    days = days,
    rt = 250 + b0[i] + (10 + b1[i]) * days +
      rnorm(length(days), sd = 20)
  )
}))
```

`lme4::lmer()` fits this model and reports the situation in one short
line: `boundary (singular) fit`. The fact is correct. What it leaves
implicit is *which* variance component reached the boundary, what the
effective rank of the random-effect covariance is, and which downstream
inference methods the boundary takes off the table.

```{r lme4-side, eval = requireNamespace("lme4", quietly = TRUE)}
m <- suppressMessages(lme4::lmer(
  rt ~ days + (1 + days | subj),
  data = sleep_like
))
m
lme4::isSingular(m)
```

`mixeff` reports the same fact, and then unpacks it. `fit_status()`
names the convergence outcome, `is_singular()` answers the yes/no
question, `diagnostics()` returns stable codes with severity and stage,
and `changes()` shows the requested-to-effective-to-fitted arc.

```{r singular-fit}
sfit <- lmm(
  rt ~ days + (1 + days | subj),
  sleep_like,
  control = mm_control(verbose = -1)
)

fit_status(sfit)
is_singular(sfit)
diagnostics(sfit)$table[, c("code", "severity", "stage", "message")]
changes(sfit)
```

The requested random-effect covariance for `(1 + days | subj)` has rank
2: one dimension for the subject intercept and one for the subject
slope. The fitted covariance has effective rank 1 — the model asked for
two random-effect dimensions, but the data support only one effective
dimension. That is recorded as the `covariance_reduced` diagnostic and
the `converged_reduced_rank` status, and the printed object carries the
tag in its header every time it prints — where `lme4` warns once at fit
time and then falls silent.

```{r singular-print}
print(sfit)
```

```{r singular-checks, include = FALSE}
stopifnot(identical(fit_status(sfit), "converged_reduced_rank"))
stopifnot(isTRUE(is_singular(sfit)))
diag_codes <- diagnostics(sfit)$table$code
stopifnot("covariance_reduced" %in% diag_codes)
ch <- changes(sfit)$table
stopifnot("design_time_reduction" %in% ch$stage ||
          "covariance_transition" %in% ch$stage)
sfit_print <- capture.output(print(sfit))
stopifnot(any(grepl("reduced_rank|boundary", sfit_print)))
```

A reduced-rank covariance does not automatically invalidate the model.
It is a labeled fact about the fit, recorded in the object rather than
printed once and lost — and it changes which inference routes the
package will certify (one sentence on that at the end of this
vignette).

## How singularity is decided: rank, not θ entries

To read the boundary structure itself, look at the model's internal
coordinates. The random-effect covariance Σ is not optimized directly.
The engine searches over a vector θ that fills a lower-triangular
factor Λ, with Σ = σ²ΛΛᵀ; θ is by construction exactly the set of free
entries of Λ. This is the parameterization Bates et al. (2015)
describe for `lme4`; `MixedModels.jl` and the engine behind `mixeff`
inherit it. `parameterization()` shows the mapping for the singular
fit:

```{r singular-parameterization}
parameterization(sfit)$table[, c("theta_name", "theta_value", "lambda_value")]
```

The *diagonal* entry for the `days` slope is pinned at numerical zero
— that is the boundary event behind the `converged_reduced_rank`
status.

It is tempting to generalize this to "a zero θ entry means the
covariance is reduced-rank". That reading is wrong. A zero
**off-diagonal** entry of Λ only says the corresponding covariance is
zero; the covariance matrix can keep full rank. Singularity is a
property of the covariance matrix's rank — equivalently its
eigenstructure — and in this parameterization the rank collapses
exactly when a **diagonal** entry of Λ is zero. A two-by-two example
makes the difference concrete:

```{r lambda-rank-demo}
l11 <- 1.2
l21 <- 0.4
l22 <- 0.9

# Zero the OFF-DIAGONAL entry (l21 = 0): zero covariance, full rank.
L_offdiag_zero <- matrix(c(l11, 0, 0, l22), nrow = 2)
Sigma_offdiag_zero <- L_offdiag_zero %*% t(L_offdiag_zero)
Sigma_offdiag_zero
eigen(Sigma_offdiag_zero, symmetric = TRUE)$values

# Zero a DIAGONAL entry (l22 = 0): the rank collapses to 1.
L_diag_zero <- matrix(c(l11, l21, 0, 0), nrow = 2)
Sigma_diag_zero <- L_diag_zero %*% t(L_diag_zero)
Sigma_diag_zero
eigen(Sigma_diag_zero, symmetric = TRUE)$values
```

With the off-diagonal at zero, both eigenvalues are positive: the
intercept and slope are uncorrelated, and each still has its own
variance — an ordinary full-rank fit (the model `(1 + days || subj)`
*asks* for exactly this shape). With a diagonal entry at zero, one
eigenvalue is zero and the two random effects collapse onto a single
dimension: that is singularity.

```{r lambda-rank-checks, include = FALSE}
stopifnot(identical(qr(Sigma_offdiag_zero)$rank, 2L))
stopifnot(all(eigen(Sigma_offdiag_zero, symmetric = TRUE)$values > 0))
stopifnot(identical(qr(Sigma_diag_zero)$rank, 1L))
ev_diag <- eigen(Sigma_diag_zero, symmetric = TRUE)$values
stopifnot(abs(ev_diag[[2L]]) < 1e-12)
stopifnot(identical(Sigma_offdiag_zero[1L, 2L], 0))
# The fitted singular case: the slope's DIAGONAL theta entry is at zero.
pt <- parameterization(sfit)$table
stopifnot(pt$theta_value[pt$theta_name == "theta[0:days,days]"] < 1e-6)
```

Consistent with this, `is_singular()` does not scan θ for zeros. It
reads the covariance status the engine recorded on the fit — the fit
status plus the per-block effective-covariance status — so its answer
is the engine's rank verdict, not an R-side reconstruction.

## Verifying an optimum: `verify_convergence()`

A convergence status describes the run that produced the fit. If you
want independent evidence, `verify_convergence()` re-runs the fit
under the engine's bounded verification workflow: a restart from the
fitted optimum, one or more jittered restarts, and (LMM-only, opt-in)
an alternate-optimizer consensus pass. The engine judges agreement
against explicit tolerances on the objective, θ, and the fixed
effects, and owns the verdict; R only formats it.

```{r verify-lmm}
vc <- verify_convergence(fit)
vc
```

Each run either agrees with the reference optimum within the printed
tolerances or it does not; the verdict summarizes the runs
(`restart_agrees`, `optimizer_consensus`, `fragile`, `unstable`, or
`not_run`). A verdict of `restart_agrees` is evidence of a *stable*
optimum — the bounded re-runs came back to the same place — not proof
of a global one. The `consensus` pass defaults to `FALSE` because this
vendored build compiles without the optional `nlopt` backend; for some
models the engine's alternate optimizer choice is an nlopt optimizer,
and its absence would read as a spurious `fragile` verdict that
reflects the build, not the fit.

Verification and rank are different questions. Running the verifier on
the singular fit from above also comes back `restart_agrees`: a
reduced-rank optimum can be a perfectly stable optimum.

```{r verify-singular}
vs <- verify_convergence(sfit)
vs$status
```

The GLMM method uses the same workflow with wider default tolerances —
objective `1e-4` and beta `1e-3`, against the LMM defaults of `1e-5`
and `1e-4` — because every GLMM objective evaluation carries
inner-PIRLS noise and the joint path's derivative-free beta search
bounds beta reproducibility at its stopping tolerance. The consensus
pass is refused for GLMM fits (the GLMM fit drivers own optimizer
selection).

```{r glmm-data, include = FALSE}
set.seed(7)
g <- factor(rep(1:12, each = 12))
x <- rnorm(144)
eta <- -0.3 + 0.8 * x + rep(rnorm(12, sd = 0.5), each = 12)
trial <- data.frame(y = rbinom(144, 1, plogis(eta)), x = x, g = g)
```

```{r verify-glmm}
gfit <- glmm(
  y ~ x + (1 | g),
  trial,
  family = binomial(),
  control = mm_control(verbose = -1)
)

gv <- verify_convergence(gfit)
gv
```

```{r verify-checks, include = FALSE}
stopifnot(identical(vc$status, "restart_agrees"))
stopifnot(all(vc$table$agrees))
stopifnot(identical(vc$tolerances$objective, 1e-5))
stopifnot(identical(vc$tolerances$beta, 1e-4))
stopifnot(identical(vs$status, "restart_agrees"))
stopifnot(identical(gv$status, "restart_agrees"))
stopifnot(all(gv$table$agrees))
stopifnot(identical(gv$tolerances$objective, 1e-4))
stopifnot(identical(gv$tolerances$beta, 1e-3))
```

## When the requested estimator does not certify (GLMM)

`glmm(method = "joint_laplace")` requests the joint-Laplace estimator
(the route that certifies Wald inference; `vignette("glmm")` is the
operational quickstart). On some designs the joint route does not
certify an optimum. The engine then returns a *labelled* fallback fit
from the fast profiled-PIRLS estimator instead of failing or — worse —
silently presenting the fallback as the thing you asked for. The
substitution crosses the boundary as a typed record, in keeping with
the package's no-silent-surgery contract:

- **The typed record.**
  `optimizer_certificate(fit)$raw$estimator_substitution` carries the
  requested method, the effective (fallback) method, the requested
  route's fit status and return code, and the engine's reason. It is
  the machine-readable contract surface; scripts should gate on it,
  not on message text.
- **The fit-time notice.** A condition of class
  `mm_estimator_substitution_notice` announces the substitution when
  the fit returns (suppressible with `mm_control(verbose = -1)`).
- **The ungated summary note.** `summary()` prints an
  `estimator substitution:` note naming both estimators regardless of
  verbosity settings, and independent of the fallback's own
  `converged_*` status — a clean-looking status line must not hide
  the fact that a different estimator than requested produced the
  numbers.
- **The checking verb.** `verify_convergence()` on a substituted fit
  re-requests the joint route; when the verification refit falls back
  the same way — the expected case — every run verifies the profiled
  objective the fitted numbers actually came from.

On an ordinary fit the record is absent, and its absence is meaningful
(no substitution happened):

```{r substitution-surface}
optimizer_certificate(gfit)$raw$estimator_substitution
```

```{r substitution-checks, include = FALSE}
stopifnot(is.null(optimizer_certificate(gfit)$raw$estimator_substitution))
```

A real example, from reproducing a published `glmer()` analysis of OSF
study data (an error-rate model with an open-practice × year
interaction): requesting `method = "joint_laplace"` with the raw,
uncentered `Year` predictor hands the engine an ill-scaled design, the
joint route does not certify, and the result comes back as a labelled
fast-PIRLS fallback — fit-time notice, summary note, and typed record
all present, with Wald inference withheld. Centering the year
(`Year - 2015`) makes the joint route certify and match `glmer()` on
the same data. mixeff flags this situation at fit time: a scaling
advisory names the ill-scaled columns, and centering or rescaling the
predictor — as the year centering does here — is the standard remedy.

## What a boundary refuses

A boundary fit also changes which inference routes the package will
certify: Satterthwaite and Kenward–Roger degrees of freedom, for
example, are refused on singular fits by mixeff's validated-support
policy, with stable reason codes and a bootstrap alternative offered
under the same labeling rules — the route map, the refusal codes, and
the alternatives are documented in `vignette("inference")`.

## References

- Bates, D., Mächler, M., Bolker, B., and Walker, S. (2015). Fitting
  linear mixed-effects models using lme4. *Journal of Statistical
  Software*, 67(1), 1–48. <https://doi.org/10.18637/jss.v067.i01>
