---
title: GLMM Boundary and Model Comparison
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLMM Boundary and Model Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-package}
library(mixeff)
```

```{r study-data, include = FALSE}
set.seed(11)
n_subj <- 30L
n_time <- 5L
study <- expand.grid(subject = factor(seq_len(n_subj)),
                     time = seq_len(n_time))
b0 <- rnorm(n_subj, sd = 0.6)[as.integer(study$subject)]
study$x <- rnorm(nrow(study))
study$z <- rnorm(nrow(study))
study$y <- 1 + 0.4 * study$x + 0.2 * study$z + b0 +
  rnorm(nrow(study), sd = 0.5)
study$hit <- rbinom(nrow(study), size = 1, prob = plogis(0.5 + study$x))
```

This page covers two things. The first is the lifecycle tools that
follow a linear mixed-model fit and that `lme4` users already reach
for: `simulate()`, `refit()`, `compare()`, multi-model `anova()`,
`drop1()`, and a small parametric-bootstrap path.

The second is the GLMM front door — `glmm()` — which accepts the
familiar `family` and `link` arguments, validates them, and currently
raises a *typed* error because the GLMM fitting path is not yet
available in this build. A typed boundary is a deliberate choice: an
analyst who tries `glmm()` on their data sees a stable refusal
message rather than a fallback fit they did not ask for.

```{r glmm-boundary, error = TRUE}
glmm(
  hit ~ x + (1 | subject),
  study,
  family = binomial(),
  control = mm_control(verbose = -1)
)
```

## Simulate and refit

```{r simulate-refit}
fit <- lmm(y ~ x + z + (1 | subject), study, control = mm_control(verbose = -1))

sims <- simulate(fit, nsim = 2, seed = 10)
head(sims)

refit_fit <- refit(fit, sims$sim_1)
fixef(refit_fit)
```

`simulate()` uses the stored fixed effects, random-effect covariance summaries,
and residual scale. `refit()` uses the stored formula and model frame with the
new response.

## Compare models

```{r compare}
reduced <- lmm(y ~ x + (1 | subject), study, control = mm_control(verbose = -1))

compare(reduced, fit)
anova(reduced, fit)
```

REML fits are refit by ML for likelihood comparison when
`refit_for_comparison = "auto"` (the default). The comparison table records
which rows were refit.

```{r drop1}
drop1(fit, test = "Chisq")
```

## Parametric bootstrap

`parametric_bootstrap()` runs the engine-certified bootstrap
likelihood-ratio test and reports the replicate accounting the Rust engine
produced. It requires ML fits; `compare(method = "bootstrap")` refits REML
to ML for you. Larger `nsim` values are slower.

```{r bootstrap}
reduced_ml <- lmm(y ~ x + (1 | subject), study, REML = FALSE,
                   control = mm_control(verbose = -1))
fit_ml <- lmm(y ~ x + z + (1 | subject), study, REML = FALSE,
              control = mm_control(verbose = -1))
boot <- parametric_bootstrap(reduced_ml, fit_ml, nsim = 2, seed = 99)
boot

compare(reduced, fit, method = "bootstrap", nsim = 2, seed = 99)
```

```{r checks, include = FALSE}
stopifnot(nrow(sims) == nrow(study))
stopifnot(inherits(refit_fit, "mm_lmm"))
cmp <- compare(reduced, fit)
stopifnot(nrow(cmp$table) == 2L)
stopifnot(any(cmp$table$refit))
stopifnot(inherits(drop1(fit), "mm_drop1"))
stopifnot(inherits(boot, "mm_parametric_bootstrap"))
```
