---
title: GLMM Fitting and Model Comparison
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GLMM Fitting and Model Comparison}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_lme4 <- requireNamespace("lme4", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = has_lme4
)
```

```{r load-package, eval = TRUE}
library(mixeff)
```

```{r lme4-missing, eval = !has_lme4, echo = FALSE, results = "asis"}
cat("The executable cbpp examples in this vignette require the `lme4` package.\n")
```

`glmm()` fits generalized linear mixed models by one of two estimators, and
the difference between them decides what inference you get:

- **`method = "pirls_profiled"` (the default)** is a fast profiled PIRLS
  fitter. It returns point estimates, fitted values, and variance
  components, but **refuses Wald standard errors, z statistics, p-values,
  and confidence intervals**: its covariance payload is a working Hessian
  that the engine does not certify for inference. The engine's own
  documentation adds a stronger warning: the profiled path is not the same
  statistical approximation as `glmer()`'s joint Laplace fit, and it can be
  less accurate for overdispersed models and models with observation-level
  random effects.
- **`method = "joint_laplace"`** maximizes the same joint Laplace objective
  as `lme4::glmer()` (at `nAGQ = 1`). It is slower, and it returns the full
  inference table — standard errors, z tests, and Wald confidence
  intervals — from a covariance the engine certifies.

Three inference routes exist, and every refusal message names them:

1. **Certified Wald** — refit with `method = "joint_laplace"`; standard
   errors, z tests, and Wald intervals come from a covariance the engine
   certifies.
2. **Parametric bootstrap** — `confint(fit, method = "bootstrap")` works on
   profiled-estimator fits for every supported family, including negative
   binomial (which has no joint-Laplace route). Replicates simulate from
   the fitted model and refit the same estimator; the result carries
   percentile intervals plus full replicate accounting (requested,
   successful, failed, Monte Carlo SE). It costs `nsim` refits. On
   `joint_laplace` fits the bootstrap is refused at this engine pin
   (replicate refits cannot re-run the joint estimator) — those fits have
   certified Wald intervals instead.
3. **Working-Hessian approximation (opt-in)** — for exploration and
   screening only; see `inference_options(fit)` for the caveat and the
   invocation.

The practical rule: explore with the default, report with `joint_laplace`
or the bootstrap.

The example uses the `lme4::cbpp` data. As in `lme4`, a binomial response
with grouped counts can be written either as a two-column
`cbind(successes, failures)` response or as a proportion with case
`weights`; the two forms give identical fits.

```{r cbpp-data}
env <- new.env(parent = emptyenv())
utils::data("cbpp", package = "lme4", envir = env)
cbpp <- get("cbpp", envir = env, inherits = FALSE)
cbpp$prop <- cbpp$incidence / cbpp$size

cbpp_family <- binomial(link = "logit")
cbind_formula  <- cbind(incidence, size - incidence) ~ period + (1 | herd)
weight_formula <- prop ~ period + (1 | herd)
```

## Family and Audit

The statistical intent is a binomial-logit GLMM with a herd random
intercept. Before fitting, the formula can be compiled and explained so the
fixed- and random-effect structure is visible before optimization starts.
(`compile_model()` requires a plain-column response, so the audit uses the
proportion form; `glmm()` itself accepts both.)

```{r audit-first}
cbpp_spec <- compile_model(weight_formula, cbpp)
explain_model(cbpp_spec)
```

## Fit

```{r fit-default}
glmm_fit <- glmm(
  cbind_formula,
  cbpp,
  family = cbpp_family,
  control = mm_control(verbose = -1)
)

glmm_fit
fixef(glmm_fit)
VarCorr(glmm_fit)
```

The weights form reproduces it exactly:

```{r fit-weights}
glmm_fit_w <- glmm(
  weight_formula,
  cbpp,
  family = cbpp_family,
  weights = cbpp$size,
  control = mm_control(verbose = -1)
)
all.equal(fixef(glmm_fit), fixef(glmm_fit_w))
```

On this default fit, `summary()` prints the estimates and withholds the
inference columns, with a note naming the alternative:

```{r summary}
summary(glmm_fit, tests = "coefficients")
```

Core extractors read from the durable R object:

```{r extractors}
head(fitted(glmm_fit))
head(residuals(glmm_fit))
head(ranef(glmm_fit)[[1L]])
c(
  logLik = as.numeric(logLik(glmm_fit)),
  deviance = deviance(glmm_fit),
  AIC = AIC(glmm_fit),
  BIC = BIC(glmm_fit)
)
```

## Inference: refit with joint_laplace

```{r fit-joint}
glmm_joint <- glmm(
  cbind_formula,
  cbpp,
  family = cbpp_family,
  method = "joint_laplace",
  control = mm_control(verbose = -1)
)

summary(glmm_joint, tests = "coefficients")
confint(glmm_joint)
```

The joint-Laplace estimates differ slightly from the profiled ones —
different estimators, same model — and match `lme4::glmer()` closely:

```{r joint-vs-glmer}
glmer_fit <- lme4::glmer(
  cbind(incidence, size - incidence) ~ period + (1 | herd),
  data = cbpp, family = binomial()
)
round(rbind(
  pirls_profiled = fixef(glmm_fit),
  joint_laplace  = fixef(glmm_joint),
  glmer          = lme4::fixef(glmer_fit)
), 4)
```

## Quadrature Sensitivity

`nAGQ` is part of the fit request and is recorded on the object. Values
above one are a profiled-path sensitivity check; the joint-Laplace path is
restricted to `nAGQ = 1`.

```{r agq-sensitivity}
glmm_fit_agq3 <- glmm(
  cbind_formula,
  cbpp,
  family = cbpp_family,
  nAGQ = 3L,
  control = mm_control(verbose = -1)
)

data.frame(
  nAGQ = c(glmm_fit$nAGQ, glmm_fit_agq3$nAGQ),
  logLik = c(as.numeric(logLik(glmm_fit)),
             as.numeric(logLik(glmm_fit_agq3))),
  AIC = c(AIC(glmm_fit), AIC(glmm_fit_agq3)),
  check.names = FALSE
)
```

On this dataset the quadrature order shifts the objective in the third
decimal and the coefficients by less than 0.005 on the logit scale —
visible, but an order of magnitude below the joint-Laplace standard
errors shown earlier.

```{r agq-check, include = FALSE}
stopifnot(max(abs(fixef(glmm_fit) - fixef(glmm_fit_agq3))) < 0.005)
```

## What is refused where

The table below runs each verb on the default fit and reports what
actually comes back — a result with numbers, a result with the inference
withheld, or a typed error. A no-error return is not the same thing as
an answer, so for the coefficient table the harness also checks whether
the standard errors are finite.

```{r boundaries}
glmm_status <- function(expr, has_numbers = NULL) {
  cnd <- tryCatch({ force(expr); NULL }, error = function(cnd) cnd)
  if (!is.null(cnd)) {
    return(data.frame(status = "typed error", class = class(cnd)[[1L]],
                      check.names = FALSE))
  }
  if (isFALSE(has_numbers)) {
    return(data.frame(status = "returns, inference withheld",
                      class = NA_character_, check.names = FALSE))
  }
  data.frame(status = "available", class = NA_character_, check.names = FALSE)
}

coef_tab <- summary(glmm_fit, tests = "coefficients")$coefficients
rbind(
  predict           = glmm_status(predict(glmm_fit)),
  fitted            = glmm_status(fitted(glmm_fit)),
  coefficient_tests = glmm_status(coef_tab,
                                  has_numbers = any(is.finite(coef_tab[, "Std. Error"]))),
  confint           = glmm_status(confint(glmm_fit)),
  simulate          = glmm_status(stats::simulate(glmm_fit, nsim = 1L)),
  refit             = glmm_status(refit(glmm_fit, fitted(glmm_fit)))
)
```

Wald `confint()` on the default fit raises `mm_inference_unavailable` (the
working-Hessian payload is not certified for Wald intervals); the same
request on the joint-Laplace fit succeeds, as shown above. The bootstrap
route complements it from the other side: `confint(glmm_fit, method =
"bootstrap")` resamples and refits the profiled estimator itself, and is
refused on joint fits (which have certified Wald instead). `simulate()`
and `refit()` are not implemented for GLMM fits.

```{r checks, include = FALSE}
stopifnot(inherits(glmm_fit, "mm_glmm"))
stopifnot(identical(glmm_fit$nAGQ, 1L))
stopifnot(identical(glmm_fit_agq3$nAGQ, 3L))
stopifnot(nobs(glmm_fit) == nrow(cbpp))
stopifnot(all(is.finite(fixef(glmm_fit))))
stopifnot(isTRUE(all.equal(fixef(glmm_fit), fixef(glmm_fit_w), tolerance = 1e-8)))
stopifnot(is.finite(as.numeric(logLik(glmm_fit))))
stopifnot(is.finite(AIC(glmm_fit)))

# Default path: estimates present, inference withheld (all SEs NA).
stopifnot(!any(is.finite(coef_tab[, "Std. Error"])))
# Bootstrap route works on the default fit (prose above claims it).
boot_ci <- confint(glmm_fit, method = "bootstrap", nsim = 25L, seed = 1L)
stopifnot(all(is.finite(boot_ci)), all(boot_ci[, 1L] < boot_ci[, 2L]))
# Joint path: full inference table, finite everywhere.
joint_tab <- summary(glmm_joint, tests = "coefficients")$coefficients
stopifnot(all(is.finite(joint_tab[, "Std. Error"])))
stopifnot(all(is.finite(confint(glmm_joint))))
# Joint-Laplace tracks glmer on this model.
stopifnot(max(abs(fixef(glmm_joint) - lme4::fixef(glmer_fit))) < 5e-3)

# Refusals stay refusals.
stopifnot(inherits(tryCatch(confint(glmm_fit), error = identity),
                   "mm_inference_unavailable"))
stopifnot(inherits(tryCatch(stats::simulate(glmm_fit, nsim = 1L),
                            error = identity),
                   "mm_inference_unavailable"))
```
