---
title: Inference Where lme4 Says No
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inference Where lme4 Says No}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-package}
library(mixeff)
```

Some mixed-model fits are exactly the cases where ordinary
asymptotic p-values (Wald, Satterthwaite, Kenward-Roger) are least
trustworthy: boundary fits, reduced-rank covariance estimates, and
designs whose degrees-of-freedom approximations stop being defined.

The mathematics behind these limits is well established. Self and
Liang (1987) showed that the likelihood-ratio statistic does not
follow its standard χ² distribution when a parameter lies on the
boundary of its parameter space; Stram and Lee (1994) specialised
the result to variance components in linear mixed models. Kenward
and Roger (1997), who introduced the most widely used small-sample
adjustment for fixed-effect tests, derived it from a Taylor
expansion of the variance estimator that requires an *interior*
solution for the variance parameters. At the boundary the
adjustment is undefined by construction. `lme4` typically returns
the Wald number alongside a one-line `boundary (singular) fit`
warning, leaving the analyst to decide what to make of the
situation.

`mixeff` does two things differently. First, it labels the
asymptotic row with a *closed-enum* reliability reason that names
why the number is suspect on this fit — for example,
`*_unavailable_at_boundary` when a variance-parameter derivative is
not defined. Second, when bootstrap is the defensible route, it
routes the analyst through a parametric bootstrap as a labelled
first-class peer to the asymptotic methods. The parametric
bootstrap has a long pedigree in this setting — Halekoh and
Højsgaard (2014) implemented it in `pbkrtest` for exactly the
small-sample and boundary cases where Kenward-Roger needs help. It
is not a fallback trick; it is the path the inference contract
documents for this situation.

## What fit are we worried about?

This small repeated-measures example has subject-specific intercepts and slopes
that are nearly collinear. That is enough to create a reduced-rank random-effect
covariance.

```{r singular-data, include = FALSE}
set.seed(3)
n_subj <- 18L
days <- 0:9
b0 <- rnorm(n_subj, sd = 30)
b1 <- 0.5 * b0 / 30 * 10 + rnorm(n_subj, sd = 0.5)

sleep_like <- do.call(rbind, lapply(seq_len(n_subj), function(i) {
  data.frame(
    subj = factor(i),
    days = days,
    rt = 250 + b0[i] + (10 + b1[i]) * days +
      rnorm(length(days), sd = 20)
  )
}))
```

```{r fit-singular}
fit <- lmm(
  rt ~ days + (1 + days | subj),
  sleep_like,
  control = mm_control(verbose = -1)
)

fit_status(fit)
is_singular(fit)
```

```{r singular-checks, include = FALSE}
stopifnot(inherits(fit, "mm_lmm"))
stopifnot(isTRUE(is_singular(fit)))
stopifnot(identical(fit_status(fit), "converged_reduced_rank"))
```

The fit is usable, but the covariance state matters for inference. The audit
verbs keep that state attached to the object.

```{r covariance-diagnostics}
diagnostics(fit)$table[, c("code", "severity", "stage", "message")]
```

## Which inference routes are available?

Ask before you test. `inference_options()` is an audit map, not a
recommender. Each row enumerates one inference method that the
engine knows about, along with a status, a warrant, and a rough
cost.

Three columns deserve a short key, because the values are closed
enums and their meanings are not self-evident.

`expected_status` takes one of three values:

- **`available`** — the method runs on this fit and produces a
  meaningful number.
- **`not_assessed`** — the inference contract refuses to run the
  method on this fit, for a structural reason named in the next
  column. This is the contract drawing a line, not failing
  silently.
- **`not_yet_wired`** — the method is part of the inference
  contract upstream but has not yet been exposed through the R
  surface. A roadmap item, not a refusal.

`expected_reliability_reason` is the closed-enum warrant — the
*why* behind the status. The values that appear on this fit:

- **`*_unavailable_at_boundary`** (here, the `satterthwaite_` and
  `kenward_roger_` variants) — the asymptotic method relies on a
  derivative of the variance estimator that, as the previous
  section explained, is undefined on a boundary fit.
- **`bootstrap_monte_carlo_replicates`** — the method's reliability
  is governed by the replicate count and the Monte-Carlo standard
  error rather than asymptotic theory. You tune it with `nsim`.
- **`inference_unavailable_by_policy`** — the engine has a refusal
  policy specific to this combination of method and fit state; the
  policy itself is the warrant.
- **`not_available`** — despite the name, this value means that
  *no additional reason applies* on this row. The Wald row is the
  case in point: the method runs, and its substantive reliability
  warrant is recorded in the main `inference_table()` (shown
  below), not duplicated here.

`approx_cost` is a rough timing estimate. `immediate` is sub-
second; everything else scales with `nsim` and the cost of one
refit.

```{r route-map}
routes <- inference_options(fit, "days", nsim = 200)
routes$table[, c("method", "expected_status",
                 "expected_reliability_reason", "approx_cost", "current")]
```

```{r route-checks, include = FALSE}
stopifnot("bootstrap" %in% routes$table$method)
satt <- routes$table[routes$table$method == "satterthwaite", ]
stopifnot(identical(satt$expected_status, "not_assessed"))
```

Read the table top to bottom. **Wald** is `available`; its
reliability warrant lives in `inference_table()` (next chunk).
**Satterthwaite** and **Kenward-Roger** refuse, both at the
boundary, for the structural reason discussed above.
**Bootstrap** is `available` at a cost of roughly four seconds for
`nsim = 200` — this is the route the contract documents for
boundary fits. **Bootstrap-LRT** is refused on this fit shape by
the engine's current policy. **Cluster bootstrap** appears in the
upstream contract but the R bridge is not yet in place; the closing
section of this vignette comes back to it as a typed refusal.

```{r wald-status}
inference_table(fit)$table[, c("term", "method", "status",
                               "reliability", "reliability_reason")]
```

## How do I test the term anyway?

For the fixed-effect term, request the bootstrap route explicitly. Rust builds
the fixed-effect-null target, simulates from that target, refits each replicate,
and returns a labelled term row.

```{r bootstrap-term}
term_boot <- test_effect(
  fit,
  "days",
  method = "bootstrap",
  bootstrap = bootstrap_control(nsim = 50, seed = 1)
)

term_boot$table[, c("term", "statistic_name", "p_value",
                    "method", "status", "reliability_reason")]
```

```{r bootstrap-term-checks, include = FALSE}
stopifnot(identical(term_boot$table$status, "available"))
stopifnot(identical(term_boot$table$method, "bootstrap"))
stopifnot(is.finite(term_boot$table$p_value))
```

The p-value is not reconstructed in R. The row carries the Rust run payload in
`details`, including replicate accounting.

```{r bootstrap-payload}
run <- term_boot$table$details[[1]]$bootstrap
data.frame(
  requested_replicates = run$requested_replicates,
  successful_replicates = run$successful_replicates,
  boundary_rate = round(run$boundary_rate, 3),
  mcse = round(run$mcse, 4)
)
```

## What about confidence intervals?

For fixed-effect confidence intervals, use `confint(method = "bootstrap")`.
This is a different bootstrap target: a full-model distribution used for
intervals and diagnostics, not for p-values.

```{r bootstrap-ci}
ci <- confint(
  fit,
  parm = "days",
  method = "bootstrap",
  bootstrap = bootstrap_control(nsim = 50, seed = 2)
)

ci
```

```{r bootstrap-ci-checks, include = FALSE}
stopifnot(all(is.finite(ci)))
stopifnot(identical(attr(ci, "method"), "bootstrap_full_model_distribution"))
payload <- attr(ci, "bootstrap")[[1]]
stopifnot(identical(payload$metadata$target$kind, "full_model_distribution"))
stopifnot(length(payload$replicate_statistics) >= payload$metadata$successful_replicates,
          length(payload$replicate_statistics) <= payload$metadata$completed_replicates)
```

The attached payload says what was simulated. A full-model bootstrap interval
does not certify a hypothesis-test p-value, and the payload records that caveat.

```{r ci-payload}
payload$metadata$target$kind
payload$metadata$notes
```

## Where does cluster bootstrap stand?

Cluster resampling is available upstream as an estimator-distribution target:
it resamples grouping levels and summarizes the resulting estimator
distribution. In the current schema it does not certify fixed-effect p-values,
so a term test refuses cleanly instead of inventing a null distribution.

```{r cluster-refusal}
cluster_row <- test_effect(
  fit,
  "days",
  method = "cluster_bootstrap",
  group = "subj"
)

cluster_row$table[, c("term", "method", "status", "p_value", "reason_code")]
```

```{r cluster-checks, include = FALSE}
stopifnot(identical(cluster_row$table$status, "not_assessed"))
stopifnot(is.na(cluster_row$table$p_value))
stopifnot(identical(cluster_row$table$reason_code,
                    "bootstrap_cluster_resample_p_value_unavailable"))
```

That refusal is part of the same contract as the available bootstrap rows:
available numbers are labelled, and unavailable numbers carry stable reasons.

## References

- Halekoh, U., and Højsgaard, S. (2014). A Kenward-Roger approximation
  and parametric bootstrap methods for tests in linear mixed models —
  the `R` package `pbkrtest`. *Journal of Statistical Software*,
  59(9), 1–32. <https://doi.org/10.18637/jss.v059.i09>
- Kenward, M. G., and Roger, J. H. (1997). Small sample inference for
  fixed effects from restricted maximum likelihood. *Biometrics*,
  53(3), 983–997. <https://doi.org/10.2307/2533558>
- Self, S. G., and Liang, K.-Y. (1987). Asymptotic properties of
  maximum likelihood estimators and likelihood ratio tests under
  nonstandard conditions. *Journal of the American Statistical
  Association*, 82(398), 605–610.
  <https://doi.org/10.1080/01621459.1987.10478472>
- Stram, D. O., and Lee, J. W. (1994). Variance components testing in
  the longitudinal mixed effects model. *Biometrics*, 50(4),
  1171–1177. <https://doi.org/10.2307/2533455>
