---
title: Inference
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-package}
library(mixeff)
```

Mixed-model users want p-values. They also want to know what those
p-values mean: which method produced them, and how trustworthy that
method is for this particular fit. `lme4::lmer()` famously declines to
print LMM p-values at all — the reason `lmerTest` exists. `mixeff`
prints them, and prints the method and a reliability grade beside each
one.

When a coefficient, contrast, or term test has an available method,
the p-value is printed with the method name. When the requested
method is unavailable on this fit, the row says so, with a stable
reason code rather than an apologetic substitute. The full vocabulary
of reason codes, and the grading rubric behind the `reliability`
column, live in the inference-method glossary on the package site
(<https://bbuchsbaum.github.io/mixeff/articles/inference-method-glossary.html>);
this vignette points there rather than restating them.

This vignette covers the whole inference surface: coefficient tests,
contrasts, term tests, and model comparisons on a well-behaved fit;
the full route map — including what is refused at a boundary fit and
why; the GLMM three-route contract; and the marginal-means verbs,
starting with the question every marginal summary should answer first:
*which estimand are you averaging toward?*

## What model are we fitting?

```{r clinic-data, include = FALSE}
set.seed(42)
n_clinic <- 12L
n_per_clinic <- 6L
clinic <- factor(rep(seq_len(n_clinic), each = n_per_clinic))
week <- rep(0:(n_per_clinic - 1L), n_clinic)
treatment <- factor(
  rep(rep(c("usual", "coached"), each = n_per_clinic), length.out = length(week)),
  levels = c("usual", "coached")
)
clinic_shift <- rnorm(n_clinic, sd = 0.45)
score <- 7.2 - 0.28 * week - 0.55 * (treatment == "coached") +
  clinic_shift[as.integer(clinic)] + rnorm(length(week), sd = 0.35)
clinic_visits <- data.frame(score, week, treatment, clinic)
```

```{r fit-model}
fit <- lmm(
  score ~ week + treatment + (1 | clinic),
  clinic_visits,
  control = mm_control(verbose = -1)
)
```

```{r fit-checks, include = FALSE}
stopifnot(inherits(fit, "mm_lmm"))
stopifnot(all(is.finite(fixef(fit))))
```

The fixed effects ask whether scores change across weeks and whether the
coached program differs from usual care, while clinic baselines are allowed to
vary.

## Coefficient p-values

Start with the ordinary coefficient table.

```{r coefficient-summary}
coef_table <- summary(fit, tests = "coefficients", method = "auto")$coefficients
knitr::kable(coef_table, digits = 4)
```

The last column tells you the method used for each available p-value.
Here `"auto"` resolved to Satterthwaite, because the fit is interior and
the finite-sample route is feasible.

`inference_table()` shows a different, lower-level view: the inference
rows the engine cached at fit time. Those rows use asymptotic Wald *z* —
the engine's always-computable baseline — so their p-values are smaller
than the Satterthwaite ones above, and their `reliability` column says
`low`. This is the raw artifact, useful for audit; for reporting, use
`summary()` or `reporting_table()`, which resolve to the best feasible
method and agree with each other row for row.

```{r inference-table}
inference_table(fit)
```

```{r inference-checks, include = FALSE}
inf <- inference_table(fit)$table
stopifnot(any(inf$status == "available"))
stopifnot(all(is.finite(inf$p_value[inf$status == "available"])))
stopifnot(all(coef_table$method == "satterthwaite"))
# The Wald-z baseline p-values are smaller than the Satterthwaite ones.
p_col <- grep("^Pr\\(", names(coef_table), value = TRUE)[[1L]]
avail <- inf$status == "available"
satt_p <- coef_table[[p_col]][match(inf$term[avail], rownames(coef_table))]
stopifnot(all(inf$p_value[avail] <= satt_p + 1e-12))
```

## Contrasts

`contrast()` is the direct route when you know the fixed-effect comparison you
want. This contrast asks whether the coached program differs from usual care.

```{r contrast-test}
L <- c(0, 0, 1)
names(L) <- names(fixef(fit))

contrast(fit, L, method = "satterthwaite")
```

```{r contrast-checks, include = FALSE}
ct <- contrast(fit, L, method = "satterthwaite")
stopifnot(identical(ct$table$status, "available"))
stopifnot(is.finite(ct$table$p_value))
```

## Term tests

Use `test_effect()` for a named fixed-effect term.

```{r term-test}
test_effect(fit, "treatment", method = "kenward_roger")
```

Single-model `anova()` gives the same kind of term-level table.

```{r anova-test}
anova(fit, type = "III", method = "kenward_roger")
```

```{r term-checks, include = FALSE}
te <- test_effect(fit, "treatment", method = "kenward_roger")
av <- anova(fit, type = "III", method = "kenward_roger")
stopifnot(identical(te$table$status, "available"))
stopifnot(any(av$table$status == "available"))
stopifnot(is.finite(te$table$p_value))
```

## Model comparisons

For nested fixed-effect comparisons, fit the reduced model and compare it with
the full model.

```{r model-comparison}
reduced <- lmm(
  score ~ week + (1 | clinic),
  clinic_visits,
  control = mm_control(verbose = -1)
)

compare(reduced, fit)
```

`compare()` records that likelihood-ratio p-values are asymptotic. If you want
a simulation-based check for a small example, use the bootstrap path. The
bootstrap p-value is `(b + 1) / (successful + 1)`, where `b` counts
replicates at or above the observed statistic and `successful` counts the
refits that completed with a finite statistic — equal to `nsim` only when
every replicate succeeds, as it does in these seeded runs. So at
`nsim = 10` the smallest reportable value is 1/11 — use hundreds of
replicates before reading the p-value as a number rather than a bound.

```{r bootstrap-comparison}
compare(reduced, fit, method = "bootstrap", nsim = 10, seed = 7)
```

```{r comparison-checks, include = FALSE}
cmp <- compare(reduced, fit)
boot <- compare(reduced, fit, method = "bootstrap", nsim = 10, seed = 7)
stopifnot(is.finite(cmp$table$p_value[[2L]]))
stopifnot(identical(boot$table$method[[2L]], "parametric_bootstrap_lrt"))
stopifnot(boot$table$status[[2L]] %in% c("available", "not_assessed"))
if (identical(boot$table$status[[2L]], "available")) {
  boot_p <- boot$table$p_value[[2L]]
  stopifnot(is.finite(boot_p))
  # The p-value sits on the (b + 1) / (successful + 1) grid; all 10 seeded
  # replicates succeed here, so the denominator is 11.
  grid_pos <- boot_p * 11
  stopifnot(abs(grid_pos - round(grid_pos)) < 1e-8)
  stopifnot(boot_p >= 1 / 11 - 1e-12)
}
```

## When standard p-values are not well justified

Mixed models sometimes converge but still leave us in a difficult
inference situation. The fit exists, the estimates are usable, and
the model may be scientifically reasonable, but some familiar
p-values are no longer well justified.

The common case is a boundary or singular fit: a variance component
estimated at or near zero, or random-effect terms estimated as
perfectly correlated — equivalently, a fitted random-effect covariance
with lower effective rank than the model requested. What those labels
mean, how they are recorded on the object, and how singularity is
decided are covered in `vignette("convergence")`; here we take the
labels as given and ask what they do to inference.

The statistical background is well known. Self and Liang (1987) showed
that likelihood-ratio statistics do not necessarily follow their
usual chi-square reference distributions when a parameter lies on
the boundary of its parameter space. Stram and Lee (1994) specialized
this issue to variance components in linear mixed models. Kenward and
Roger (1997) derived their fixed-effect adjustment from a Taylor
expansion of the variance estimator; that derivation assumes a regular
interior solution for the variance parameters, and at a boundary those
derivatives may not exist or may not behave regularly. Parametric
bootstrap methods have a long history as the better-supported route in
this setting; Halekoh and Højsgaard (2014) implemented them in
`pbkrtest` for small-sample and boundary cases.

`mixeff` handles this by making the inference rules explicit. Asymptotic
rows are labeled with a reliability reason from a fixed vocabulary,
such as `asymptotic_wald_z_fallback` or `*_unavailable_at_boundary`,
and bootstrap inference is exposed as a labeled method rather than an
informal workaround. The goal is not to say that a singular fit is
automatically useless; it is to keep the covariance state attached to
the inference result, so that available numbers and unavailable numbers
are both documented.

This small repeated-measures example has subject-specific intercepts
and slopes that are nearly collinear by construction — enough to
produce a reduced-rank random-effect covariance estimate.

```{r singular-data, include = FALSE}
set.seed(3)
n_subj <- 18L
days <- 0:9
b0 <- rnorm(n_subj, sd = 30)
b1 <- 0.5 * b0 / 30 * 10 + rnorm(n_subj, sd = 0.5)

sleep_like <- do.call(rbind, lapply(seq_len(n_subj), function(i) {
  data.frame(
    subj = factor(i),
    days = days,
    rt = 250 + b0[i] + (10 + b1[i]) * days +
      rnorm(length(days), sd = 20)
  )
}))
```

```{r fit-singular}
sfit <- lmm(
  rt ~ days + (1 + days | subj),
  sleep_like,
  control = mm_control(verbose = -1)
)

fit_status(sfit)
is_singular(sfit)
```

```{r singular-checks, include = FALSE}
stopifnot(inherits(sfit, "mm_lmm"))
stopifnot(isTRUE(is_singular(sfit)))
stopifnot(identical(fit_status(sfit), "converged_reduced_rank"))
```

## The route map

Before testing a term, ask which inference routes are defined for this fit.
`inference_options()` is an audit map, not a recommender. Each row describes one
method that `mixeff` knows about. The row says whether the method can run, why
it can or cannot run, and what the rough computational cost is.

The printed object is the reader-facing map. It keeps the raw enum
columns in `routes$table` for scripts, but explains the route status
with display columns. The complete contract vocabulary is in the
[inference-method glossary](https://bbuchsbaum.github.io/mixeff/articles/inference-method-glossary.html).

```{r route-map}
routes <- inference_options(sfit, "days", nsim = 200)
routes
```

```{r route-checks, include = FALSE}
stopifnot("bootstrap" %in% routes$table$method)
satt <- routes$table[routes$table$method == "satterthwaite", ]
stopifnot(identical(satt$expected_status, "not_assessed"))
stopifnot(identical(satt$expected_reliability_reason,
                    "satterthwaite_unavailable_at_boundary"))
```

Read this table as follows. The Wald route can run immediately, but it
is labeled as a low-reliability asymptotic fallback. Satterthwaite and
Kenward–Roger are refused on this fit by mixeff's validated-support
policy: both approximations are derived under asymptotics that assume
the variance parameters lie in the interior of their parameter space,
and this fit has a variance parameter on the boundary. Other packages
will compute degrees of freedom here (`lmerTest` returns a number on
this same fit); `mixeff` declines to, and states the refusal as a
stable reason code — `satterthwaite_unavailable_at_boundary`,
`kenward_roger_unavailable_at_boundary` — you can test for. That is a
policy about what the package will certify, not a theorem that the
numbers other packages print are wrong.

The parametric bootstrap route can run and is the documented
fixed-effect testing route for this fit. The bootstrap LRT route is
refused here because this model was fit by REML, while that route
requires an ML fit (`bootstrap_lrt_requires_ml`). The cluster bootstrap
route is refused for fixed-effect p-values because its current target
is an estimator distribution, not a null hypothesis test. Profile
confidence intervals are not certified for this boundary fit
(`profile_ci_unavailable_at_boundary`).

```{r wald-status}
inference_table(sfit)$table[, c("term", "method", "status",
                                "reliability", "reliability_reason")]
```

`asymptotic_wald_z_fallback` on these rows is the engine's reason: a
*t* reference distribution was the target, the degrees of freedom could
not be computed at this boundary fit, and a standard normal was
substituted — which is why the row is graded `low` rather than hidden.
The *z* test drops the finite-sample correction, so treat its p-values
as optimistic rather than cautious.

```{r wald-status-checks, include = FALSE}
sinf <- inference_table(sfit)$table
stopifnot(all(sinf$reliability[sinf$status == "available"] == "low"))
stopifnot(all(sinf$reliability_reason[sinf$status == "available"] ==
                "asymptotic_wald_z_fallback"))
```

The important point is that `mixeff` distinguishes three cases:

1. A method can run and is considered available.
2. A method can produce a number, but the number is labeled as low reliability.
3. A method is refused because the method's assumptions or target are not
   defined for this fit.

That distinction is especially useful in scripts and reports, because
unavailable inference is recorded with stable reason codes rather than
disappearing as an error message.

## How do I test the term anyway?

For the fixed-effect term `days`, request the parametric bootstrap route
explicitly. Here the bootstrap target is a fixed-effect null model. `mixeff`
simulates data under the null hypothesis for the tested term, refits the model
to each simulated data set, and compares the observed test statistic with the
bootstrap reference distribution.

The small `nsim` value below keeps the vignette fast, and it earns a `low`
reliability grade: per the package's grade rubric (see the
inference-method glossary linked above), a certified bootstrap reaches
`moderate` at 999 or more successful replicates with a finite Monte
Carlo SE, and stays `low` below that floor. The reason to use bootstrap
here is not the grade — it is that the p-value rests on simulation
rather than on asymptotics that mixeff refuses to certify at the
boundary.

```{r bootstrap-term}
term_boot <- test_effect(
  sfit,
  "days",
  method = "bootstrap",
  bootstrap = bootstrap_control(nsim = 50, seed = 1)
)

term_boot$table[, c("term", "statistic_name", "p_value", "method",
                    "status", "reliability", "reliability_reason")]
```

```{r bootstrap-term-checks, include = FALSE}
stopifnot(identical(term_boot$table$status, "available"))
stopifnot(identical(term_boot$table$method, "bootstrap"))
stopifnot(is.finite(term_boot$table$p_value))
```

The p-value is returned as part of the labeled result row. It is not
reconstructed by hand in R.

The result also carries a bootstrap payload in `details`, including the
requested number of replicates, the number of successful refits, the boundary
rate among bootstrap refits, and the Monte Carlo standard error.

```{r bootstrap-payload}
run <- term_boot$table$details[[1]]$bootstrap
data.frame(
  requested_replicates = run$requested_replicates,
  successful_replicates = run$successful_replicates,
  boundary_rate = round(run$boundary_rate, 3),
  mcse = round(run$mcse, 4)
)
```

The boundary rate is not a failure count. It reports how often the bootstrap
refits also ended on a covariance boundary. That is useful diagnostic
information: it tells you that the boundary behavior is not just a one-off
feature of the observed data set.

With `nsim = 50`, the bootstrap p-value is also necessarily coarse. For example,
a p-value of about 0.0196 corresponds to the smallest non-zero value possible
under a common plus-one bootstrap adjustment, `1 / (successful + 1)` — here
all 50 replicates succeeded, as the accounting above shows, so 1/51. Use more
replicates when the p-value itself will be reported or compared with a decision
threshold.

## What about confidence intervals?

For fixed-effect confidence intervals, use `confint(method = "bootstrap")`.
This bootstrap has a different target from the fixed-effect test above. It
simulates from the full fitted model and summarizes the resulting estimator
distribution. That target is appropriate for percentile-style confidence
intervals and diagnostics, but it is not the same as a null distribution for a
fixed-effect p-value.

```{r bootstrap-ci}
ci <- confint(
  sfit,
  parm = "days",
  method = "bootstrap",
  bootstrap = bootstrap_control(nsim = 50, seed = 2)
)

ci
```

```{r bootstrap-ci-checks, include = FALSE}
stopifnot(all(is.finite(ci)))
stopifnot(identical(attr(ci, "method"), "bootstrap_full_model_distribution"))
payload <- attr(ci, "bootstrap")[[1]]
stopifnot(identical(payload$metadata$target$kind, "full_model_distribution"))
stopifnot(length(payload$replicate_statistics) >= payload$metadata$successful_replicates,
          length(payload$replicate_statistics) <= payload$metadata$completed_replicates)
```

The attached bootstrap payload records what was simulated and why the interval
should not be reinterpreted as a fixed-effect hypothesis test.

```{r ci-payload}
payload <- attr(ci, "bootstrap")[[1]]
payload$metadata$target$kind
payload$metadata$notes
```

This distinction is central:

- `test_effect(..., method = "bootstrap")` uses a fixed-effect null target and
  returns a p-value.
- `confint(..., method = "bootstrap")` uses a full-model estimator-distribution
  target and returns an interval.
- The second result should not be used to reverse-engineer the first.

## Where does cluster bootstrap stand?

Cluster resampling is available as an estimator-distribution target. It
resamples grouping levels and summarizes how estimates vary across those
resamples.

In the current schema, that target does not certify fixed-effect p-values.
Therefore a request for a fixed-effect cluster-bootstrap p-value refuses cleanly
instead of inventing a null distribution.

```{r cluster-refusal}
cluster_row <- test_effect(
  sfit,
  "days",
  method = "cluster_bootstrap",
  group = "subj"
)

cluster_row$table[, c("term", "method", "status", "p_value", "reason_code")]
```

```{r cluster-checks, include = FALSE}
stopifnot(identical(cluster_row$table$status, "not_assessed"))
stopifnot(is.na(cluster_row$table$p_value))
stopifnot(identical(cluster_row$table$reason_code,
                    "bootstrap_cluster_resample_p_value_unavailable"))
```

That refusal is part of the same inference contract as the available bootstrap
result. Available numbers are labeled with their method and reliability reason.
Unavailable numbers are labeled with stable reason codes explaining why they
are unavailable.

## Variance-component boundary tests

The boundary likelihood-ratio route is for random-effect variance components,
not ordinary fixed effects.

This is a different testing problem. For a variance component, the null
hypothesis can put the parameter exactly on the boundary of the parameter space,
because variances cannot be negative. In the simplest one-component case, the
*asymptotic* reference distribution is the Self-Liang 50:50 mixture,
`0.5 * chi-square(0) + 0.5 * chi-square(1)`. In finite samples the true
null distribution puts more mass at zero than the mixture, so the test
runs conservative (Crainiceanu and Ruppert 2004); with 18 subjects here,
read a marginal mixture p-value with that in mind.

Fit a simpler random-intercept model by ML:

```{r boundary-lrt}
ri_fit <- lmm(
  rt ~ days + (1 | subj),
  sleep_like,
  REML = FALSE,
  control = mm_control(verbose = -1)
)

re_lrt <- test_random_effect(ri_fit, "subj", method = "boundary_lrt")
re_lrt
```

```{r boundary-lrt-table}
reporting_table(re_lrt)$table[, c("term", "statistic", "p_value",
                                  "reference_distribution", "status")]
```

```{r boundary-lrt-checks, include = FALSE}
stopifnot(inherits(re_lrt, "mm_random_effect_test"))
stopifnot(identical(re_lrt$table$status, "available"))
stopifnot(identical(re_lrt$table$reference_distribution,
                    "0.5 * chi-square(0) + 0.5 * chi-square(1)"))
stopifnot(is.finite(re_lrt$table$p_value))
```

The reporting table keeps the reference distribution attached to the result.

Asking for the same method on a fixed effect returns a typed refusal.

```{r boundary-fixed-refusal}
fixed_boundary <- test_effect(ri_fit, "days", method = "boundary_lrt")
fixed_boundary$table[, c("term", "method", "status", "reason_code")]
```

Boundary likelihood-ratio theory applies to variance-component tests,
where the null hypothesis places a variance on the edge of the parameter
space. A fixed-effect null does not put any parameter on a boundary, so
the mixture distribution has nothing to correct there — which is why the
route refuses rather than running.

```{r boundary-fixed-checks, include = FALSE}
stopifnot(identical(fixed_boundary$table$status, "unsupported"))
stopifnot(identical(fixed_boundary$table$reason_code,
                    "boundary_lrt_not_applicable_to_fixed_effects"))
```

## Profile confidence intervals

Profile-likelihood intervals are available for variance parameters
under both fitting criteria, but fixed-effect (beta) profile intervals
require an ML fit: under the REML default, the beta rows are recorded
as unavailable with the reason code
`profile_beta_unavailable_under_reml`, not silently approximated.

```{r profile-ci}
prof <- confint(fit, method = "profile")
prof
attr(prof, "mm_profile")$table[
  ,
  c("parameter", "parameter_kind", "lower", "upper", "reason_code")
]
```

```{r profile-checks, include = FALSE}
ptab <- attr(prof, "mm_profile")$table
beta_rows <- ptab$parameter_kind == "beta"
stopifnot(any(beta_rows), any(!beta_rows))
stopifnot(all(is.na(ptab$lower[beta_rows])))
stopifnot(all(ptab$reason_code[beta_rows] ==
                "profile_beta_unavailable_under_reml"))
stopifnot(all(is.finite(ptab$lower[!beta_rows])),
          all(is.finite(ptab$upper[!beta_rows])))
```

Refit with `REML = FALSE` when you want beta profile intervals. On
boundary fits, profile intervals are refused outright
(`profile_ci_unavailable_at_boundary`), as the route map above showed.

## GLMM inference: the three-route contract

For generalized linear mixed models the contract is stricter, because
the default estimator is not certified for Wald inference. `glmm()`'s
default `method = "pirls_profiled"` is a fast profiled PIRLS fitter: it
returns point estimates, fitted values, and variance components, but
**refuses Wald standard errors, z statistics, p-values, and confidence
intervals on every route**. Its covariance payload is a working Hessian
the engine does not certify, and the engine's own documentation warns
that the profiled path is a different statistical approximation than
`glmer()`'s joint Laplace fit — it can be less accurate for
overdispersed models and models with observation-level random effects.

Exactly three routes lead to reportable GLMM inference, and every
refusal names them:

1. **Certified joint Wald.** Refit with `method = "joint_laplace"`;
   standard errors, z tests, and Wald intervals come from a covariance
   the engine certifies. `vignette("glmm")` is the operational
   quickstart for this route, including its parity with `glmer()` and
   its runtime notice.
2. **Parametric bootstrap on profiled fits.**
   `confint(fit, method = "bootstrap")` simulates from the fitted model
   and refits the same profiled estimator. It works for every supported
   family, including negative binomial — which has no joint-Laplace
   route, so the bootstrap is its only interval path — and costs `nsim`
   refits. On `joint_laplace` fits the bootstrap is refused at this
   engine pin; those fits have certified Wald intervals instead.
3. **Working-Hessian approximation, explicit opt-in.** For exploration
   and screening only. Every unlocked row is labeled
   `wald_z_working_hessian` with status `available_noninferential` and
   grade `moderate`; `inference_options(fit)` states the caveat and the
   invocation, and the aphantasia reproduction article
   (<https://bbuchsbaum.github.io/mixeff/articles/reproducing-aphantasia.html>)
   is the worked example of why this route is opt-in.

The practical rule: explore with the default, report with
`joint_laplace` or the bootstrap.

Here is the contract on a small simulated binomial GLMM. On the
default fit, `summary()` prints the estimates and withholds the
inference columns:

```{r glmm-data, include = FALSE}
set.seed(7)
g <- factor(rep(1:12, each = 12))
x <- rnorm(144)
eta <- -0.3 + 0.8 * x + rep(rnorm(12, sd = 0.5), each = 12)
trials <- data.frame(y = rbinom(144, 1, plogis(eta)), x = x, g = g)
```

```{r glmm-fit}
gfit <- glmm(
  y ~ x + (1 | g),
  trials,
  family = binomial(),
  control = mm_control(verbose = -1)
)

gtab <- summary(gfit, tests = "coefficients")$coefficients
gtab
```

A Wald `confint()` on the same fit is a typed refusal, not an error to
guess about:

```{r glmm-refusal}
err <- tryCatch(confint(gfit), error = identity)
class(err)[[1L]]
writeLines(strwrap(conditionMessage(err), width = 72))
```

The bootstrap route works on the same fit, and the interval carries its
accounting: requested, successful, and failed replicate counts,
replicate-SD standard errors, the Monte Carlo SE, and a reliability
grade. The replicate floor for a `moderate` grade is the rubric's
999-successful-replicates convention (see the glossary linked above);
the small, seeded run below stays `low` by that rubric.

```{r glmm-bootstrap-ci}
gci <- confint(gfit, method = "bootstrap", nsim = 30, seed = 42)
gci

acct <- attr(gci, "mm_bootstrap")
data.frame(
  requested = acct$requested,
  successful = acct$successful,
  failed = acct$failed,
  mcse = round(acct$mcse, 4),
  reliability = acct$reliability$reliability
)
attr(gci, "mm_method")
```

```{r glmm-checks, include = FALSE}
stopifnot(inherits(gfit, "mm_glmm"))
# Default path: estimates present, inference withheld (all SEs NA).
stopifnot(!any(is.finite(gtab[, "Std. Error"])))
# Wald confint refuses with a typed condition naming the alternatives.
stopifnot(inherits(err, "mm_inference_unavailable"))
stopifnot(grepl("joint_laplace", conditionMessage(err), fixed = TRUE))
stopifnot(grepl("bootstrap", conditionMessage(err), fixed = TRUE))
# Bootstrap interval: finite, ordered, with exact replicate accounting.
stopifnot(all(is.finite(gci)), all(gci[, 1L] < gci[, 2L]))
stopifnot(identical(acct$requested, 30L))
stopifnot(identical(acct$requested, acct$successful + acct$failed))
stopifnot(is.finite(acct$mcse))
stopifnot(identical(acct$reliability$reliability, "low"))
stopifnot(identical(attr(gci, "mm_method"),
                    "glmm_parametric_bootstrap_percentile"))
stopifnot(identical(attr(gci, "mm_estimator"), "pirls_profiled"))
```

The estimator mechanics — how the two estimators differ, the `nAGQ`
sensitivity check, and the joint-Laplace parity with `glmer()` — are
covered in `vignette("glmm")`. If a `joint_laplace` request does not
certify and the engine returns a labelled fallback fit, that
substitution is a typed record with its own notices; see
`vignette("convergence")`.

## Estimands: what are you averaging?

A fixed-effect coefficient is not a group mean. For a model with
multiple predictors and interactions, `fixef(fit)["trtactive"]` is the
treatment effect *at the reference level of every other predictor* —
one cell of the design, not a summary over it. Before averaging cells
into something more interpretable, name the estimand. Four quantities
are easy to conflate:

1. **A prediction at a covariate cell** — the model's fixed-effect
   prediction at one row of the reference grid. This is what
   `mm_predictions()` returns.
2. **An equal-weight reference-grid average** — the average of cell
   predictions over the grid, weighting every cell equally regardless
   of how many observations fall in it. This is what `mm_means()`
   returns, and what `mm_comparisons()` differences.
3. **An empirical-distribution average** — the average of predictions
   over the observed covariate rows, weighted as the data are
   weighted. `mixeff` has no dedicated verb for this; it is what you
   get by averaging `predict()` over the model frame.
4. **A causal average treatment effect** — the expected difference in
   potential outcomes under treatment versus control. Only this one is
   a causal quantity, and no mixed-model contrast supplies it by
   itself: it requires identification assumptions (randomization, or
   conditional ignorability given the model's covariates) that are
   properties of the study design, not of the fit.

The first three are model-based summaries that differ only in how they
weight covariate cells; the marginal verbs below compute (1) and (2)
with standard errors from the fixed-effect covariance. Whether any of
them deserves a causal reading is a question about the design, and the
fitted model does not certify it.

`mixeff` computes these with four functions — `mm_grid()`,
`mm_predictions()`, `mm_means()`, `mm_comparisons()` — that route all
inference through the same machinery as `contrast()`. Each row in the
returned table carries `method`, `status`, `reliability`, and `reason`
columns naming how it was computed and how far to trust it (the column
vocabularies are in the glossary linked above).

### The study

A rehabilitation trial assigns patients to coached or usual-care treatment.
Each patient is assessed before and after the intervention. Patients are
grouped within clinics.

```{r rehab-data, include = FALSE}
set.seed(101)
n_rclinic <- 4L
n_rsubj   <- 24L
spc       <- n_rsubj %/% n_rclinic   # subjects per clinic

rclinic <- factor(rep(paste0("C", seq_len(n_rclinic)), each = spc))
# Deliberately unbalanced assignment: C1 is mostly control, C3 mostly
# active. The imbalance is the point of this section.
n_control <- c(5L, 3L, 1L, 3L)
trt <- factor(
  unlist(lapply(seq_len(n_rclinic), function(i) {
    c(rep("control", n_control[[i]]), rep("active", spc - n_control[[i]]))
  })),
  levels = c("control", "active")
)
rsubj <- factor(paste0("S", seq_len(n_rsubj)))

rehab <- data.frame(
  subj   = rep(rsubj,   each = 2L),
  clinic = rep(rclinic, each = 2L),
  trt    = rep(trt,     each = 2L),
  time   = factor(rep(c("pre", "post"), times = n_rsubj),
                  levels = c("pre", "post"))
)

rclinic_re <- c(-1.2, 0.4, 0.9, -0.1)
rsubj_re   <- rnorm(n_rsubj, sd = 0.8)

rehab$score <- 52 +
  -2.1 * (rehab$trt  == "active") +
  -3.8 * (rehab$time == "post") +
  -2.9 * (rehab$trt  == "active") * (rehab$time == "post") +
  rclinic_re[as.integer(rehab$clinic)] +
  rsubj_re[as.integer(rehab$subj)] +
  rnorm(nrow(rehab), sd = 0.4)
```

```{r show-rehab}
head(rehab)
```

Clinics are not balanced between treatment arms: clinic C1 has five
control patients and one active, clinic C3 the reverse. So raw group
means are confounded with clinic effects — each arm's mean leans on the
clinics that arm over-samples, and the clinics differ — and a naive
`tapply(rehab$score, rehab$trt, mean)` will not answer the treatment
question.

```{r rehab-fit}
rfit <- lmm(
  score ~ trt * time + (1 | clinic) + (1 | subj),
  rehab,
  control = mm_control(verbose = -1)
)
summary(rfit, tests = "coefficients")
```

The interaction coefficient `trtactive:timepost` tells you the
*additional* post-treatment change for the active arm relative to
control — a cell-specific quantity, estimand (1) territory. Summaries
over cells are the marginal verbs' job.

### Reference grids with `mm_grid()`

`mm_grid()` constructs the cross-product of all fixed-predictor levels. By
default, factor predictors expand to all their levels; numeric predictors
collapse to their mean.

```{r grid}
grd <- mm_grid(rfit, specs = ~ trt * time)
grd
```

The grid has four rows — one for each treatment × timepoint cell — and
retains the model matrix needed for inference.

### Cell predictions with `mm_predictions()`

`mm_predictions()` evaluates the fixed-effect prediction at each grid row,
with a confidence interval from the certified covariance. These are
estimand (1): the four population-level cell means.

```{r predictions}
preds <- mm_predictions(rfit, specs = ~ trt * time)
preds$table[, c("label", "estimate", "conf_low", "conf_high", "method")]
```

Each row carries its inference method so the provenance is visible
without digging into model objects.

### Marginal means with `mm_means()`

Marginal means average the reference grid over dimensions you want to
collapse — estimand (2), the equal-weight reference-grid average. Here:
average over timepoints to get a treatment-level summary.

```{r means-trt}
mt <- mm_means(rfit, specs = ~ trt)
mt$table[, c("label", "estimate", "conf_low", "conf_high", "method")]
```

Compare these to the raw means:

```{r raw-means}
raw_means <- tapply(rehab$score, rehab$trt, mean)
raw_means

marg_means <- setNames(mt$table$estimate, mt$table$label)
imbalance_pull <- max(abs(
  raw_means[c("control", "active")] -
    marg_means[c("trt=control", "trt=active")]
))
round(imbalance_pull, 2)
```

The clinic imbalance pulls each raw mean `r round(imbalance_pull, 2)`
points toward the clinics its arm over-samples; the marginal means
average over the model's fixed-effect grid instead and are free of that
shift. The size of the gap depends on how unbalanced the design is and
how large the group effects are.

### Pairwise comparisons with `mm_comparisons()`

`mm_comparisons()` takes all pairwise differences among the marginal means
and applies the same inference method.

```{r comparisons-trt}
ctr <- mm_comparisons(rfit, specs = ~ trt)
ctr$table[, c("label", "estimate", "conf_low", "conf_high", "p_value", "method")]
```

The `active - control` row is the equal-weight reference-grid contrast
— estimand (2) in the list above, averaged over both timepoints —
tested with a Satterthwaite *t* and graded `moderate` reliability in
the same row. In a randomized design you may argue it also estimates
the causal quantity (4); that argument rests on the design's
identification assumptions, and it is yours to make — the fitted model
does not certify it.

```{r marginal-checks, include = FALSE}
stopifnot(inherits(grd, "mm_grid"), nrow(grd$grid) == 4L)
stopifnot(nrow(preds$table) == 4L)
stopifnot(all(is.finite(preds$table$estimate)))
stopifnot(nrow(mt$table) == 2L)
stopifnot(all(is.finite(mt$table$estimate)))
stopifnot(is.finite(imbalance_pull), imbalance_pull > 0)
cmp_row <- ctr$table[ctr$table$label == "trt=active - trt=control", ]
stopifnot(nrow(cmp_row) == 1L)
stopifnot(identical(cmp_row$method, "satterthwaite"))
stopifnot(is.finite(cmp_row$p_value))
```

### Conditional comparisons with `by =`

The `by` argument splits comparisons within levels of another variable —
the analogue of simple effects in a factorial design.

```{r comparisons-by-time}
ctr_by <- mm_comparisons(rfit, specs = ~ trt | time)
ctr_by$table[, c("label", "estimate", "conf_low", "conf_high", "p_value", "method")]
```

Two rows: the treatment difference *at pre-intervention* and the treatment
difference *at post-intervention*. The post-intervention gap is larger because
the interaction drives additional improvement in the active arm.

### Constraining the grid with `at =`

`at` pins predictors to chosen values instead of expanding or averaging
them: a factor to a subset of its levels, a numeric predictor to specific
points rather than its mean.

```{r at-example}
mt_time <- mm_means(rfit, specs = ~ time, at = list(trt = "active"))
mt_time$table[, c("label", "estimate", "conf_low", "conf_high")]
```

This gives the pre/post means *within the active arm* only, holding `trt`
constant at `"active"`.

### Custom contrasts with `mm_lincomb()`

For hypotheses that are not pairwise differences of marginal means, build
the contrast weights directly with `mm_lincomb()`. The interaction effect
expressed as a contrast: (active post − active pre) − (control post − control pre).

```{r lincomb}
beta <- fixef(rfit)
names(beta)

# Interaction row: coefficient named "trtactive:timepost" (lme4-identical)
w <- setNames(numeric(length(beta)), names(beta))
w["trtactive:timepost"] <- 1

lc <- mm_lincomb(rfit, weights = w)
lc[, c("estimate", "lower", "upper", "p_value", "method")]
```

```{r lincomb-checks, include = FALSE}
stopifnot(is.finite(lc$estimate), is.finite(lc$p_value))
stopifnot(nzchar(lc$method))
```

`mm_lincomb()` applies the same contract-preserving inference as `contrast()`,
so the method and status fields are populated identically. A row whose
status is anything other than `"available"` contains `NA` where the
refused numbers would go, with a stable `reason` code saying why; the
code vocabulary is in the glossary linked above. For the `emmeans`
bridge — and the audit columns that do *not* survive the trip through
`emmeans` — see `vignette("lme4-migration")`.

## The covariance contract

The quantities above all lean on the fixed-effect covariance, and
`vcov()` is explicit about what it returns. For full-rank fits it is
the model-based covariance from the engine's
`fixed_effect_covariance_matrix` payload, with method and status
attributes attached:

```{r vcov-contract}
V <- vcov(fit)
V
attr(V, "mm_method")
attr(V, "mm_status")
```

For rank-deficient or otherwise uncertified fits, `vcov()` returns an
`NA` matrix carrying an `mm_unavailable_reason` attribute instead of
fabricated numbers — the same available/unavailable/typed-error
contract as every other surface in this vignette. Prediction standard
errors and intervals (population versus conditional, and why the
conditional SE can be *smaller*) are documented in `vignette("lmm")`.

```{r vcov-checks, include = FALSE}
stopifnot(all(is.finite(V)))
stopifnot(isTRUE(all.equal(V, t(V))))
stopifnot(identical(attr(V, "mm_status"), "available"))
stopifnot(is.null(attr(V, "mm_unavailable_reason")))
```

## Summary

On an interior LMM fit, `summary()`, `contrast()`, `test_effect()`,
`anova()`, and `compare()` return labeled finite-sample inference, with
`"auto"` resolving to Satterthwaite.

On the boundary fit in this vignette:

- the model converges, but the random-effect covariance is reduced rank
  (`vignette("convergence")` for what that label establishes);
- Wald inference is available but labeled as low reliability;
- Satterthwaite and Kenward–Roger routes are refused by mixeff's
  validated-support policy: their degrees-of-freedom approximations
  assume interior variance parameters, which this fit does not have;
- parametric bootstrap testing is available for the fixed effect;
- bootstrap confidence intervals are available, but they use a
  full-model estimator-distribution target, not a fixed-effect null
  target;
- cluster bootstrap currently reports estimator distributions, not
  fixed-effect p-values;
- boundary LRTs are available for variance-component tests, not
  fixed-effect tests.

For GLMMs, three routes lead to reportable inference: certified joint
Wald (`method = "joint_laplace"`), the parametric bootstrap on profiled
fits, and the labeled working-Hessian opt-in for screening.

Every route either returns a labeled result or refuses with a stable
reason.

## References

- Crainiceanu, C. M., and Ruppert, D. (2004). Likelihood ratio tests
  in linear mixed models with one variance component. *Journal of the
  Royal Statistical Society: Series B*, 66(1), 165–185.
  <https://doi.org/10.1111/j.1467-9868.2004.00438.x>
- Halekoh, U., and Højsgaard, S. (2014). A Kenward-Roger approximation
  and parametric bootstrap methods for tests in linear mixed models —
  the `R` package `pbkrtest`. *Journal of Statistical Software*,
  59(9), 1–32. <https://doi.org/10.18637/jss.v059.i09>
- Kenward, M. G., and Roger, J. H. (1997). Small sample inference for
  fixed effects from restricted maximum likelihood. *Biometrics*,
  53(3), 983–997. <https://doi.org/10.2307/2533558>
- Self, S. G., and Liang, K.-Y. (1987). Asymptotic properties of
  maximum likelihood estimators and likelihood ratio tests under
  nonstandard conditions. *Journal of the American Statistical
  Association*, 82(398), 605–610.
  <https://doi.org/10.1080/01621459.1987.10478472>
- Stram, D. O., and Lee, J. W. (1994). Variance components testing in
  the longitudinal mixed effects model. *Biometrics*, 50(4),
  1171–1177. <https://doi.org/10.2307/2533455>
