---
title: Migrating from lme4
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from lme4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
has_lme4 <- requireNamespace("lme4", quietly = TRUE)
```

```{r load}
library(mixeff)
```

```{r sleepstudy-helper, include = FALSE}
# vignette-local convenience so chunks run whether or not lme4 is attached;
# use lme4::sleepstudy directly in your own code.
sleepstudy_data <- function() {
  if (requireNamespace("lme4", quietly = TRUE)) {
    e <- new.env()
    utils::data("sleepstudy", package = "lme4", envir = e)
    e$sleepstudy
  } else {
    set.seed(1)
    data.frame(
      Reaction = rnorm(180, 300, 50),
      Days = rep(0:9, 18),
      Subject = factor(rep(seq_len(18), each = 10))
    )
  }
}
```

`mixeff` fits lme4-style formulas with the familiar extractors and, on
its supported envelope, gives statistical answers tested against pinned
lme4 references within documented tolerances. It is not a literal drop-in —
you call `lmm()` / `glmm()` rather than `lmer()` / `glmer()`, and requests
outside the envelope are changed or declined with a diagnostic, never
silently. This vignette is the verb-for-verb,
argument-for-argument map; the generated tables below are the envelope's
exact boundary. It is the migration entry in the package's six-vignette
set: the support contract itself is stated in
`vignette("mixeff", package = "mixeff")`, the fitting workflows in
`vignette("lmm", package = "mixeff")` and
`vignette("glmm", package = "mixeff")`, convergence and boundary semantics
in `vignette("convergence", package = "mixeff")`, and the inference
contract in `vignette("inference", package = "mixeff")`.

## The two edits

An `lmer` script becomes an `lmm` script with two changes: the fitting verb and
the control object.

```{r fit, eval = TRUE}
fit <- lmm(Reaction ~ Days + (Days | Subject), sleepstudy_data(),
           control = mm_control(verbose = -1))
fixef(fit)
```

```{r fit-lme4, eval = has_lme4}
m <- lme4::lmer(Reaction ~ Days + (Days | Subject), lme4::sleepstudy)
lme4::fixef(m)
```

(`sleepstudy_data()` above just returns `lme4::sleepstudy` when lme4 is
installed; use `lme4::sleepstudy` directly in your own code.)

## Verb map

| lme4 | mixeff | Notes |
|------|--------|-------|
| `lmer(y ~ x + (x \| g), data)` | `lmm(y ~ x + (x \| g), data)` | same formula language, incl. `(x\|\|g)`, `(1\|g1/g2)`, crossed; fixed-part `factor()`, `poly()`, `scale()`, `ns()`/`bs()`, `cut()`, `relevel()`, `offset()`, `^`, `%in%` and `(a + b)^2` work with lme4's coefficient names (evaluated in R and sent to the engine as columns); such transforms inside a random-effect bar are refused |
| `glmer(y ~ ..., family = binomial)` | `glmm(y ~ ..., family = binomial())` | pass a family **object** (`binomial()`), not a string |
| `lmerControl(...)` / `glmerControl(...)` | `mm_control(optimizer=, max_feval=, verbose=, start=, optCtrl=, disp_method=, disp_dof_correction=, max_phi_iter=)` | `optCtrl` names `maxfun`/`maxeval`, `ftol_*`, `xtol_*`, `rhobeg` are mapped (others refused); no `check.conv.*` (no post-hoc convergence warnings): use `optimizer_certificate()` / `verify_convergence()`; lme4 2.1-0's GLMM dispersion controls map one to one: `disp_method = "moment"`/`"old/buggy"`, `disp_dof_correction`, and `maxPhiIter` -> `max_phi_iter` (defaults `"moment"`, `TRUE`, `100`, as lme4 2.1-0); they only affect Gamma, inverse-Gaussian and Gaussian non-identity-link fits (elsewhere a `mm_control_ignored_notice` says they were ignored; lme4 is silent) and carry over to `refit()`/`update()` |
| `fixef`, `vcov` | same generics, same values | except for Gamma, inverse-Gaussian and Gaussian non-identity-link GLMMs where `vcov()` comes from RX (`nAGQ = 0` or the Hessian fallback): lme4 2.1-0 multiplies by `sigma()^2` although its weights already carry `1/phi`; mixeff reports the unscaled RX covariance (same for `predict(se.fit = TRUE)`) |
| `sigma` | identical | 1 for binomial, Poisson and negative-binomial GLMMs (the NB theta is `getME(fit, "glmer.nb.theta")`); Gamma, inverse-Gaussian and Gaussian non-identity-link GLMMs report `sqrt(phi)` with `phi = deviance / (n - rank([X, Z]))`, as lme4 >= 2.1-0 (older lme4 used a different estimate, so those values, and their `logLik()`, theta and `VarCorr()`, match lme4 >= 2.1-0 only) |
| `ranef` | same generic | `condVar = TRUE` for LMMs and GLMMs (GLMM conditional variances from the Laplace approximation at the fitted modes, as lme4; for Gamma, inverse-Gaussian and Gaussian non-identity-link GLMMs lme4 2.1-0 multiplies them by an extra `sigma()^2`, mixeff does not) |
| `coef` | identical | every fixed-effect column per group (fixed-only columns repeated), lme4's column order |
| `VarCorr` | identical shape | named list of covariance matrices with `stddev`/`correlation` attributes and `sc`/`useSc`, so `VarCorr(m)$Subject` is the matrix; prints like lme4 (plus `[boundary]` markers); mixeff's long table is still `VarCorr(m)$table` (and `$residual_sd`); `as.data.frame()` matches lme4; GLMMs have `useSc = FALSE` (no Residual row, absolute SDs) with `sc = sigma()`, as lme4 >= 2.1-0 |
| `logLik`, `AIC`, `BIC`, `nobs` | identical | `AIC(m1, m2)` / `BIC(m1, m2)` return stats' `df`/`AIC` data frame |
| `deviance` | identical | LMM: as lme4 for ML fits (REML fits return the REML criterion without lme4's deprecation warning); GLMM: sum of squared deviance residuals, as lme4 |
| `REMLcrit` | `REMLcrit()` | exported by mixeff (lme4's is not a generic; whichever package is attached last masks the other, and mixeff's forwards lme4 fits to lme4) |
| `family`, `weights`, `model.frame` | identical | `weights()` returns ones for unweighted fits and `type = "working"` for GLMMs; `model.frame()` has lme4's transformed columns and `terms` attribute (the raw variables the engine used are `fit$model_frame`) |
| `confint` | same generic, **different default** | Wald (`"asymptotic"`, lme4 spelling `"Wald"`) by default, where lme4 defaults to profile; returns fixed effects only. `"profile"` (LMMs) returns lme4's rows (`.sig01`, ..., `.sigma`, fixed effects; REML fits profiled on the ML deviance, as lme4), and accepts `threads =`; `"bootstrap"` (lme4 spelling `"boot"`); GLMM availability per fit type is the route matrix in `vignette("inference", package = "mixeff")` |
| `predict`, `fitted` | identical | `re.form = NULL/NA/~0` or a partial formula such as `~(1 \| g)` (no `se.fit`/`interval` for partial ones), `allow.new.levels`, `na.action` (default `na.pass`), `se.fit`, `interval`; `newdata` is narrowed to the needed variables and coerced like the training frame; an `offset =` fit offset must be resupplied with `predict(offset =)` (lme4 silently uses 0) |
| `residuals` | identical | LMM default `"response"`, `"pearson"`/`"deviance"` are `sqrt(w) * (y - mu)`; GLMM default `"deviance"`, plus `"pearson"`, `"working"`, `"response"`; `scaled = TRUE` divides by `sigma()` once |
| `simulate`, `refit` | LMM and GLMM | lme4's `re.form`/`use.u` meaning (default `NA`: new random effects; `NULL`: condition on the BLUPs) and RNG order, so seeded draws match lme4; LMM draws honour prior weights (lme4 ignores them); Gamma (shape `w/phi`), Gaussian non-identity-link and inverse-Gaussian fits draw random effects on the absolute scale reported by lme4 >= 2.1-0, and Gamma and unweighted Gaussian draws match lme4 2.1-0's; Gaussian draws honour prior weights (lme4 ignores them); inverse-Gaussian draws use shape `w/phi` (needs statmod), where lme4 2.1-0 uses `w/sigma` and so simulates variance `sigma*mu^3` instead of the fitted `phi*mu^3`; a factor binomial response simulates as 0/1; a partial `re.form` such as `~(1 \| g)` conditions on those terms and redraws the others, as lme4; `newdata` and `newparams` are refused |
| `plot(fit)`, `dotplot(ranef(fit, condVar = TRUE))`, `qqmath(ranef(...))` | base-graphics `plot()`/`qqnorm()`; lattice methods | `plot(fit)` draws Pearson residuals vs fitted with base graphics (no formula interface); `plot(ranef(fit, condVar = TRUE))` gives a caterpillar plot |
| `hatvalues`, `cooks.distance`, `influence` | same generics | LMM values match lme4; GLMM hat values use the IRLS working weights (as `glm` does), so they differ from lme4's; `influence()` takes `groups=` and `ncores=` and records failed refits instead of stopping |
| `performance::r2()`, `performance::icc()`, `insight::get_variance()` | same calls, or `mm_r2()`, `mm_icc()`, `mm_variance_components()` | Nakagawa R2/ICC agree with performance on lme4 fits; `by_group = TRUE` is not supported |
| `update(fit, . ~ . - x)` | same generic | a formula that needs new columns re-evaluates the original `data`, as lme4 does; arguments `lmm()`/`glmm()` cannot carry over (e.g. `subset`, `contrasts`) are refused rather than dropped; the fit's `na.action` carries over |
| `anova(m1, m2)`, `drop1` | LRT with automatic ML refit | `anova(m1, m2)` is lme4's data frame (`npar AIC BIC logLik -2*log(L) Chisq Df Pr(>Chisq)`, as lme4 >= 2.0); single-model `anova()` is lmerTest's (`Sum Sq Mean Sq NumDF DenDF F value Pr(>F)`, Type I/II/III, Satterthwaite/KR) and LMM-only; mixeff's provenance table is `$table`; a multi-column term such as `poly(x, 2)` is tested as one multi-df row, like lmerTest, unless another term contains it (then per column) |
| `getME` | `getME()` | LMM and GLMM; lme4's components including `u`, `b`, `L`, `RX`, `RZX`, `devcomp`, `lower`, `Gp`, `Ztlist`, `Tlist`, `ST`, `glmer.nb.theta`, `"ALL"`; `devfun` is refused (the optimizer lives in the engine) |
| `isSingular` | `is_singular(fit, tol = 1e-4)` | lme4's rule (a zero-bounded theta below `tol`), LMM and GLMM; `lme4::isSingular()` is not a generic, so call `is_singular()` |
| `ngrps` | identical | interaction/nested groups are named as in lme4 (`a:b`; `b:a` for the nested term of `a/b`) with levels `x:y` |
| `broom.mixed::tidy/glance/augment` | supported | registered for `mm_lmm`/`mm_glmm` |
| `emmeans::emmeans(fit, ~ x)` | supported | estimates and SEs agree with lme4's; mixeff's status/reliability/reason audit columns do not survive the emmeans trip, and a profiled (`pirls_profiled`) GLMM refuses inference through emmeans |
| `lmerTest` p-values in `summary()` | built in | Satterthwaite/Kenward-Roger native, no extra package; `summary()$coefficients` is lmerTest's matrix (per-row method labels in `summary()$coef_table`); lmerTest's `ddf = "Satterthwaite"`/`"Kenward-Roger"`/`"lme4"` is accepted by `summary()` and `anova()` |
| `bootMer(m, FUN, nsim)` | `mm_bootmer(fit, FUN, nsim)` | LMM; a `boot`-compatible result for `boot::boot.ci()`; `type = "semiparametric"` refused |
| `rePCA`, `lmList`, `allFit` | `rePCA()`, `mm_lmlist()`, `mm_allfit()` | `mm_*` names avoid masking lme4's non-generic functions |
| `pbkrtest::KRmodcomp(big, small)` | `compare(small, big, method = "kenward_roger")` | also `"satterthwaite"` (`SATmodcomp`); KR F is pbkrtest's scaled `Ftest` row (unscaled `FtestU` in `$fixed_f$unscaled_statistic`) |
| `lmerTest::step(m)` | `step(fit)` | random terms tested with the boundary-corrected LRT; whole terms only |
| `nlmer`, `devFunOnly = TRUE` | not provided | nonlinear mixed models are a non-goal; the engine owns the objective |

## Argument map for `lmm()` / `glmm()`

| lme4 argument | mixeff | Notes |
|---------------|--------|-------|
| `REML` | `lmm(..., REML=)` | same |
| `weights` | `weights=` | a column of `data` or a vector, as in lme4 (LMM and GLMM) |
| `offset` | `offset=` | a column of `data`, an expression, or a vector (LMM and GLMM); `offset()` formula terms are added to it. LMMs fit `y - offset` and add it back in `fitted()`/`predict()` |
| `subset` | `subset=` | LMM and GLMM; formula transforms are evaluated before subsetting, as in `model.frame()`; unused factor levels are dropped afterwards |
| `na.action` | `na.action=` | LMM and GLMM; default `getOption("na.action")` (`na.omit`) as in lme4; only model variables, `weights` and `offset` count; dropped rows are announced (`mm_rows_dropped` message) and recorded in `na.action(fit)`; `na.exclude` pads `fitted()`/`residuals()`/`predict()`; `na.fail`/`na.pass` raise `mm_data_error` |
| `contrasts` | partial | unordered factors use treatment coding, ordered factors `contr.poly` (both matching R/lme4 defaults). Any other coding is refused with an `mm_arg_error` rather than silently replaced, whether it is requested through `contrasts=`, a contrast attached to the factor (`contrasts(f) <- ...`), or `options(contrasts=)`; use `relevel()` to change the reference level, or build coded columns as numeric predictors |
| character predictors | converted to factors | sorted levels, as lme4's `factor()` conversion does; logical predictors are coded as factors with levels `FALSE`/`TRUE`, as `model.matrix()` does (e.g. `lFALSE`, `lTRUE` without an intercept) |
| numeric/character grouping (accepted silently) | coerced to factor | non-factor grouping columns are coerced to factor, announced via an `mm_grouping_coercion_notice`; silence it with `mm_control(verbose = -1)`; an ordered grouping factor is used as an unordered one (grouping needs only its levels) |
| `family = "binomial"` | `family = binomial()` | string families are not accepted |
| `nAGQ` | `glmm(..., nAGQ=)` | `1` (default) is joint Laplace, as in `glmer`; `nAGQ = 0` selects the profiled fast path (`method = "pirls_profiled"`, lme4's PIRLS-only estimate); `>1` runs on the profiled path only and, as in lme4, needs a single scalar random effect (selected with an `mm_estimator_notice` when `method` is not given) |
| `control = lmerControl(optimizer=, optCtrl=)` | `mm_control(optimizer=, max_feval=, optCtrl=, ...)` | the engine picks a default optimizer; `mm_control()` can override it or cap the evaluation budget |
| `start` | `mm_control(start=)` | theta warm starts |

## Four differences to check when porting

**1. Coefficient names match lme4 exactly.** Since 0.2.0, `fixef()`,
`summary()` tables, `vcov()` dimnames, and `mm_lincomb()` weight names use
lme4's naming and column order (`"recipeB"`, `"temperature.L"`,
`"recipeB:temperature.L"`), so lme4 code keyed on coefficient names carries
over without renaming.
(Earlier versions used an engine encoding like `"recipe: B"`; if you wrote
normalization shims for those, delete them.)

**2. Grouped binomial responses.** `glmm()` accepts the
`cbind(successes, failures)` spelling like `glmer`:

```{r cbind, eval = FALSE}
glmm(cbind(incidence, size - incidence) ~ period + (1 | herd),
     lme4::cbpp, family = binomial())
```

**3. The GLMM estimator.** `glmm()` defaults to `method = "joint_laplace"`,
glmer's joint Laplace (`nAGQ = 1`) estimator, with certified Wald inference.
The fast profiled (PIRLS) estimator — lme4's `nAGQ = 0` — is opt-in; its
coefficients do **not** match `glmer(nAGQ = 1)` exactly. Negative-binomial
families (no joint route yet), `nAGQ > 1`, and
`inference = "working_hessian"` need the profiled path: without an explicit
`method`, `glmm()` selects it and prints an `mm_estimator_notice` saying so.

```{r joint, eval = FALSE}
glmm(y ~ x + (1 | g), data, family = binomial())   # joint Laplace (glmer)
glmm(y ~ x + (1 | g), data, family = binomial(),
     method = "pirls_profiled")                      # fast path (nAGQ = 0)
```

**4. `||` expands exactly like lme4's, factors included.** mixeff expands
`(1 + x + f || g)` the way lme4's `expandDoubleVerts()` does:
`(1 | g) + (0 + x | g) + (0 + f | g)` — the numeric columns get
independent variances, while the factor `f` keeps its full within-factor
covariance block, coded like R's `model.matrix()` (so `(0 + f + h | g)` is
full rank, as in lme4). The parameter count, `df`, AIC and the optimum
therefore match lme4, and `VarCorr()` lists the pieces as separate entries
named `g`, `g.1`, ... exactly as lme4 does (also in
`as.data.frame(VarCorr(fit))`). To fix *every* covariance at zero —
including those among a factor's columns (MixedModels.jl's `zerocorr()`) —
write `diag(1 + f | g)`, which has no lme4 equivalent. mixeff still notes
a factor inside `||` at compile time with an info diagnostic
(`covariance_assumption`, reason `double_bar_factor_term`).

```{r doublebar-factor, eval = FALSE}
# lme4 semantics: independent intercept and x slope; the factor keeps its block
glmm(y ~ cond * x + (1 + cond + x || subj), data, family = binomial())
# same model, written out
glmm(y ~ cond * x + (1 | subj) + (0 + x | subj) + (0 + cond | subj),
     data, family = binomial())

# MixedModels.jl zerocorr semantics: every covariance fixed at zero
glmm(y ~ cond * x + diag(1 + cond + x | subj), data, family = binomial())
```

Comparing a `||` model with its correlated `|` counterpart adds only
correlation parameters; the likelihood-ratio test (and
`test_random_effect()`'s boundary route) then uses the ordinary
chi-square, as `anova()` on lme4 fits does.

**5. Random-effect terms are presented in lme4's order.** `ranef()`,
`VarCorr()`, `getME()` (`theta`, `Lambdat`, `Zt`, `cnms`, ...) and the
`.sig01`, ... rows of `confint(method = "profile")` use lme4's `mkReTrms`
term order (formula order, reordered by decreasing number of levels only
when a later term has more levels than an earlier one). The engine fits
the same model with its own internal (MixedModels.jl) order; the
translation is exact.

**6. Profile intervals use lme4's scale and names, but Wald is the
default.** `confint(fit, method = "profile")` returns lme4's rows
(`.sig01`, ..., `.sigma`, then the fixed effects) on the
standard-deviation/correlation scale, numbered as lme4 >= 2.0 numbers them
(term by term, each term's standard deviations before its correlations;
lme4 1.1 interleaved them in lower-triangle order), and, like lme4,
profiles a REML fit on
the ML deviance. lme4's `confint()` defaults to the profile; mixeff's
default stays the Wald interval (the profile refits the model many times),
so pass `method = "profile"` to reproduce lme4's default output.
Bootstrap and profile computations accept `threads =` (via
`bootstrap_control(threads = )` for bootstraps); results are identical for
every thread count, and the default of 1 respects CRAN's two-thread policy.

**7. Rank-deficient fixed effects.** As in lme4, of two collinear columns
the earlier is kept; `fixef()` omits the dropped coefficient (use
`fixef(fit, add.dropped = TRUE)` for an `NA` in its place), and `vcov()` and
`summary()` cover the estimable coefficients only.

## The supported envelope, generated

The tables below are rendered from the package's machine-readable support
registry (`inst/support/model-support.json`, via `supported_models()`), so
this vignette cannot drift from what the code enforces — the test suite
holds the fitting code, this registry, and the `glmm()` documentation to
the same envelope.

```{r support-matrix, echo = FALSE}
sm <- supported_models()
knitr::kable(sm$glmm_families,
             col.names = c("family", "links", "estimators", "notes"),
             caption = "Supported GLMM family/link/estimator cells.")
knitr::kable(sm$features[, c("feature", "lmm", "glmm", "notes")],
             caption = "Feature support (see notes for mixeff-specific semantics).")
```

```{r support-matrix-checks, include = FALSE}
stopifnot(nrow(sm$glmm_families) >= 4L, nrow(sm$features) >= 5L)
stopifnot(identical(
  sort(sm$glmm_families$family),
  sort(names(mixeff:::mm_glmm_supported_family_links()))
))
```

Explicit non-goals: `r paste(sm$non_goals, collapse = "; ")`.

## What is `NA`-with-a-reason (and why)

Where an inference method is unavailable, `mixeff` returns `NA` with a
machine-readable reason or raises a typed condition, rather than printing a
number without comment:

| Situation | lme4 (+lmerTest) | mixeff |
|-----------|------|--------|
| `NA` in a model variable | silently dropped (`na.omit`) | dropped under the same `na.action`, announced with a typed `mm_rows_dropped` message and recorded in `na.action(fit)` |
| Boundary (singular) fit | one-time warning | persistent `[boundary]` tag + effective rank |
| Satterthwaite df at a boundary | lmerTest prints df anyway | refused with a reason; use bootstrap |
| Prediction SE for an unseen grouping level | — | `NA` with the reason in `mm_reason` |
| GLMM `confint()` | computed | Wald intervals on `joint_laplace` fits via `method = "asymptotic"` (lowercase; lme4's `method = "Wald"` spelling is not accepted); parametric-bootstrap intervals via `method = "bootstrap"` on profiled-estimator fits (like `bootMer`; refused on `joint_laplace` fits at this engine pin); Wald routes refused on `pirls_profiled` fits |

Use `inference_options(fit)` to see, before you run anything, which inference
routes are available on a given fit and why.
