---
title: Migrating from lme4
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from lme4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
has_lme4 <- requireNamespace("lme4", quietly = TRUE)
```

```{r load}
library(mixeff)
```

```{r sleepstudy-helper, include = FALSE}
# vignette-local convenience so chunks run whether or not lme4 is attached;
# use lme4::sleepstudy directly in your own code.
sleepstudy_data <- function() {
  if (requireNamespace("lme4", quietly = TRUE)) {
    e <- new.env()
    utils::data("sleepstudy", package = "lme4", envir = e)
    e$sleepstudy
  } else {
    set.seed(1)
    data.frame(
      Reaction = rnorm(180, 300, 50),
      Days = rep(0:9, 18),
      Subject = factor(rep(seq_len(18), each = 10))
    )
  }
}
```

`mixeff` fits lme4-style formulas with the familiar extractors and, on
its supported envelope, gives statistical answers tested against pinned
lme4 references within documented tolerances. It is not a literal drop-in —
you call `lmm()` / `glmm()` rather than `lmer()` / `glmer()`, and requests
outside the envelope are changed or declined with a diagnostic, never
silently. This vignette is the verb-for-verb,
argument-for-argument map; the generated tables below are the envelope's
exact boundary. It is the migration entry in the package's six-vignette
set: the support contract itself is stated in
`vignette("mixeff", package = "mixeff")`, the fitting workflows in
`vignette("lmm", package = "mixeff")` and
`vignette("glmm", package = "mixeff")`, convergence and boundary semantics
in `vignette("convergence", package = "mixeff")`, and the inference
contract in `vignette("inference", package = "mixeff")`.

## The two edits

An `lmer` script becomes an `lmm` script with two changes: the fitting verb and
the control object.

```{r fit, eval = TRUE}
fit <- lmm(Reaction ~ Days + (Days | Subject), sleepstudy_data(),
           control = mm_control(verbose = -1))
fixef(fit)
```

```{r fit-lme4, eval = has_lme4}
m <- lme4::lmer(Reaction ~ Days + (Days | Subject), lme4::sleepstudy)
lme4::fixef(m)
```

(`sleepstudy_data()` above just returns `lme4::sleepstudy` when lme4 is
installed; use `lme4::sleepstudy` directly in your own code.)

## Verb map

| lme4 | mixeff | Notes |
|------|--------|-------|
| `lmer(y ~ x + (x \| g), data)` | `lmm(y ~ x + (x \| g), data)` | same formula language, incl. `(x\|\|g)`, `(1\|g1/g2)`, crossed |
| `glmer(y ~ ..., family = binomial)` | `glmm(y ~ ..., family = binomial())` | pass a family **object** (`binomial()`), not a string |
| `lmerControl(...)` / `glmerControl(...)` | `mm_control(verbose=, max_feval=)` | optimizer/tolerance knobs are engine-chosen (see below) |
| `fixef`, `ranef`, `VarCorr`, `coef`, `sigma`, `vcov` | identical | same generics |
| `logLik`, `AIC`, `BIC`, `deviance`, `nobs`, `confint` | identical | `confint` supports Wald, profile, and bootstrap on LMMs; Wald/profile/bootstrap availability per GLMM fit type is the route matrix in `vignette("inference", package = "mixeff")` |
| `predict`, `fitted`, `residuals`, `simulate`, `refit` | identical (LMM) | `predict()` supports `re.form = NULL/NA`, `se.fit`, `interval`; `simulate()`/`refit()` are LMM-only and refuse GLMM fits with a typed error |
| `update(fit, . ~ . - x)` | identical | formula edits, `REML=`, `weights=`, etc. |
| `anova(m1, m2)`, `drop1`, `getME`, `ngrps`, `isSingular` | identical | `isSingular()` is `is_singular()` |
| `broom.mixed::tidy/glance/augment` | identical | registered for `mm_lmm`/`mm_glmm` |
| `emmeans::emmeans(fit, ~ x)` | identical | mixeff registers an emmeans basis on the same covariance payload, so estimates and SEs agree; mixeff's status/reliability/reason audit columns do not survive the emmeans trip, and a default profiled GLMM refuses inference through emmeans too |
| `lmerTest` p-values in `summary()` | built in | Satterthwaite/Kenward-Roger native, no extra package |

## Argument map for `lmm()` / `glmm()`

| lme4 argument | mixeff | Notes |
|---------------|--------|-------|
| `REML` | `lmm(..., REML=)` | same |
| `weights` | `weights=` | LMM and GLMM |
| `offset` | `glmm(..., offset=)` | GLMM only; LMM in-fit offset is not yet supported |
| `subset` | `lmm(..., subset=)` | supported for `lmm()` |
| `na.action` | `lmm(..., na.action=)` | **default refuses NA**; pass `na.action = na.omit` for lme4's complete-case behavior |
| `contrasts` | partial | unordered factors use treatment coding, ordered factors `contr.poly` (both matching R/lme4 defaults); other codings are refused — recode the factor. Contrast attributes set on factor columns and `options(contrasts=)` are currently ignored — the fit uses treatment coding — a known gap (bd-01KZ050ADJ2Y7SS20YH9CTE2NS); recode factors explicitly when porting |
| numeric/character grouping (accepted silently) | coerced to factor | non-factor grouping columns are coerced to factor, announced via an `mm_grouping_coercion_notice`; silence it with `mm_control(verbose = -1)` |
| `family = "binomial"` | `family = binomial()` | string families are not accepted |
| `nAGQ` | `glmm(..., nAGQ=)` | `>1` on the profiled path |
| `control = lmerControl(optimizer=, optCtrl=)` | `mm_control(optimizer=, max_feval=, ...)` | the engine picks a default optimizer; `mm_control()` can override it or cap the evaluation budget |
| `start` | `mm_control(start=)` | theta warm starts |

## Four differences to check when porting

**1. Coefficient names match lme4 exactly.** Since 0.2.0, `fixef()`,
`summary()` tables, `vcov()` dimnames, and `mm_lincomb()` weight names use
lme4's naming and column order (`"recipeB"`, `"temperature.L"`,
`"recipeB:temperature.L"`), so lme4 code keyed on coefficient names carries
over without renaming.
(Earlier versions used an engine encoding like `"recipe: B"`; if you wrote
normalization shims for those, delete them.)

**2. Grouped binomial responses.** `glmm()` accepts the
`cbind(successes, failures)` spelling like `glmer`:

```{r cbind, eval = FALSE}
glmm(cbind(incidence, size - incidence) ~ period + (1 | herd),
     lme4::cbpp, family = binomial())
```

**3. The default GLMM estimator is not glmer's.** `glmm()` defaults to a fast
profiled (PIRLS) estimator whose coefficients do **not** match `glmer()`
exactly; it prints a notice saying so. For glmer-equivalent (joint Laplace)
estimates, ask for them:

```{r joint, eval = FALSE}
glmm(y ~ x + (1 | g), data, family = binomial(),
     method = "joint_laplace")
```

**4. `||` with a factor means full decorrelation.** In mixeff, `||` fixes
*every* covariance in the block at zero — including the covariances among a
factor's level contrasts (each treatment-coded contrast gets an independent
variance), matching `MixedModels.jl`'s `zerocorr()`. These are mixeff's own
semantics, stated directly: lme4 documents its `||` as intended for numeric
predictors, and as currently implemented it leaves factor terms intact (the
factor keeps its full within-factor covariance block) rather than promising
a factor behavior. So `(1 + cond + x || subj)` with a
factor `cond` fits a strictly larger model in lme4 than in mixeff. The two
therefore disagree on the parameter count, hence on `df` and AIC — and on
the optimum itself whenever the fitted within-factor covariance is
non-zero. mixeff announces
the situation at compile time with an info diagnostic
(`covariance_assumption`, reason `double_bar_factor_term`). To reproduce
lme4's model family exactly, write the expansion explicitly and give the
factor its own correlated block:

```{r doublebar-factor, eval = FALSE}
# mixeff `||`: independent variances for every column, factor levels included
glmm(y ~ cond * x + (1 + cond + x || subj), data, family = binomial())

# lme4-equivalent family: the factor keeps its within-factor covariance block
glmm(y ~ cond * x + (1 | subj) + (0 + cond | subj) + (0 + x | subj),
     data, family = binomial())
```

## The supported envelope, generated

The tables below are rendered from the package's machine-readable support
registry (`inst/support/model-support.json`, via `supported_models()`), so
this vignette cannot drift from what the code enforces — the test suite
holds the fitting code, this registry, and the `glmm()` documentation to
the same envelope.

```{r support-matrix, echo = FALSE}
sm <- supported_models()
knitr::kable(sm$glmm_families,
             col.names = c("family", "links", "estimators", "notes"),
             caption = "Supported GLMM family/link/estimator cells.")
knitr::kable(sm$features[, c("feature", "lmm", "glmm", "notes")],
             caption = "Feature support (see notes for mixeff-specific semantics).")
```

```{r support-matrix-checks, include = FALSE}
stopifnot(nrow(sm$glmm_families) >= 4L, nrow(sm$features) >= 5L)
stopifnot(identical(
  sort(sm$glmm_families$family),
  sort(names(mixeff:::mm_glmm_supported_family_links()))
))
```

Explicit non-goals: `r paste(sm$non_goals, collapse = "; ")`.

## What is `NA`-with-a-reason (and why)

Where an inference method is unavailable, `mixeff` returns `NA` with a
machine-readable reason or raises a typed condition, rather than printing a
number without comment:

| Situation | lme4 (+lmerTest) | mixeff |
|-----------|------|--------|
| `NA` in a model variable | silently dropped | refused unless `na.action = na.omit` |
| Boundary (singular) fit | one-time warning | persistent `[boundary]` tag + effective rank |
| Satterthwaite df at a boundary | lmerTest prints df anyway | refused with a reason; use bootstrap |
| Prediction SE for an unseen grouping level | — | `NA` with the reason in `mm_reason` |
| GLMM `confint()` | computed | Wald intervals on `joint_laplace` fits via `method = "asymptotic"` (lowercase; lme4's `method = "Wald"` spelling is not accepted); parametric-bootstrap intervals via `method = "bootstrap"` on profiled-estimator fits (like `bootMer`; refused on `joint_laplace` fits at this engine pin); Wald routes refused on the default `pirls_profiled` fit |

Use `inference_options(fit)` to see, before you run anything, which inference
routes are available on a given fit and why.
