---
title: Linear Mixed Models
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear Mixed Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-package}
library(mixeff)
```

You use a linear mixed model when observations are not independent:
visits within clinics, trials within people, students within
classrooms. The standard R answer is `lme4::lmer()`. `mixeff::lmm()`
accepts the same formula language — `(1 | group)`, `(x | group)`,
`(1 + x || group)`, `(1 | a/b)`, crossed blocks — and the fitted
object answers to the common generics: `fixef()`, `ranef()`,
`VarCorr()`, `predict()`, `summary()`, `confint()`, `anova()`, and
friends. The supported surface is documented and closed rather than
open-ended: `vignette("mixeff", package = "mixeff")` states the
contract, and `vignette("lme4-migration", package = "mixeff")` gives
the exact verb-and-argument map.

What `lmm()` adds is that the fitted object carries the design, the
convergence status, and the inference labels with it. You can audit,
summarize, save, and reload without recomputing.

This vignette is the LMM walkthrough: reading a random-effects
formula before you fit, fitting, reading the summary, the standard
extractors, prediction, and weights — followed by a longer section on
what the random-effects spellings themselves mean and the
compile-time design help.

## The data

```{r simulate-data, include = FALSE}
set.seed(1)
n_subjects <- 18L
sleep <- expand.grid(
  Days = 0:9,
  Subject = factor(seq_len(n_subjects))
)
subject_intercept <- rnorm(n_subjects, sd = 25)
subject_slope <- rnorm(n_subjects, sd = 6)
subject_index <- as.integer(sleep$Subject)
sleep$Reaction <- 250 + 10.5 * sleep$Days +
  subject_intercept[subject_index] +
  subject_slope[subject_index] * sleep$Days +
  rnorm(nrow(sleep), sd = 25)
```

The example simulates reaction times (ms) for 18 subjects over ten
days of sleep deprivation. Each subject has a different baseline and
a different response to continued sleep loss. The data are generated
inside this vignette, so the walkthrough runs without an example-data
package.

```{r show-data}
str(sleep)
```

The key structure: `Subject` appears 10 times in the data — once per
day. Those 10 observations are *not* independent. They share the
subject's baseline speed, their individual sensitivity to sleep loss,
and any other between-person trait that we haven't measured. If you
fit a plain OLS regression, those correlations inflate your Type I
error rate and make standard errors too small.

A mixed model handles this by giving each subject its own intercept
*and* its own slope — their personal baseline and their personal rate
of slowing — and then estimating a population distribution over those
person-level parameters.

```{r individual-trajectories, echo = FALSE, fig.width = 6, fig.height = 4}
subject_lines <- unstack(sleep, Reaction ~ Subject)
matplot(0:9, subject_lines, type = "l", lty = 1, col = "grey70",
        xlab = "Days of sleep deprivation", ylab = "Reaction time (ms)",
        main = "Individual subject trajectories")
abline(lm(Reaction ~ Days, data = sleep), col = "steelblue", lwd = 2)
```

The grey lines are individuals; the blue line is the population
average. Subjects vary in their starting point *and* in how fast they
slow down — which is exactly what a random-intercept-and-slope model
captures.

## Step 1: Read the formula before you fit

Before fitting, compile the model to see what `mixeff` understands
about your formula. This separates formula interpretation from
optimization; you can catch mis-specified random effects without
paying the cost of a full fit.

```{r compile}
spec <- compile_model(Reaction ~ Days + (Days | Subject), sleep)
explain_model(spec)
```

This pre-fit explanation confirms:

- one fixed effect (`Days`) plus an intercept,
- one random-effects block with a correlated intercept and slope per
  subject,
- the formula has been canonicalized to the explicit
  `(1 + Days | Subject)` form that the optimizer will receive.

Use `audit(spec)` when you want the deeper design audit. Its compact
print starts with the audit summary and requested model;
`print(audit(spec), full = TRUE)` shows the complete upstream report.

If you had written `(1 | Subject)` by mistake (random intercepts
only), the explanation would show only one random-effects column per
subject — useful to verify before a long fit. The second half of this
vignette works through the random-effects spellings, canonical forms,
and design diagnostics in detail.

## Step 2: Fit

```{r fit}
fit <- lmm(Reaction ~ Days + (Days | Subject), sleep)
```

`lmm()` returns an `mm_lmm` object. The bundled Rust engine performs
the optimization; `lmm()` is the R entry point and result container.

```{r fit-checks, include = FALSE}
stopifnot(inherits(fit, "mm_lmm"))
stopifnot(all(is.finite(fixef(fit))))
stopifnot(length(predict(fit)) == nrow(sleep))
```

Printing the fit gives the formula, convergence status, likelihood
summary, residual scale, and fixed effects.

```{r print-fit}
fit
```

## Step 3: Read the summary

```{r summary}
summary(fit, tests = "coefficients")
```

The summary has four blocks, in print order.

**Fit status.** Check this line first: `converged_interior` is the
status every other block of the summary assumes. When a fit lands
somewhere else — a variance component at zero, a reduced-rank
covariance — the status says so and the available inference methods
change with it. What the statuses mean, and what a boundary does to
the rest of the output, is the subject of
`vignette("convergence", package = "mixeff")`.

**Variance components.** `Subject (Intercept)` is the estimated
between-subject spread in baseline reaction time; `Subject Days` is
the estimated between-subject spread in sensitivity to sleep loss;
`correlation` is their estimated association. The residual standard
deviation is the within-subject noise left over.

**Fixed effects.** `Estimate` is the population-level coefficient.
The `Days` estimate is the fitted average change in reaction time per
day; the simulation used a population slope of 10.5 ms per day. The
`method` column names how each p-value was computed — here
`satterthwaite`, a finite-sample *t* test whose `df` column is on the
scale of the 18 subjects, not the 180 raw rows. Which methods exist,
when each applies, and how to choose among them is covered in
`vignette("inference", package = "mixeff")`.

**Inference status.** One row per coefficient stating how much to
trust the test: `status` says whether it was computed, `reliability`
grades it, and `reliability_reason` names the engine's reason for
that grade — here `satterthwaite_finite_difference_approximation`,
the standard route for this method. The Rust engine assigns these
grades and reasons; the R side passes them through unchanged. The
closed vocabularies behind these columns are defined in the
[inference-method
glossary](https://bbuchsbaum.github.io/mixeff/articles/inference-method-glossary.html)
on the package site.

```{r summary-method-checks, include = FALSE}
s_auto <- summary(fit, method = "auto")
stopifnot(all(s_auto$inference$table$method == "satterthwaite"))
stopifnot("satterthwaite_finite_difference_approximation" %in%
            s_auto$inference$table$reliability_reason)
ctab <- s_auto$coefficients
p_col <- grep("^Pr\\(|^p\\.value$", names(ctab), value = TRUE)
stopifnot(length(p_col) == 1L)
stopifnot(any(is.finite(ctab[[p_col]])))
stopifnot(all(is.finite(ctab$df)), all(ctab$df < 40))
```

## Step 4: Extract components

The lme4-compatible extractors work on `mm_lmm` objects.

```{r fixed-effects}
fixef(fit)
sigma(fit)
logLik(fit)
```

`VarCorr()` reports fitted variance components, and `ranef()` returns
conditional random effects by grouping factor.

```{r random-effects}
VarCorr(fit)
head(ranef(fit)$Subject)
```

```{r extractor-checks, include = FALSE}
stopifnot(nrow(ranef(fit)$Subject) == length(unique(sleep$Subject)))
stopifnot(is.finite(as.numeric(logLik(fit))))
stopifnot(is.finite(deviance(fit)))
```

```{r confint}
confint(fit, method = "asymptotic")
```

`confint()` with `method = "asymptotic"` gives Wald intervals. The
`(Intercept)` interval is the population mean reaction time on Day 0;
the `Days` interval is the range of plausible slopes. The other
interval methods, and when each is available, are covered in
`vignette("inference", package = "mixeff")`.

Design matrices can be rebuilt from the stored formula and model
frame — including after a save/reload cycle
(`vignette("mixeff", package = "mixeff")` covers persistence).
`getME()` provides a small familiar subset for code that expects
lme4-style names.

```{r design-matrices}
X <- model.matrix(fit, type = "fixed")
Z <- model.matrix(fit, type = "random")

dim(X)
dim(Z)
class(Z)

getME(fit, c("theta", "beta", "cnms"))
```

```{r design-checks, include = FALSE}
stopifnot(nrow(X) == nrow(sleep))
stopifnot(nrow(Z) == nrow(sleep))
stopifnot(inherits(Z, "dgCMatrix"))
```

For report-ready data-frame tables of the same fit — design summary,
random terms, fixed effects — see the [reporting
article](https://bbuchsbaum.github.io/mixeff/articles/reporting-lmms.html)
on the package site.

## Step 5: Prediction and residuals

For fitted data, `predict()` returns in-sample fitted values. Use
`re.form = NA` for the fixed-effect part only.

```{r predictions}
prediction_check <- data.frame(
  Reaction = sleep$Reaction,
  fitted = predict(fit),
  fixed_only = predict(fit, re.form = NA),
  residual = residuals(fit)
)
head(prediction_check)
```

```{r prediction-checks, include = FALSE}
stopifnot(all(is.finite(prediction_check$fitted)))
stopifnot(all(is.finite(prediction_check$residual)))
stopifnot(max(abs(prediction_check$Reaction -
                    prediction_check$fitted -
                    prediction_check$residual)) < 1e-8)
```

### Prediction standard errors and intervals

Population-level prediction standard errors and intervals are
available via `re.form = NA` (the Wald SE of the fixed-effect linear
predictor):

```{r predict-se-population}
pop <- predict(fit, re.form = NA, se.fit = TRUE)
head(pop$se.fit)
head(predict(fit, re.form = NA, interval = "confidence"))
```

*Conditional* prediction standard errors (the default,
`re.form = NULL`) come from the engine's prediction-variance payload,
which combines the fixed-effect Wald variance with the random-effect
(BLUP) variance and the fixed/random covariance. The covariance term
can be negative, so the conditional SE can come out *smaller* than
the population SE. Conditional confidence and prediction intervals
come from the same payload.

```{r predict-se-conditional}
pred <- predict(fit, se.fit = TRUE)
head(data.frame(population_se = pop$se.fit, conditional_se = pred$se.fit))
head(predict(fit, interval = "confidence"))
```

```{r predict-se-checks, include = FALSE}
stopifnot(all(is.finite(pop$se.fit)))
stopifnot(all(is.finite(pred$se.fit)), all(pred$se.fit > 0))
```

Rows the engine cannot certify — for example an unseen grouping level
predicted with `allow.new.levels = TRUE`, where there is no posterior
variance for the missing level — return `NA` with the engine's reason
in the `mm_reason` attribute rather than a fabricated number. That
refusal vocabulary is part of the inference contract
(`vignette("inference", package = "mixeff")`).

## Step 6: Case weights

`lmm()` accepts case weights: one finite, positive value per row,
evaluated in `data` like the formula variables. A common use is rows
that summarize different numbers of underlying trials.

```{r fit-weights}
sleep$n_trials <- rep(c(1, 2), length.out = nrow(sleep))

fit_w <- lmm(Reaction ~ Days + (Days | Subject), sleep,
             weights = n_trials,
             control = mm_control(verbose = -1))
fixef(fit_w)
head(fit_w$weights)
```

The weights are stored on the fit and reused by `refit()` and the
inference routines. Invalid weights — wrong length, negative, or
non-finite — are refused with a typed `mm_data_error` rather than
silently recycled.

```{r weights-checks, include = FALSE}
stopifnot(isTRUE(all.equal(fit_w$weights, sleep$n_trials)))
stopifnot(all(is.finite(fixef(fit_w))))
bad <- tryCatch(
  lmm(Reaction ~ Days + (Days | Subject), sleep, weights = rep(1, 5),
      control = mm_control(verbose = -1)),
  error = identity
)
stopifnot(inherits(bad, "mm_data_error"))
```

For contrasts, term tests, marginal means, and model comparisons on a
fit like this one, continue with
`vignette("inference", package = "mixeff")`. The rest of this
vignette turns to the random-effects formulas themselves.

## Random-effects formulas in depth

Random-effects formulas are compact, and compactness hides
assumptions. Seven characters — `(x | g)` — commit the analyst to
a specific covariance structure, a specific number of free
parameters, and a specific identification claim about what the data
can support. Two formulas that differ by a single character — `||`
for `|`, say — fit different models.

`mixeff` shares the formula language with `lme4`, so the notation
you already read transfers directly. What it adds is a way to read
the formula *before* the optimizer runs: the named form of every
random term, the plain-language scope, the parameter count, and the
design facts that support or refuse the formula on this particular
data.

The goal of this section is not to pick a model for you. It is to
show what the model you wrote can and cannot express, and to make
the difference between nearby spellings legible.

We use a second small repeated-measures dataset with one row per
subject per time point.

```{r study-data, include = FALSE}
set.seed(11)
n_subj <- 30L
n_time <- 5L
study <- expand.grid(subject = factor(seq_len(n_subj)),
                     time = seq_len(n_time))
subj_dose <- rnorm(n_subj, sd = 1)
study$dose <- subj_dose[as.integer(study$subject)]
b0 <- rnorm(n_subj, sd = 0.6)[as.integer(study$subject)]
b1 <- rnorm(n_subj, sd = 0.2)[as.integer(study$subject)]
study$x <- rnorm(nrow(study))
study$z <- rnorm(nrow(study))
study$score <- 2 + 0.3 * study$time + b0 + b1 * study$time +
  rnorm(nrow(study), sd = 0.4)
study$y <- 1 + 0.4 * study$x + 0.2 * study$z + b0 +
  rnorm(nrow(study), sd = 0.5)
study$hit <- rbinom(nrow(study), size = 1, prob = plogis(0.5 + study$x))

# A between-subject covariate is required to demonstrate the
# structural-refusal case.
between_study <- study
between_study$dose <- between_study$dose

# A small-N variant to demonstrate the low-information-budget case.
low_n <- 4L
low_study <- subset(study, as.integer(subject) <= low_n)
low_study$subject <- factor(low_study$subject)
```

```{r study-preview}
head(study)
```

### What does `(1 | subject)` say?

A random intercept is the simplest random-effects structure:
subjects may differ in average score, but the fixed `time` effect
is the same across subjects. It is often the right choice, and it
is always a reasonable starting point. `explain_model()` spells
out what the model assumes and reports that no random slopes were
added.

```{r intercept-only-explanation}
intercept_only <- compile_model(score ~ time + (1 | subject), study)
explain_model(intercept_only)
```

```{r intercept-only-checks, include = FALSE}
intercept_only_explanation <- explain_model(intercept_only)
intercept_only_slopes <-
  intercept_only_explanation$cards[[1L]]$blocks[[1L]]$slopes
stopifnot(length(intercept_only_slopes) == 0L)
stopifnot(any(vapply(
  intercept_only_explanation$diagnostics,
  function(x) identical(x$code, "scope_note"),
  logical(1)
)))
```

Each random-effects card has two header lines worth pausing on.
`wrote:` is the term as you typed it. `canonical:` is what the
parser hands downstream — what the optimizer, the audit, and the
inference contract will actually see. They are identical here
because `(1 | subject)` is already in canonical form — nothing was
rewritten. The two diverge for
more elaborate formulas. `(1 | clinic/site)` is rewritten as
`(1 | clinic) + (1 | clinic:site)`; implicit intercepts are made
explicit; double-bar shorthand expands to its split-block form.
The `canonical:` line is where you read off the model the engine
will actually fit.

The remaining lines decode the term further. `named form:`
restates it as a function call that names the grouping factor, the
intercept, the slopes (if any), and the covariance family.
`scope:` is a plain-language sentence. `covariance` and `support`
report parameter cost and design sufficiency.

The `Design notes:` section at the bottom records facts about the
data that the fitted formula did not use. Here `time` varies
within each subject, so a subject-level `time` slope is
structurally possible. That is reported as a note, not a
correction: whether to add `(time | subject)` is a scientific
question about whether subjects plausibly differ in their `time`
slopes, not a mechanical one about what the package can do.

### What changes when you add a random slope?

`(1 + time | subject)` requests a two-dimensional random effect for each
subject: one baseline coefficient and one `time` coefficient, with a fitted
covariance between them.

```{r full-slope-explanation}
full <- compile_model(score ~ time + (1 + time | subject), study)
explain_model(full)
```

After fitting, `parameterization()` shows the optimizer's *internal*
coordinates: the vector θ the engine actually searched over, and
where each entry sits in the random-effects covariance factor.

```{r full-parameterization}
full_fit <- lmm(
  score ~ time + (1 + time | subject),
  study,
  control = mm_control(verbose = -1)
)
parameterization(full_fit)$table[, c("theta_name", "theta_value", "lambda_value")]
```

What θ is, why the engine optimizes it rather than the covariance
matrix directly, and how boundary fits are read off this table are
covered in `vignette("convergence", package = "mixeff")`.

### What do split blocks and `||` mean?

The formula `(1 | subject) + (0 + time | subject)` uses two separate blocks for
the same grouping factor. That fixes the intercept-slope covariance to zero.

```{r split-blocks}
split_blocks <- compile_model(
  score ~ time + (1 | subject) + (0 + time | subject),
  study
)
explain_model(split_blocks)
```

`(1 + time || subject)` is a shorthand for the same split-block model.

```{r double-bar}
double_bar <- compile_model(score ~ time + (1 + time || subject), study)
explain_model(double_bar)
```

```{r split-checks, include = FALSE}
split_text <- explain_model(split_blocks)$text
double_text <- explain_model(double_bar)$text
stopifnot(grepl("separate random-effect blocks fix the covariance", split_text, fixed = TRUE))
stopifnot(grepl("double-bar syntax fixes the covariance", double_text, fixed = TRUE))
stopifnot(grepl("covariance_assumption", split_text, fixed = TRUE))
stopifnot(grepl("covariance_assumption", double_text, fixed = TRUE))
```

One caveat: when a **factor** appears inside `||`, mixeff's expansion
differs from lme4's — `vignette("lme4-migration", package =
"mixeff")` states the exact semantics and the explicit spelling that
reproduces lme4's model family.

### Nested and crossed grouping

Nesting is written with `/` and expands mechanically:
`(1 | clinic/ward)` means one intercept block per clinic plus one per
ward-within-clinic. Crossed factors are simply two blocks —
`(1 | subject) + (1 | item)` — and nothing is rewritten, because that
spelling is already canonical. `mm_parse_formula()` shows the
canonical string for any formula without touching data:

```{r canonical-spellings}
mm_parse_formula("rt ~ days + (1 | subject/task)")
mm_parse_formula("rt ~ days + (1 | subject) + (1 | item)")
mm_parse_formula("rt ~ days + (1 + days || subject)")
```

```{r canonical-checks, include = FALSE}
stopifnot(identical(
  mm_parse_formula("rt ~ days + (1 | subject/task)"),
  "rt ~ 1 + days + (1 | subject) + (1 | subject:task)"
))
stopifnot(identical(
  mm_parse_formula("rt ~ days + (1 | subject) + (1 | item)"),
  "rt ~ 1 + days + (1 | subject) + (1 | item)"
))
stopifnot(identical(
  mm_parse_formula("rt ~ days + (1 + days || subject)"),
  "rt ~ 1 + days + (1 + days || subject)"
))
```

The `explain_model()` cards agree with these canonicalizations — they
show the model the engine actually fits — with one display difference:
`mm_parse_formula()` keeps a `||` term as written, while the cards
additionally show it as the split blocks the engine fits.

### What are the three kinds of help?

The compile-time diagnostics separate three situations that are often mixed
together in ordinary model output.

First, a structural impossibility: a requested random slope cannot be estimated
if that variable does not vary within the group. Here `dose` is constant within
each subject.

```{r structural-impossibility}
impossible <- compile_model(score ~ dose + (1 + dose | subject), between_study)
explain_model(impossible)
```

Second, low information budget: a requested full covariance can be estimable
in principle while still having little grouping-level support. The package
reports the parameter count and observed levels.

```{r low-information}
low_info <- compile_model(score ~ time + (1 + time | subject), low_study)
diagnostics(low_info)$table[, c("code", "severity", "message")]
```

Third, unmodeled-but-possible: a fixed effect varies within group, but the
random-effects formula does not include the corresponding slope. This is the
quiet design note you saw for `(1 | subject)`.

```{r unmodeled-possible}
diagnostics(intercept_only)$table[, c("code", "severity", "message")]
```

```{r help-checks, include = FALSE}
stopifnot("structural_refusal" %in% diagnostics(impossible)$table$code)
stopifnot("covariance_too_rich" %in% diagnostics(low_info)$table$code)
stopifnot("scope_note" %in% diagnostics(intercept_only)$table$code)
```

### How do you inspect nearby formulas?

`random_options()` is a map of nearby random-effect spellings for one grouping
factor. It marks the current spelling and reports parameter costs and support
facts. It does not rank the rows.

```{r random-options-map}
random_options(intercept_only, group = subject, slope = time)
```

The same information is available as a data frame if you want to build a
custom report.

```{r random-options-table}
opts <- random_options(intercept_only, group = subject, slope = time)
opts$options[, c("formula", "theta_parameters", "design_status", "current")]
```

### Why is there no recommendation row?

`mixeff` treats formula explanations as an audit problem, not a model-selection
problem. The package reports the current model, nearby spellings, assumptions,
and data-support facts. Choosing among scientifically different random-effects
structures remains part of the analysis design.

`compare_covariance()` gives another view of the same principle: full,
diagonal, and scalar covariance families are displayed side by side, without a
preferred row.

```{r compare-covariance}
compare_covariance(full)
```

```{r no-recommendation-check, include = FALSE}
opts <- random_options(intercept_only, group = subject, slope = time)
cmp <- compare_covariance(full)
stopifnot(!"recommended" %in% tolower(names(opts$options)))
stopifnot(!"recommended" %in% tolower(names(cmp$table)))
```

### Where do fitted-model changes appear?

After fitting, `changes()` records the journey from the formula you
wrote to the model the engine actually solved. Each row carries the
stage at which something happened, the term it affected, the status
of the change, and three side-by-side renderings: what was
*requested*, what was *effective* (after design parsing and
identifiability checks), and what was *fitted* (after the optimizer
landed). A reduced-rank covariance estimate appears here as a
labeled fact, not an instruction to rewrite the formula (boundary
and reduced-rank outcomes are the subject of
`vignette("convergence", package = "mixeff")`).

The full row carries a lot of text. We show it in two passes so
neither pass wraps awkwardly in the rendered vignette. The first
pass names the stage of each change and its status:

```{r fitted-changes-status}
knitr::kable(
  changes(full_fit)$table[, c("stage", "term_id", "status", "detail")]
)
```

The second pass shows the requested → effective → fitted arc
itself, where the three columns are the substance of the report:

```{r fitted-changes-arc}
knitr::kable(
  changes(full_fit)$table[, c("stage", "requested", "effective", "fitted")]
)
```

## Where to go next

| Question | Where |
|---|---|
| What is the supported envelope, and how do fits persist across sessions? | `vignette("mixeff", package = "mixeff")` |
| How are p-values computed, and which method is right for my fit? | `vignette("inference", package = "mixeff")` |
| How do I compute marginal means and treatment contrasts? | `vignette("inference", package = "mixeff")` |
| What do convergence statuses, boundaries, and singular fits mean? | `vignette("convergence", package = "mixeff")` |
| Can I fit a binomial/Poisson/negative-binomial GLMM? | `vignette("glmm", package = "mixeff")` |
| I'm coming from lme4 — what's different? | `vignette("lme4-migration", package = "mixeff")` |
| How do I write up the results for a paper? | the [reporting article](https://bbuchsbaum.github.io/mixeff/articles/reporting-lmms.html) on the package site |
