---
title: Marginal Means and Comparisons
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Marginal Means and Comparisons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-package}
library(mixeff)
```

A fixed-effects coefficient is not a group mean. For a model with multiple
predictors and interactions, `fixef(fit)["trtactive"]` is the treatment
effect *at the reference level of every other predictor* — not the
average treatment effect across the population. Marginal means give you the
latter: averages over a reference grid of predictor combinations, with
standard errors from the fixed-effect covariance.

`mixeff` computes these with four functions — `mm_grid()`,
`mm_predictions()`, `mm_means()`, `mm_comparisons()` — that route all
inference through the same machinery as `contrast()`. Each row in the
returned table carries `method`, `status`, `reliability`, and `reason`
columns naming how it was computed and how far to trust it.

## The study

A rehabilitation trial assigns patients to coached or usual-care treatment.
Each patient is assessed before and after the intervention. Patients are
grouped within clinics.

```{r rehab-data, include = FALSE}
set.seed(101)
n_clinic <- 4L
n_subj   <- 24L
spc      <- n_subj %/% n_clinic   # subjects per clinic

clinic <- factor(rep(paste0("C", seq_len(n_clinic)), each = spc))
# Deliberately unbalanced assignment: C1 is mostly control, C3 mostly
# active. The imbalance is the point of this vignette.
n_control <- c(5L, 3L, 1L, 3L)
trt <- factor(
  unlist(lapply(seq_len(n_clinic), function(i) {
    c(rep("control", n_control[[i]]), rep("active", spc - n_control[[i]]))
  })),
  levels = c("control", "active")
)
subj <- factor(paste0("S", seq_len(n_subj)))

rehab <- data.frame(
  subj   = rep(subj,   each = 2L),
  clinic = rep(clinic, each = 2L),
  trt    = rep(trt,    each = 2L),
  time   = factor(rep(c("pre", "post"), times = n_subj), levels = c("pre", "post"))
)

clinic_re <- c(-1.2, 0.4, 0.9, -0.1)
subj_re   <- rnorm(n_subj, sd = 0.8)

rehab$score <- 52 +
  -2.1 * (rehab$trt  == "active") +
  -3.8 * (rehab$time == "post") +
  -2.9 * (rehab$trt  == "active") * (rehab$time == "post") +
  clinic_re[as.integer(rehab$clinic)] +
  subj_re[as.integer(rehab$subj)] +
  rnorm(nrow(rehab), sd = 0.4)
```

```{r show-data}
head(rehab)
```

Clinics are not balanced between treatment arms: clinic C1 has five
control patients and one active, clinic C3 the reverse. So raw group
means are confounded with clinic effects — each arm's mean leans on the
clinics that arm over-samples, and the clinics differ — and a naive
`tapply(rehab$score, rehab$trt, mean)` will not answer the treatment
question.

## Fit the model

```{r fit}
fit <- lmm(score ~ trt * time + (1 | clinic) + (1 | subj), rehab)
summary(fit, tests = "coefficients")
```

The interaction coefficient `trtactive:timepost` tells you the
*additional* post-treatment change for the active arm relative to control.
It is not the average treatment effect. For that you need marginal means.

## Reference grids with `mm_grid()`

`mm_grid()` constructs the cross-product of all fixed-predictor levels. By
default, factor predictors expand to all their levels; numeric predictors
collapse to their mean.

```{r grid}
g <- mm_grid(fit, specs = ~ trt * time)
g
```

The grid has four rows — one for each treatment × timepoint cell — and
retains the model matrix needed for inference.

## Cell predictions with `mm_predictions()`

`mm_predictions()` evaluates the fixed-effect prediction at each grid row,
with a confidence interval from the certified covariance.

```{r predictions}
preds <- mm_predictions(fit, specs = ~ trt * time)
preds$table[, c("label", "estimate", "conf_low", "conf_high", "method")]
```

These are the four population-level cell means. Each row carries its
inference method so the provenance is visible without digging into model
objects.

## Marginal means with `mm_means()`

Marginal means average the reference grid over dimensions you want to
collapse. Here: average over timepoints to get the *overall* treatment effect.

```{r means-trt}
mt <- mm_means(fit, specs = ~ trt)
mt$table[, c("label", "estimate", "conf_low", "conf_high", "method")]
```

Compare these to the raw means:

```{r raw-means}
tapply(rehab$score, rehab$trt, mean)
```

The clinic imbalance pulls each raw mean about 0.44 points toward the
clinics its arm over-samples; the marginal means average over the model's
fixed-effect grid instead and are free of that shift. The size of the gap
depends on how unbalanced the design is and how large the group effects
are.

## Pairwise comparisons with `mm_comparisons()`

`mm_comparisons()` takes all pairwise differences among the marginal means
and applies the same inference method.

```{r comparisons-trt}
ct <- mm_comparisons(fit, specs = ~ trt)
ct$table[, c("label", "estimate", "conf_low", "conf_high", "p_value", "method")]
```

The `active - control` row is the average treatment effect across both
timepoints, tested with a Satterthwaite *t* and graded `moderate`
reliability in the same row.

## Conditional comparisons with `by =`

The `by` argument splits comparisons within levels of another variable —
the analogue of simple effects in a factorial design.

```{r comparisons-by-time}
ct_by <- mm_comparisons(fit, specs = ~ trt | time)
ct_by$table[, c("label", "estimate", "conf_low", "conf_high", "p_value", "method")]
```

Two rows: the treatment difference *at pre-intervention* and the treatment
difference *at post-intervention*. The post-intervention gap is larger because
the interaction drives additional improvement in the active arm.

## Constraining the grid with `at =`

`at` pins predictors to chosen values instead of expanding or averaging
them: a factor to a subset of its levels, a numeric predictor to specific
points rather than its mean.

```{r at-example}
mt_time <- mm_means(fit, specs = ~ time, at = list(trt = "active"))
mt_time$table[, c("label", "estimate", "conf_low", "conf_high")]
```

This gives the pre/post means *within the active arm* only, holding `trt`
constant at `"active"`.

## Custom contrasts with `mm_lincomb()`

For hypotheses that are not pairwise differences of marginal means, build
the contrast weights directly with `mm_lincomb()`. The interaction effect
expressed as a contrast: (active post − active pre) − (control post − control pre).

```{r lincomb}
beta <- fixef(fit)
names(beta)

# Interaction row: coefficient named "trtactive:timepost" (lme4-identical)
w <- setNames(numeric(length(beta)), names(beta))
w["trtactive:timepost"] <- 1

lc <- mm_lincomb(fit, weights = w)
lc[, c("estimate", "lower", "upper", "p_value", "method")]
```

`mm_lincomb()` applies the same contract-preserving inference as `contrast()`,
so the method and status fields are populated identically.

## The `emmeans` bridge

When `emmeans` is installed, `mixeff` registers a basis so you can call
`emmeans::emmeans()` directly on `mm_lmm` objects.

```{r emmeans-bridge, eval = requireNamespace("emmeans", quietly = TRUE)}
if (requireNamespace("emmeans", quietly = TRUE)) {
  em <- emmeans::emmeans(fit, ~ trt)
  print(em)
  print(pairs(em))
}
```

`emmeans` uses the same `mixedmodels.fixed_effect_covariance_matrix` payload
as `mm_means()`, so the point estimates and standard errors agree. The bridge
prints an informational message noting which covariance it used.

**When to prefer the native verbs over `emmeans`:** `mm_means()` and
`mm_comparisons()` carry the full `status`, `reliability`, and `reason`
row-level audit fields from the underlying `contrast()` call. `emmeans`
does not propagate these fields; if a row is `"unavailable"` for a
documented reason, that information disappears in the `emmeans` output.
Use `emmeans` when you need what it does and `mm_means()` does not —
Tukey correction, back-transformation, custom contrast grammars. Use the
native verbs when you need each row to say how it was computed and why,
because those columns do not survive the trip through `emmeans`.

## Reading `status` and `reason`

Every table returned by `mm_means()`, `mm_comparisons()`, and
`mm_predictions()` carries these columns:

| Column | Values |
|---|---|
| `status` | `"available"`, `"p_value_unavailable"`, `"not_estimable"`, `"not_assessed"`, or `"unsupported"` |
| `reliability` | `"high"`, `"moderate"`, `"low"`, or `"not_available"` |
| `reason` | stable code when a row is degraded or refused, `NA` otherwise (see `vignette("inference-method-glossary")`) |
| `method` | the inference method that was applied |

On the fits in this vignette every row reads `status = "available"`,
`reliability = "moderate"`, `method = "satterthwaite"`. A row whose status
is anything other than `"available"` contains `NA` where the refused
numbers would go, and `reason` says why (rank deficiency, missing
covariance payload, and so on). These codes come from the engine's fixed
vocabulary, so scripts can test for them by string.

## Where to read next

- `vignette("inference")` — how the underlying `contrast()` machinery works
  and what inference methods are available.
- `vignette("inference-method-glossary")` — reference table of every
  `reason` code.
- `vignette("reporting-lmms")` — building a full results section from a
  fitted object.
