---
title: Introduction and the Supported Model Contract
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction and the Supported Model Contract}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-package}
library(mixeff)
```

If you fit mixed-effects models in R, you most likely use
`lme4::lmer()`. `mixeff::lmm()` speaks the same formula language, and
the extractors — `fixef()`, `ranef()`, `VarCorr()`, `predict()`,
`simulate()`, `anova()`, `summary()`, `update()` — do what you expect.
On the package's supported envelope, statistical answers agree with
`lme4` within documented tolerances on the parity datasets shipped
with the package; the envelope itself is documented, closed, and
machine-readable (`supported_models()`). It is not a literal
*drop-in*: you call `lmm()` rather than `lmer()`, results are not
bit-exact, and when the package changes or declines a request it says
so, in a diagnostic you can read from the fitted object.

The reason to switch is what `mixeff` does around the fit. On the
scaling benchmark shipped with the package it ran faster than `lme4`
at the largest tested scale of each of five common designs — all
small fits in absolute terms; the benchmarking article on the package
site,
<https://bbuchsbaum.github.io/mixeff/articles/benchmarking.html>,
reports the numbers and what they do and do not establish. And it
makes four things explicit that `lme4` leaves implicit:

1.  **The formula stays familiar.** Anything you would hand to
    `lmer()` you can hand to `lmm()`.
2.  **Singular fits become labeled facts.** A reduced-rank
    random-effect covariance is reported on the fitted object, not as
    a single console warning.
3.  **Inference labels its method.** Each p-value carries the method
    that produced it. Where the package declines to compute
    Satterthwaite or Kenward–Roger degrees of freedom — at a boundary
    fit, for example — a parametric bootstrap is offered under the
    same labeling rules.
4.  **The fitted object is a record.** A saved model reopens with the
    same coefficients, the same diagnostics, and the same method
    labels, without depending on the original Rust handle.

Behind the four is one design habit. In ordinary mixed-model work,
three questions come up again and again: what does this
random-effects formula actually mean; are the p-values or tests
available, and by what method; can the model state be reconstructed
after the fit has been saved. `mixeff` makes those questions part of
the fitted object.

This page demonstrates each of the four on one small dataset, then
states the supported-model contract and the persistence guarantees.
The other vignettes go deeper on every step.

## The dataset

A small repeated-measures study: 18 subjects, 10 daily reaction-time
measurements each, intercepts and slopes that are nearly perfectly
correlated by construction. This design produces a singular fit in
`lme4` and in `mixeff` alike.

```{r make-data}
set.seed(3)
n_subj <- 18L
days   <- 0:9
b0     <- rnorm(n_subj, sd = 30)
b1     <- 0.5 * b0 / 30 * 10 + rnorm(n_subj, sd = 0.5)

sleep_like <- do.call(rbind, lapply(seq_len(n_subj), function(i) {
  data.frame(
    subj = factor(i),
    days = days,
    rt   = 250 + b0[i] + (10 + b1[i]) * days +
           rnorm(length(days), sd = 20)
  )
}))

head(sleep_like, 4)
```

## A. The formula stays familiar

`mixeff` keeps the lme4 random-effects syntax. If you can read `(x | g)`, you
can read `mixeff`.

```{r fit-model}
fit <- lmm(
  rt ~ days + (1 + days | subj),
  sleep_like,
  control = mm_control(verbose = -1)
)
fit
```

```{r familiar-checks, include = FALSE}
stopifnot(inherits(fit, "mm_lmm"))
stopifnot(all(is.finite(fixef(fit))))
stopifnot(length(fitted(fit)) == nrow(sleep_like))
```

`fixef()`, `ranef()`, `VarCorr()`, `predict()`, `simulate()`, `anova()`, and
`summary()` all do what you expect.

## B. When a fit is degenerate, the object says so

By construction, this design pushes the random-effect covariance to
reduced rank. `mixeff` fits the model and records the outcome on the
object, where you can query it later:

```{r singular-status}
fit_status(fit)
is_singular(fit)
```

```{r singular-checks, include = FALSE}
stopifnot(identical(fit_status(fit), "converged_reduced_rank"))
stopifnot(isTRUE(is_singular(fit)))
```

What this status means, which variance component reached the
boundary, the diagnostic codes behind the label (`diagnostics()`),
the requested-to-effective-to-fitted record (`changes()`), and what a
boundary does to downstream inference are the subject of
`vignette("convergence", package = "mixeff")`.

## C. Inference labels its method

On the inference side, `mixeff` first reports which methods are
available on your fit, and then lets you run the one you choose under
the same labeling rules.

`inference_options()` enumerates the inference methods for the
current fit, gives each one a fixed-vocabulary *reason* for its
status (the vocabularies are defined in the [inference-method
glossary](https://bbuchsbaum.github.io/mixeff/articles/inference-method-glossary.html)
on the package site), and names the function you would call to invoke
it.

```{r inference-options}
opt <- inference_options(fit, "days", nsim = 200)
opt$table[, c("method", "expected_status",
              "expected_reliability_reason", "current")]
```

```{r options-checks, include = FALSE}
stopifnot(inherits(opt, "mm_inference_options"))
stopifnot(any(opt$table$expected_status == "available"))
stopifnot(any(opt$table$expected_status != "available"))
```

Some routes are immediately available on this fit; others decline
with a stable reason code rather than printing a number without
comment. Why a boundary fit takes some methods off the table is
explained in `vignette("convergence", package = "mixeff")`; the
routes themselves — including the parametric bootstrap that yields a
defensible p-value on a fit like this one — are walked through in
`vignette("inference", package = "mixeff")`.

Every reported quantity is in one of three states: available, with a
named method and reason; unavailable, with a stable reason code; or a
typed error.

## The supported model contract

The envelope is machine-readable. `supported_models()` returns the
same registry the fitting code enforces
(`inst/support/model-support.json`); the test suite holds the fitting
code, the registry, and the documentation to the same envelope.

```{r supported-envelope}
sm <- supported_models()
knitr::kable(sm$glmm_families[, c("family", "links", "estimators")],
             caption = "GLMM family/link/estimator cells, from the registry.")
knitr::kable(sm$features[, c("feature", "lmm", "glmm")],
             caption = "Feature support, from the registry.")
```

```{r envelope-checks, include = FALSE}
stopifnot(inherits(sm, "mm_model_support"))
stopifnot(nrow(sm$glmm_families) >= 4L, nrow(sm$features) >= 5L)
stopifnot(length(sm$non_goals) >= 1L, all(nzchar(sm$non_goals)))
```

Explicit non-goals, from the same registry:
`r paste(sm$non_goals, collapse = "; ")`.

These are the short forms. `vignette("lme4-migration", package =
"mixeff")` renders the full registry tables with their notes columns
and is the verb-for-verb, argument-for-argument map for porting an
lme4 script.

Two boundary conditions of the contract are worth stating up front:

- Estimates are not bit-for-bit identical to `lme4`; the bar is
  statistical agreement within documented tolerances on the parity
  datasets shipped with the package. Where results are *expected* to
  differ, the differences are classified in
  `inst/extdata/expected-mismatches.json`, with tolerances enforced
  by the test suite.
- Generalized fits carry their own estimator and inference contract;
  see `vignette("glmm", package = "mixeff")` for the estimators and
  `vignette("inference", package = "mixeff")` for the inference
  routes.

Each limit is reported by the package itself, as a reason code or a
typed error, when you hit it.

## D. The fit is the record

The fitted object is a serializable record. `saveRDS()` followed by
`readRDS()` and `revive()` reproduces the audit trail and the extractors
without depending on the original Rust handle.

```{r round-trip}
path <- tempfile(fileext = ".rds")
saveRDS(fit, path)
restored_fit <- revive(readRDS(path))

identical(fixef(restored_fit),             fixef(fit))
identical(changes(restored_fit)$table,     changes(fit)$table)
identical(diagnostics(restored_fit)$table, diagnostics(fit)$table)
```

```{r round-trip-checks, include = FALSE}
stopifnot(identical(fixef(restored_fit), fixef(fit)))
stopifnot(identical(changes(restored_fit)$table, changes(fit)$table))
stopifnot(identical(diagnostics(restored_fit)$table, diagnostics(fit)$table))
```

A reviewer reading the `.rds` later sees the same convergence status,
the same reduced-rank diagnostic, the same method labels on the same
coefficients.

### Saving, reloading, and reviving in practice

A mixed-model fit often outlives the R session that produced it: you
fit a model today, save it with the analysis, hand the script to a
collaborator, and reopen the fit six months later for a contrast, a
revision, or a referee response. `mixeff` stores the fitted values,
the random-effects design, the convergence record, and the inference
labels inside the R object, so each of those tasks works after
`readRDS()` without recomputing the fit.

The round trip above used the singular fit; the same guarantees hold
for an ordinary interior fit.

```{r persist-data, include = FALSE}
set.seed(42)
n_clinic <- 12L
n_per_clinic <- 6L
clinic <- factor(rep(seq_len(n_clinic), each = n_per_clinic))
week <- rep(0:(n_per_clinic - 1L), n_clinic)
treatment <- factor(
  rep(rep(c("usual", "coached"), each = n_per_clinic), length.out = length(week)),
  levels = c("usual", "coached")
)
clinic_shift <- rnorm(n_clinic, sd = 0.45)
score <- 7.2 - 0.28 * week - 0.55 * (treatment == "coached") +
  clinic_shift[as.integer(clinic)] + rnorm(length(week), sd = 0.35)
clinic_visits <- data.frame(score, week, treatment, clinic)
```

```{r persist-fit}
fit2 <- lmm(
  score ~ week + treatment + (1 | clinic),
  clinic_visits,
  control = mm_control(verbose = -1)
)

path2 <- tempfile(fileext = ".rds")
saveRDS(fit2, path2)

restored <- readRDS(path2)
restored <- revive(restored)
```

`revive()` resets the object's process-local cache. It is cheap and
safe to call after every `readRDS()`, but the extractors and
inference functions shown below also work on the plain `readRDS()`
result — the durable values live in the object itself, and the Rust
handle is deliberately absent after reload.

```{r persist-extractors}
fixef(restored)
head(predict(restored))
reporting_table(restored, "fixed_effects")
```

```{r persist-checks, include = FALSE}
stopifnot(identical(fixef(restored), fixef(fit2)))
stopifnot(length(predict(restored)) == nrow(clinic_visits))
stopifnot(identical(reporting_table(restored, "fixed_effects"),
                    reporting_table(fit2, "fixed_effects")))
```

Conditional variances for random effects also survive the round trip.
With `condVar = TRUE`, each grouping table carries a finite `postVar`
array.

```{r persist-condvar}
re <- ranef(restored, condVar = TRUE)
attr(re, "mm_unavailable_reason")
dim(attr(re$clinic, "postVar"))
```

```{r condvar-checks, include = FALSE}
stopifnot(is.null(attr(re, "mm_unavailable_reason")))
stopifnot(is.array(attr(re$clinic, "postVar")))
stopifnot(all(is.finite(attr(re$clinic, "postVar"))))
```

Two related surfaces are documented elsewhere: the honest-covariance
contract behind `vcov()` is part of the inference contract
(`vignette("inference", package = "mixeff")`), and rebuilding design
matrices (`model.matrix()`, `getME()`) from the stored formula and
model frame is shown in `vignette("lmm", package = "mixeff")`.

### The artifact contract

The structured JSON artifacts stored in the fitted object are the
source of truth; external pointers such as the Rust handle are
caches. `mm_json_known_schemas()` lists the artifact schemas this
version of the package understands:

```{r schemas}
head(mm_json_known_schemas())
```

```{r schema-checks, include = FALSE}
schemas <- mm_json_known_schemas()
stopifnot(is.data.frame(schemas))
stopifnot(all(c("name", "version") %in% names(schemas)))
stopifnot(nrow(schemas) >= 1L)
```

## Lower-level tools

Most users start with `lmm()`, `glmm()`, `summary()`, `contrast()`,
`test_effect()`, `compare()`, and `reporting_table()`. The
lower-level functions are there when you need them:

- `mm_parse_formula()` checks formula syntax and returns the
  canonical spelling (demonstrated in
  `vignette("lmm", package = "mixeff")`).
- `compile_model()` builds a pre-fit model specification;
  `explain_model()` translates it into named forms and plain-language
  scopes (also in `vignette("lmm", package = "mixeff")`).
- `diagnostics()` and `changes()` expose model-state checks.
- `mm_json_known_schemas()` lists the structured artifact schemas
  understood by this version of the package.

The Rust backend does not change how you write R: you fit the model,
get the numbers, and the status of those numbers stays attached to
the object.

## Where to read next

- `vignette("lmm", package = "mixeff")` — linear mixed models
  end-to-end: reading random-effects formulas, fitting, summaries,
  extractors, weights, and prediction.
- `vignette("glmm", package = "mixeff")` — the generalized families
  and estimators.
- `vignette("convergence", package = "mixeff")` — convergence
  statuses, boundary and singular fits, and verification.
- `vignette("inference", package = "mixeff")` — p-values, contrasts,
  term tests, bootstrap methods, confidence intervals, and marginal
  means.
- `vignette("lme4-migration", package = "mixeff")` — the verb-for-verb
  map and the generated envelope tables.
- On the package site: the
  [benchmarking article](https://bbuchsbaum.github.io/mixeff/articles/benchmarking.html),
  the
  [inference-method glossary](https://bbuchsbaum.github.io/mixeff/articles/inference-method-glossary.html),
  [reporting workflows](https://bbuchsbaum.github.io/mixeff/articles/reporting-lmms.html),
  and a
  [full published-study reproduction](https://bbuchsbaum.github.io/mixeff/articles/reproducing-aphantasia.html).
