---
title: Why mixeff?
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Why mixeff?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-package}
library(mixeff)
```

If you fit mixed-effects models in R, you most likely use
`lme4::lmer()`. `mixeff::lmm()` aims to be *functionally equivalent*:
the formula language is the same, and the extractors —
`fixef()`, `ranef()`, `VarCorr()`, `predict()`, `simulate()`,
`anova()`, `summary()`, `update()` — do what you expect. Statistical
answers agree with `lme4` within documented tolerances on the parity
datasets shipped with the package. It is not a literal *drop-in*: you
call `lmm()` rather than `lmer()`, results are not bit-exact, and when
the package changes or declines a request it says so, in a diagnostic
you can read from the fitted object.

The reason to switch is what `mixeff` does around the fit. On the
scaling benchmark shipped with the package it ran two to three and a
half times faster than `lme4` at the largest tested scale of each of
five common designs (all small fits in absolute terms; see
`vignette("benchmarking")` for what that does and does not
establish). And it makes four things explicit that `lme4` leaves
implicit:

1.  **The formula stays familiar.** Anything you would hand to
    `lmer()` you can hand to `lmm()`.
2.  **Singular fits become labeled facts.** A reduced-rank
    random-effect covariance is reported with codes, severity, and
    effective rank, not as a single console warning.
3.  **Inference labels its method.** Each p-value carries the method
    that produced it. Where the package declines to compute
    Satterthwaite or Kenward–Roger degrees of freedom — at a boundary
    fit, for example — a parametric bootstrap is offered under the
    same labeling rules.
4.  **The fitted object is a record.** A saved model reopens with the
    same coefficients, the same diagnostics, and the same method
    labels, without depending on the original Rust handle.

This page demonstrates each of the four on one small dataset.

## The dataset

A small repeated-measures study: 18 subjects, 10 daily reaction-time
measurements each, intercepts and slopes that are nearly perfectly
correlated by construction. This design produces a singular fit in
`lme4` and in `mixeff` alike.

```{r make-data}
set.seed(3)
n_subj <- 18L
days   <- 0:9
b0     <- rnorm(n_subj, sd = 30)
b1     <- 0.5 * b0 / 30 * 10 + rnorm(n_subj, sd = 0.5)

sleep_like <- do.call(rbind, lapply(seq_len(n_subj), function(i) {
  data.frame(
    subj = factor(i),
    days = days,
    rt   = 250 + b0[i] + (10 + b1[i]) * days +
           rnorm(length(days), sd = 20)
  )
}))

head(sleep_like, 4)
```

## A. The formula stays familiar

`mixeff` keeps the lme4 random-effects syntax. If you can read `(x | g)`, you
can read `mixeff`.

```{r fit-model}
fit <- lmm(
  rt ~ days + (1 + days | subj),
  sleep_like,
  control = mm_control(verbose = -1)
)
fit
```

```{r familiar-checks, include = FALSE}
stopifnot(inherits(fit, "mm_lmm"))
stopifnot(all(is.finite(fixef(fit))))
stopifnot(length(fitted(fit)) == nrow(sleep_like))
```

`fixef()`, `ranef()`, `VarCorr()`, `predict()`, `simulate()`, `anova()`, and
`summary()` all do what you expect.

## B. When a fit is degenerate, you find out *which* part

`lme4::lmer()` fits this model and reports the situation in one short
line: `boundary (singular) fit`. The fact is correct. What it leaves
implicit is *which* variance component reached the boundary, what the
effective rank of the random-effect covariance is, and which
downstream inference methods the boundary takes off the table.

```{r lme4-side, eval = requireNamespace("lme4", quietly = TRUE)}
m <- suppressMessages(lme4::lmer(
  rt ~ days + (1 + days | subj),
  data = sleep_like
))
m
lme4::isSingular(m)
```

`mixeff` reports the same fact, and then unpacks it. `fit_status()` names
the convergence outcome, `changes()` shows the requested-to-effective
transition, and `diagnostics()` returns stable codes.

```{r mixeff-side}
fit_status(fit)
is_singular(fit)
changes(fit)
diagnostics(fit)$table[, c("code", "severity", "stage", "message")]
```

```{r diag-checks, include = FALSE}
stopifnot(identical(fit_status(fit), "converged_reduced_rank"))
stopifnot(isTRUE(is_singular(fit)))
diag_codes <- diagnostics(fit)$table$code
stopifnot("covariance_reduced" %in% diag_codes)
ch <- changes(fit)$table
stopifnot("design_time_reduction" %in% ch$stage ||
          "covariance_transition" %in% ch$stage)
```

A reduced-rank covariance is now a labeled fact about the fit,
recorded in the object rather than printed once and lost.

## C. Which inference methods survive a boundary fit

On the inference side, `mixeff` first reports which methods (Wald z,
Satterthwaite, Kenward–Roger, bootstrap) are reliable on your fit,
and then lets you run the one you choose under the same labeling
rules.

`inference_options()` enumerates the inference methods for the
current fit, gives each one a
[fixed-vocabulary *reason*](inference-method-glossary.html) for its
status, and names the function you would call to invoke it.

```{r inference-options}
opt <- inference_options(fit, "days", nsim = 200)
opt$table[, c("method", "expected_status",
              "expected_reliability_reason", "current")]
```

```{r options-checks, include = FALSE}
stopifnot(inherits(opt, "mm_inference_options"))
satt_row <- opt$table[opt$table$method == "satterthwaite", ]
stopifnot(identical(satt_row$expected_status, "not_assessed"))
stopifnot(identical(satt_row$expected_reliability_reason,
                    "satterthwaite_unavailable_at_boundary"))
```

Two routes are available on this fit: asymptotic Wald z (immediate, but
labeled `low` reliability), and bootstrap (~seconds, labeled by
replicate count and Monte-Carlo SE). Satterthwaite and Kenward–Roger
refuse with a stable reason — `*_unavailable_at_boundary`. Other
packages will compute degrees of freedom here (`lmerTest` returns a
number on this same fit); `mixeff` declines to, because those
approximations rest on an asymptotic argument about the variance
parameters that does not hold when one of them sits on the boundary.
The refusal is a policy about what to report, stated as a reason code
you can test for.

The asymptotic Wald row carries its own reason. `summary()` prints
`reliability_reason` next to `reliability`:

```{r summary-status}
inf <- inference_table(fit)$table
inf[, c("term", "method", "status", "reliability", "reliability_reason")]
```

```{r summary-status-checks, include = FALSE}
stopifnot("reliability_reason" %in% names(inf))
stopifnot(all(inf$reliability[inf$status == "available"] == "low"))
stopifnot(all(nzchar(inf$reliability_reason)))
# The prose below names this reason code; keep them in step.
stopifnot(all(inf$reliability_reason[inf$status == "available"] ==
                "asymptotic_wald_z_fallback"))
```

`asymptotic_wald_z_fallback` is the engine's reason: a t reference
distribution was the target, the degrees of freedom could not be
computed at this boundary fit, and a standard normal was substituted —
which is why the row is graded `low` rather than hidden.

For a defensible p-value on this same fit, route through `contrast()`
with `method = "bootstrap"`. The Rust engine simulates from the
constrained null, refits each replicate, and returns a labeled
inference row plus a run payload (boundary rate, MCSE, replicate
count) for audit.

```{r bootstrap-answer}
ct <- contrast(fit, c(0, 1), method = "bootstrap",
               bootstrap = bootstrap_control(nsim = 200, seed = 1))
ct$table[, c("contrast", "estimate", "p_value",
             "method", "status", "reliability")]

run <- ct$table$details[[1]]$bootstrap
data.frame(
  successful_replicates = run$successful_replicates,
  boundary_rate         = round(run$boundary_rate, 3),
  mcse                  = round(run$mcse, 4)
)
```

```{r bootstrap-checks, include = FALSE}
stopifnot(identical(ct$table$status, "available"))
stopifnot(is.finite(ct$table$p_value))
stopifnot(run$successful_replicates >= 100)
```

The bootstrap p-value is `available`, the method is named, and the run
payload records how the simulation went. The boundary rate is high
because a singular fit produces singular replicates; the payload
reports that rather than smoothing it over. `mcse` quantifies the
Monte-Carlo uncertainty of the p-value estimate; raise `nsim` for a
tighter MCSE.

Every reported quantity is in one of three states: available, with a
named method and reason; unavailable, with a stable reason code; or a
typed error.

## D. The fit is the record

The fitted object is a serializable record. `saveRDS()` followed by
`readRDS()` and `revive()` reproduces the audit trail and the extractors
without depending on the original Rust handle.

```{r round-trip}
path <- tempfile(fileext = ".rds")
saveRDS(fit, path)
restored <- revive(readRDS(path))

identical(fixef(restored),               fixef(fit))
identical(changes(restored)$table,       changes(fit)$table)
identical(diagnostics(restored)$table,   diagnostics(fit)$table)
```

```{r round-trip-checks, include = FALSE}
stopifnot(identical(fixef(restored), fixef(fit)))
stopifnot(identical(changes(restored)$table, changes(fit)$table))
stopifnot(identical(diagnostics(restored)$table, diagnostics(fit)$table))
```

A reviewer reading the `.rds` later sees the same convergence status,
the same reduced-rank diagnostic, the same method labels on the same
coefficients.

## Current limits

- Estimates are not bit-for-bit identical to `lme4`; the bar is
  statistical agreement within documented tolerances on the parity
  datasets shipped with the package.
- `glmm()`'s default estimator (`pirls_profiled`) returns point
  estimates quickly but refuses Wald standard errors, z statistics,
  and p-values; refit with `method = "joint_laplace"` when you need
  glmer-comparable inference (`vignette("glmm", package = "mixeff")`).
- Profile-likelihood confidence intervals for fixed effects require an
  ML fit; under the REML default they are refused with the reason code
  `profile_beta_unavailable_under_reml` (variance parameters are
  profiled under both criteria).
- `simulate()` is implemented for LMM fits only.

Each limit is reported by the package itself, as a reason code or a
typed error, when you hit it.

## Where to read next

- `vignette("lmm-basics", package = "mixeff")` — fitting and the standard
  extractors at a slower pace.
- `vignette("inference", package = "mixeff")` — coefficient tests,
  contrasts, term tests, and model comparisons.
- `vignette("demystifying-formulas", package = "mixeff")` — what
  `(1 | g)`, `(x | g)`, split blocks, and `||` actually mean.
- `vignette("saving-and-reviving", package = "mixeff")` — saving,
  reloading, and reviving fitted models in detail.
