---
title: "Reproducing the Loo Aphantasia GLMMs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the Loo Aphantasia GLMMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
run_live <- identical(tolower(Sys.getenv("MIXEFF_RUN_APHANTASIA_VIGNETTE")),
                      "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-package}
library(mixeff)
```

The fixture bundled with `mixeff` is an anonymized copy of the trial-level data
used by the revision 3 Loo aphantasia manuscript analysis. It is intended as a
real GLMM reproduction target: large enough to expose optimizer and naming
drift, but small enough to ship as package test data.

```{r load-fixture}
fixture_candidates <- c(
  system.file("extdata", "aphantasia", package = "mixeff"),
  file.path("..", "inst", "extdata", "aphantasia"),
  file.path("inst", "extdata", "aphantasia")
)
fixture_dir <- fixture_candidates[
  dir.exists(fixture_candidates) & nzchar(fixture_candidates)
][1L]
trials <- readRDS(file.path(fixture_dir, "trials.rds"))
metadata <- readRDS(file.path(fixture_dir, "metadata.rds"))
reference <- jsonlite::fromJSON(file.path(fixture_dir, "reference.json"))

c(
  trials = nrow(trials),
  participants = length(unique(trials$participant)),
  metadata_rows = nrow(metadata)
)
```

The primary analysis uses the occluded trials, excludes the four intermediate
VVIQ controls, and models accuracy with crossed participant and item effects.

```{r prepare-data}
excluded <- unlist(reference$excluded_participants, use.names = FALSE)

primary <- subset(
  trials,
  bubbled == "yes" & !is.na(correct) & !participant %in% excluded
)

prepare_model_data <- function(dat, stimtype = FALSE) {
  out <- transform(
    dat,
    participant = factor(participant),
    item = factor(trial_image),
    group = factor(ifelse(aphantasia == "yes", "aphant", "control"),
                   levels = c("control", "aphant")),
    mask = factor(ifelse(back_masked == "yes", "masked", "unmasked"),
                  levels = c("unmasked", "masked")),
    block = factor(block_num),
    soa_log = log(SOA)
  )
  out$soa_s <- as.numeric(scale(out$soa_log))
  if (stimtype) {
    out$stimtype <- factor(
      ifelse(out$bubbled == "yes", "occluded", "intact"),
      levels = c("intact", "occluded")
    )
  }
  out
}

primary_dat <- prepare_model_data(primary)
table(primary_dat$group, primary_dat$mask)
```

The live fits are opt-in because the primary GLMM alone takes several minutes
on a laptop. Set `MIXEFF_RUN_APHANTASIA_VIGNETTE=true` to execute the model
chunks while rendering.

```{r primary-fit, eval = run_live}
primary_fit <- glmm(
  correct ~ group * mask * soa_s + block +
    (1 + mask + soa_s || participant) + (1 | item),
  primary_dat,
  family = binomial(),
  control = mm_control(verbose = -1)
)

c(
  logLik = as.numeric(logLik(primary_fit)),
  AIC = AIC(primary_fit)
)
fixef(primary_fit)
```

Without live refitting, the frozen lme4 reference records the target values the
integration test compares against.

```{r primary-reference}
unlist(reference$models$primary[c("nobs", "logLik", "AIC")])
unlist(reference$models$primary$fixef)
```

The same fixture supports the manuscript sensitivity and specificity fits:
the sensitivity model assigns the four intermediate VVIQ participants to the
control group, the intact model repeats the GLMM on intact images, and the
combined model tests the group by masking cost difference between occluded and
intact stimuli.

```{r additional-glmms, eval = run_live}
sensitivity <- subset(trials, bubbled == "yes" & !is.na(correct))
sensitivity$aphantasia[sensitivity$participant %in% excluded] <- "no"
sensitivity_dat <- prepare_model_data(sensitivity)

intact <- subset(
  trials,
  bubbled == "no" & !is.na(correct) & !participant %in% excluded
)
intact_dat <- prepare_model_data(intact)

combined <- subset(trials, !is.na(correct) & !participant %in% excluded)
combined_dat <- prepare_model_data(combined, stimtype = TRUE)

sensitivity_fit <- glmm(
  correct ~ group * mask * soa_s + block +
    (1 + mask + soa_s || participant) + (1 | item),
  sensitivity_dat,
  family = binomial(),
  control = mm_control(verbose = -1)
)

intact_fit <- glmm(
  correct ~ group * mask * soa_s + block +
    (1 + mask + soa_s || participant) + (1 | item),
  intact_dat,
  family = binomial(),
  control = mm_control(verbose = -1)
)

combined_fit <- glmm(
  correct ~ group * mask * soa_s * stimtype + block +
    (1 + mask + soa_s || participant) + (1 | item),
  combined_dat,
  family = binomial(),
  control = mm_control(verbose = -1)
)

rbind(
  sensitivity = c(logLik = as.numeric(logLik(sensitivity_fit)),
                  AIC = AIC(sensitivity_fit)),
  intact = c(logLik = as.numeric(logLik(intact_fit)), AIC = AIC(intact_fit)),
  combined = c(logLik = as.numeric(logLik(combined_fit)),
               AIC = AIC(combined_fit))
)
```

The RT sensitivity is a Gaussian LMM over correct trials with finite positive
reaction times.

```{r rt-lmm, eval = run_live}
rt_dat <- subset(primary_dat, correct == 1 & is.finite(rt) & rt > 0)
rt_dat$log_rt <- log(rt_dat$rt)

rt_fit <- lmm(
  log_rt ~ group * mask * soa_s + block +
    (1 | participant) + (1 | item),
  rt_dat,
  REML = FALSE,
  control = mm_control(verbose = -1)
)

c(logLik = as.numeric(logLik(rt_fit)), AIC = AIC(rt_fit))
fixef(rt_fit)
```

## Inferential surfaces

Since `mixeff-rs` started serializing the GLMM fixed-effect covariance
artifact (pin 5e72e0b), the inferential surfaces the manuscript actually
reports are available through three `mixeff` primitives. The chunks
below run only when live fitting is enabled
(`MIXEFF_RUN_APHANTASIA_VIGNETTE=true`); otherwise the same calls
remain valid against any locally-built fit.

`summary(fit, tests = "coefficients")` returns a Wald-z fixed-effect
table built from the PIRLS/Laplace working-Hessian covariance:

```{r primary-summary, eval = run_live}
sm <- summary(primary_fit, tests = "coefficients")
round(sm$coefficients, 3)
sm$vcov_status
```

The status block flags `reliability = "moderate"`: the working-Hessian
flavor is close to but not bit-identical with `lme4::vcov(glmer_fit)`.
SE estimates on this dataset drift by ~5-10% in absolute terms, without
flipping any of the manuscript's qualitative conclusions.

The manuscript's primary estimand --- the difference-in-differences
contrast at the centered SOA and at the focal 25 ms SOA --- is a
linear combination of fixed effects, and `mm_lincomb()` is its
front door:

```{r primary-lincomb, eval = run_live}
soa_s_25 <- (log(0.025) - mean(primary_dat$soa_log)) /
              sd(primary_dat$soa_log)

dd_center <- mm_lincomb(
  primary_fit,
  c("group: aphant:mask: masked" = 1)
)
dd_25 <- mm_lincomb(
  primary_fit,
  c("group: aphant:mask: masked"       = 1,
    "group: aphant:mask: masked:soa_s" = soa_s_25)
)
list(centered_soa = dd_center, ms25 = dd_25)
```

Compare directly against the `lme4` reference frozen in the fixture:

```{r dd-vs-reference}
as.data.frame(reference$inference$primary_dd)
```

Both contrasts reproduce the manuscript's sign and significance class:
negative (larger masking cost in aphantasia), CIs excluding zero, and
*p* below the conventional α = .05 at both 25 ms and the centered SOA.

`emmeans` works on `mm_glmm` via `emm_basis.mm_glmm`. Population-level
cell means at the centered SOA, on the response (probability) scale,
recover the same group × mask × group pattern the manuscript reports:

```{r primary-emm, eval = run_live && requireNamespace("emmeans", quietly = TRUE)}
em <- emmeans::emmeans(
  primary_fit, ~ mask | group,
  at   = list(soa_s = 0),
  type = "response"
)
as.data.frame(summary(em))
```

The reverse-pairwise contrast inside each group is the
group-conditional masking cost on the log-odds scale:

```{r primary-emm-contrast, eval = run_live && requireNamespace("emmeans", quietly = TRUE)}
emmeans::contrast(
  emmeans::emmeans(primary_fit, ~ mask | group,
                   at = list(soa_s = 0)),
  method = "revpairwise"
)
```

## Out of scope here

The full sensitivity, intact-stimulus, combined-stimtype, log-RT, S1
random-effects-spec stability, S7 age-covariate, and S9 folder-based
age-matched analyses are reproducible against the same fixture and
reference via `tests/testthat/test-aphantasia-reproduction.R` when
`MIXEFF_RUN_APHANTASIA=true` is set. The S3 leave-one-participant-out
sweep, S4 specification curve, and S5 rstanarm posterior are
intentionally not part of the regular reproduction surface: the first
two are heavy opt-in jobs, and `mixeff` is not a Bayesian engine.

## Caveats

* GLMM Wald inference in `mixeff` is the PIRLS/Laplace working-Hessian
  flavor, advertised as `mm_reliability = "moderate"`. Absolute SEs
  drift 5--10% versus `lme4::vcov()` on this dataset.
* `predict(mm_glmm_fit, ...)` is not yet certified for population-level
  prediction; that lives behind its own bead. Use `emmeans(..., type =
  "response")` for marginal probabilities.
* Coefficient names follow `mixeff`'s `"group: aphant"` scheme, not
  `lme4`'s compact `"groupaphant"`. For copy-pasting hand-written
  lincombs from `lme4` user code, `gsub(": ", "", x, fixed = TRUE)`
  recovers the compact form.

## Citation

If you use the bundled fixture or reproduce these analyses in
downstream work, please cite:

> Loo, C., & Buchsbaum, B. R. (2026). Fragile recurrent processing in
> aphantasia: Evidence from visual pattern completion. *Manuscript
> submitted for publication.*
