---
title: "Your First Mixed Model with mixeff"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Your First Mixed Model with mixeff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-packages}
library(mixeff)
```

```{r simulate-data, include = FALSE}
set.seed(1)
n_subjects <- 18L
sleep <- expand.grid(
  Days = 0:9,
  Subject = factor(seq_len(n_subjects))
)
subject_intercept <- rnorm(n_subjects, sd = 25)
subject_slope <- rnorm(n_subjects, sd = 6)
subject_index <- as.integer(sleep$Subject)
sleep$Reaction <- 250 + 10.5 * sleep$Days +
  subject_intercept[subject_index] +
  subject_slope[subject_index] * sleep$Days +
  rnorm(nrow(sleep), sd = 25)
```

This vignette fits one model, end-to-end, and explains every number. After
reading it you will be able to fit a linear mixed model, read its output,
run an inference test, and save the result. The other vignettes go deeper on
any step.

## The data

The example simulates reaction times (ms) for 18 subjects over ten days of
sleep deprivation. Each subject has a different baseline and a different
response to continued sleep loss. The data are generated inside this
vignette, so the tutorial runs without an example-data package.

```{r load-data}
str(sleep)
```

The key structure: `Subject` appears 10 times in the data — once per day.
Those 10 observations are *not* independent. They share the subject's
baseline speed, their individual sensitivity to sleep loss, and any other
between-person trait that we haven't measured. If you fit a plain OLS
regression, those correlations inflate your Type I error rate and make
standard errors too small.

A mixed model handles this by giving each subject its own intercept *and*
its own slope — their personal baseline and their personal rate of slowing —
and then estimating a population distribution over those person-level
parameters.

```{r individual-trajectories, echo = FALSE, fig.width = 6, fig.height = 4}
subject_lines <- unstack(sleep, Reaction ~ Subject)
matplot(0:9, subject_lines, type = "l", lty = 1, col = "grey70",
        xlab = "Days of sleep deprivation", ylab = "Reaction time (ms)",
        main = "Individual subject trajectories")
abline(lm(Reaction ~ Days, data = sleep), col = "steelblue", lwd = 2)
```

The grey lines are individuals; the blue line is the population average.
Subjects vary in their starting point *and* in how fast they slow down —
which is exactly what a random-intercept-and-slope model captures.

## Step 1: Compile the model

Before fitting, compile the model to see what `mixeff` understands about your
formula. This separates formula interpretation from optimization; you can
catch mis-specified random effects without paying the cost of a full fit.

```{r compile}
spec <- compile_model(Reaction ~ Days + (Days | Subject), sleep)
explain_model(spec)
```

This pre-fit explanation confirms:

- one fixed effect (`Days`) plus an intercept,
- one random-effects block with a correlated intercept and slope per subject,
- the formula has been canonicalized to the explicit `(1 + Days | Subject)`
  form that the optimizer will receive.

Use `audit(spec)` when you want the deeper design audit. Its compact
print starts with the audit summary and requested model;
`print(audit(spec), full = TRUE)` shows the complete upstream report.

If you had written `(1 | Subject)` by mistake (random intercepts only), the
explanation would show only one random-effects column per subject — useful to
verify before a long fit.

## Step 2: Fit

```{r fit}
fit <- lmm(Reaction ~ Days + (Days | Subject), sleep)
```

`lmm()` returns an `mm_lmm` object. The bundled Rust engine performs the
optimization; `lmm()` is the R entry point and result container.

## Step 3: Read the summary

```{r summary}
summary(fit, tests = "coefficients")
```

The summary has four blocks, in print order.

**Fit status.** Check this line first. `converged_interior` is the good
outcome: the optimizer found a clean solution and every variance component
is comfortably positive, so you can read the rest of the summary at face
value. The status changes when something needs your attention. If a
variance component collapses to zero — the data show no detectable
variation for that term — the status reports a *boundary* fit, and the
affected row is flagged in the variance-components table. The p-values
below then switch from Satterthwaite *t* to asymptotic Wald *z*, labeled
as such and graded `low` reliability: the *z* test drops the
finite-sample correction, so treat its p-values as optimistic rather
than cautious.

**Variance components.** `Subject (Intercept)` is the estimated
between-subject spread in baseline reaction time; `Subject Days` is the
estimated between-subject spread in sensitivity to sleep loss; `correlation`
is their estimated association. The residual standard deviation is the
within-subject noise left over.

**Fixed effects.** `Estimate` is the population-level coefficient.
The `Days` estimate is the fitted average change in reaction time per day;
the simulation used a population slope of 10.5 ms per day. The `method`
column names how each p-value was computed — here `satterthwaite`, a
finite-sample *t* test whose `df` column is on the scale of the 18 subjects,
not the 180 raw rows.
Every test statistic and p-value in this table names the method behind it.
On fits where Satterthwaite degrees of freedom cannot be computed (for
example a variance component at the boundary), the summary shows labeled
asymptotic Wald *z* rows instead; the `method` column tells you which one
you got.

**Inference status.** One row per coefficient stating how much to trust the
test: `status` says whether it was computed, `reliability` grades it, and
`reliability_reason` names the engine's reason for that grade — here
`satterthwaite_finite_difference_approximation`, meaning the degrees of
freedom come from a finite-difference approximation, the standard route for
this method (graded `moderate`). Any further engine notes print underneath.
The Rust engine assigns these grades and reasons; the R side passes them
through unchanged.

## Step 4: Extract components

The lme4-compatible extractors work on `mm_lmm` objects:

```{r extractors}
fixef(fit)
```

```{r varcorr}
VarCorr(fit)
```

```{r confint}
confint(fit, method = "asymptotic")
```

`confint()` with `method = "asymptotic"` gives Wald intervals (fast; use
`"bootstrap"` for small samples). The `(Intercept)` interval is the
population mean reaction time on Day 0; the `Days` interval is the
range of plausible slopes.

## Step 5: Test a specific claim

`contrast()` tests a linear combination of fixed-effect coefficients and
returns a one-row table that names the method behind the p-value.

```{r contrast}
ct <- contrast(fit, c("(Intercept)" = 0, Days = 1))
ct$table[, c("estimate", "std_error", "p_value", "method", "status", "reliability")]
```

The `estimate` is the Days slope. `status = "available"` says the test was
computed; `reliability` grades it on a fixed scale — `high`, `moderate`,
`low`, or `not_available` — and here reads `moderate`, the standard grade
for finite-difference Satterthwaite. If you request a method the fit cannot
support (Satterthwaite on a boundary fit, say), the row comes back with
`status = "not_assessed"` and a `reason_code` naming the problem instead
of a number.

## Step 6: Compare specific conditions

`mm_comparisons()` answers questions like "how much does the population
mean slow between Day 0 and Day 9?" directly, so you do not assemble the
contrast by hand.

```{r comparisons}
cmp <- mm_comparisons(fit, specs = "Days", at = list(Days = c(0, 9)))
cmp$table[, c("label", "estimate", "conf_low", "conf_high", "p_value")]
```

The `at` argument pins `Days` to exactly those two values. The difference
`Days=9 - Days=0` is nine times the fitted slope and includes a confidence
interval.

## Step 7: Save and reload

Mixed-model fits can take minutes to hours on large data sets. `mixeff` stores
everything needed to revive the result later.

```{r save-reload}
tmp <- tempfile(fileext = ".rds")
saveRDS(fit, tmp)

fit2 <- revive(readRDS(tmp))
stopifnot(isTRUE(all.equal(fixef(fit), fixef(fit2))))
```

Everything the extractors and inference functions need is stored in the R
object itself, so `readRDS()` alone is enough: `fixef()`, `contrast()`,
`mm_means()`, and the reporting tables all work on the reloaded fit. The
Rust handle is deliberately left absent after reloading. `revive()` resets
the object's process-local cache; calling it after `readRDS()` is cheap
and recommended, but the current bridge does not require it. See
`vignette("saving-and-reviving")` for the full discussion.

## Where to go next

| Question | Vignette |
|---|---|
| What do the random-effects formula options mean? | `vignette("demystifying-formulas")` |
| How are p-values computed and which method is right for my fit? | `vignette("inference")` |
| How do I compute marginal means and treatment contrasts? | `vignette("marginal-effects")` |
| How do I write up the results for a paper? | `vignette("reporting-lmms")` |
| I'm coming from lme4 — what's different? | `vignette("lme4-migration")` |
| Can I fit a binomial/Poisson GLMM? | `vignette("glmm")` |
