---
title: Benchmarking mixeff
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking mixeff}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-package}
library(mixeff)
```

Speed matters most when a workflow repeats the same fit many times.
Parametric bootstrap is the canonical case: a 1,000-replicate
bootstrap means roughly 1,000 refits, plus the original fit and
summary work. The same logic applies to cluster bootstrap,
simulation studies, multi-seed reanalysis, and anything whose inner
loop is "fit the same model again with new data". A per-refit
speedup multiplies by the replicate count.

On the LMM scaling benchmark shipped with the package, `mixeff` ran
2.0 to 3.5 times faster than `lme4` at the largest tested scale of
each design (1.3 to 3.5 across the whole grid). This
vignette shows the figures from a single run of that benchmark and
then documents the harness used to make them. The numbers are
machine- and model-dependent — all of these fits complete in tens of
milliseconds — so the scripts also write CSV files that you can
re-plot and rerun on the designs you actually care about.

## What does the benchmark show?

The benchmark sweeps five mixed-model designs — varying rows,
varying subject levels, adding random slopes, crossing subject and
item, and combining crossed grouping with a random slope — across a
log-spaced grid of scale values. At each cell, `mixeff::lmm()` and
`lme4::lmer()` fit the same model five times. The summary table
records median elapsed seconds and speedup over `lme4` per cell.
The shipped CSV comes from one run on 2026-07-30: macOS arm64
(M-series), R 4.5.1, mixeff 0.2.0 built with the release profile,
engine snapshot `4a2abb3`. If any of those differ on your machine,
rerun the harness rather than reusing these numbers.

```{r benchmark-data}
benchmark_path <- system.file(
  "extdata", "lme4-scaling-summary.csv",
  package = "mixeff"
)
benchmark <- read.csv(benchmark_path, stringsAsFactors = FALSE)

benchmark$scenario_label <- factor(
  benchmark$scenario,
  levels = c("rows", "groups", "slopes", "crossed", "crossed_slope"),
  labels = c(
    "vary rows (1 | subject)",
    "vary subject levels (1 | subject)",
    "add random slope (1 + x | subject)",
    "crossed (1 | subject) + (1 | item)",
    "crossed + slope (1 + x | subject) + (1 | item)"
  )
)

head(benchmark[, c("scenario", "scale_value", "engine",
                   "median_sec", "speedup_vs_lme4")])
```

```{r benchmark-checks, include = FALSE}
stopifnot(nrow(benchmark) > 0L)
stopifnot(all(c("rows", "groups", "slopes", "crossed",
                "crossed_slope") %in% benchmark$scenario))
stopifnot(all(c("lme4", "mixeff") %in% benchmark$engine))
# Derive the prose numbers from the shipped CSV (two significant digits).
mixeff_cells <- benchmark[benchmark$engine == "mixeff", ]
stopifnot(all(mixeff_cells$speedup_vs_lme4 > 1))  # faster in every cell
largest_scale <- do.call(rbind, lapply(
  split(mixeff_cells, mixeff_cells$scenario),
  function(d) d[which.max(d$scale_value), , drop = FALSE]
))
largest_speedup <- signif(largest_scale$speedup_vs_lme4, 2)
stopifnot(min(largest_speedup) == 2.0, max(largest_speedup) == 3.5)
stopifnot(signif(min(mixeff_cells$speedup_vs_lme4), 2) == 1.3)
stopifnot(signif(max(mixeff_cells$speedup_vs_lme4), 2) == 3.5)
stopifnot(identical(
  mixeff_cells$scenario[which.max(mixeff_cells$speedup_vs_lme4)],
  "crossed_slope"
))
```

The first figure plots the median fit time against design scale, on
log-log axes, with one panel per scenario. `mixeff` runs below
`lme4` everywhere on this grid. The largest speedup is on the one
design that combines crossed grouping with a random slope; the other
four designs do not order by structural complexity (the plain
grouped random intercept ranks second).

```{r fit-time-plot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 4.5, fig.alt = "Median fit time vs design scale, log-log, by scenario, for mixeff and lme4."}
library(ggplot2)

ggplot(benchmark,
       aes(scale_value, median_sec * 1000, colour = engine,
           group = engine)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ scenario_label, scales = "free_x", ncol = 2) +
  scale_x_log10() +
  scale_y_log10() +
  scale_colour_manual(values = c(lme4 = "#888888", mixeff = "#3366cc")) +
  labs(
    x = "design scale (rows, subject levels, etc.)",
    y = "median fit time (ms, log scale)",
    colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top",
        panel.grid.minor = element_blank())
```

The second figure folds the same data into one panel: the speedup
of `mixeff` over `lme4` at each cell. The dashed reference line at
1× would mean the two engines are equally fast.

```{r speedup-plot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 4, fig.alt = "Speedup of mixeff over lme4 vs design scale, by scenario."}
mixeff_rows <- subset(benchmark, engine == "mixeff")

ggplot(mixeff_rows,
       aes(scale_value, speedup_vs_lme4,
           colour = scenario_label, group = scenario_label)) +
  geom_hline(yintercept = 1, linetype = 2, colour = "grey50") +
  geom_line() +
  geom_point() +
  scale_x_log10() +
  scale_y_continuous(breaks = seq(0, 8, by = 1)) +
  labs(
    x = "design scale",
    y = "speedup over lme4 (×)",
    colour = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom",
        legend.direction = "vertical",
        panel.grid.minor = element_blank()) +
  guides(colour = guide_legend(ncol = 1))
```

Two caveats before reading the speedup figure. First, the absolute
timings run from about 3 ms to 35 ms and the recorded medians are
whole milliseconds, so timer resolution alone can move a ratio
substantially — one millisecond on the 4 ms side of a 12 ms / 4 ms
cell moves the ratio from 3.0 to 2.4 or 4.0; five replications per
cell do little to smooth that. Second, within a single design the
speedup does not trend reliably with scale. What the grid does
support: `mixeff` ran faster in every cell, and the single largest
speedup is on the design combining crossed grouping with a random
slope.

## What should you measure?

There are two separate speed questions.

First, measure model fitting. This asks how quickly each engine solves the same
mixed model as the number of rows, groups, slopes, or crossed random effects
changes.

Second, measure bootstrap routes. This asks how long the user-facing inference
verbs take when they repeatedly refit models.

```{r refit-counts}
data.frame(
  bootstrap_replicates = c(50L, 200L, 1000L),
  approximate_refits = c(50L, 200L, 1000L)
)
```

The arithmetic is simple but it decides the workflow: per-refit speed
differences multiply by `nsim`.

## How do you benchmark fitting?

Use the scaling benchmark when you want to compare `mixeff::lmm()` with
`lme4::lmer()` over several synthetic LMM designs.

```{r fit-benchmark-command, eval = FALSE}
system2(
  "Rscript",
  c(
    "inst/benchmarks/lme4-scaling.R",
    "--reps=5",
    "--scenarios=rows,groups,slopes,crossed",
    "--no-plots"
  )
)
```

That script writes one row per engine, design, scale value, and timing
repetition, plus a summary table with median elapsed seconds and speed relative
to `lme4`.

```{r fit-benchmark-files}
data.frame(
  file = c(
    "benchmarks/lme4-scaling/lme4-scaling-raw.csv",
    "benchmarks/lme4-scaling/lme4-scaling-summary.csv"
  ),
  contents = c(
    "one timing row per engine/design/repetition",
    "median seconds, fits per second, and relative speed"
  )
)
```

## How do you benchmark bootstrap?

Use the bootstrap benchmark when you want to time the bootstrap and
simulation-based inference routes themselves.

```{r bootstrap-benchmark-command, eval = FALSE}
system2(
  "Rscript",
  c(
    "inst/benchmarks/bootstrap-inference.R",
    "--nsim=200",
    "--reps=3"
  )
)
```

The benchmark includes `test_effect(method = "bootstrap")`,
`confint(method = "bootstrap")`, `compare(method = "bootstrap")`, and
`lme4::bootMer()` as an R-engine baseline. If `pbkrtest` is installed, it also
times `pbkrtest::PBmodcomp()`.

```{r bootstrap-routes}
data.frame(
  route = c(
    "mixeff_test_effect_bootstrap",
    "mixeff_confint_bootstrap",
    "mixeff_compare_bootstrap_lrt",
    "lme4_bootMer_fixef_distribution",
    "pbkrtest_PBmodcomp"
  ),
  target = c(
    "term-level fixed-effect p-value",
    "fixed-effect confidence interval",
    "model comparison p-value",
    "fixed-effect bootstrap distribution",
    "parametric bootstrap model comparison"
  )
)
```

## What does inference method choice cost?

Speed is only one part of the inference decision. The route table also needs
calibration evidence: how often a method rejects under a null, how often it
detects a small alternative, and whether interval routes cover that small
alternative.

The shipped artifact is produced by the simulation study's **release mode**:
2,500 replications per analytic route (Monte Carlo SE at a true 0.05 rate:
about 0.0044), 500 per bootstrap route (about 0.0097), and 1,000 for the
certified joint-Laplace GLMM route. Every rate below carries a Wilson 95%
interval, and every cell records requested, evaluated, refused, failed, and
boundary replicate counts — nothing is silently dropped. The study covers
four Gaussian LMM regimes (interior, boundary, reduced-rank, small-group)
and one binomial GLMM fixture with three routes (certified joint Wald, the
opt-in working-Hessian approximation, and the parametric bootstrap). That is
the scope of the claim: calibration evidence **for these regimes**, not a
general guarantee — a broader condition matrix (crossed effects, rare
events, overdispersion) is future work.

The artifact's provenance (package commit, engine pin, lockfile checksum,
exact invocation, seeds) ships next to it as a manifest:

```{r inference-simulation-data}
locate_extdata <- function(file) {
  path <- system.file("extdata", file, package = "mixeff")
  if (nzchar(path)) return(path)
  local_path <- file.path("inst", "extdata", file)
  if (file.exists(local_path)) local_path else file.path("..", "..", local_path)
}
simulation <- read.csv(locate_extdata("inference-method-simulation-summary.csv"),
                       stringsAsFactors = FALSE)
manifest <- jsonlite::read_json(
  locate_extdata("inference-method-simulation-manifest.json")
)
cat(sprintf(
  "mode: %s | analytic reps: %d | bootstrap reps: %d | joint reps: %d\npackage %s @ %.9s | engine %.9s\nreproduce: %s\n",
  manifest$mode, manifest$reps_analytic, manifest$reps_bootstrap,
  manifest$reps_joint, manifest$package_version, manifest$package_commit,
  manifest$rust_pin, manifest$invocation
))
```

```{r inference-simulation-checks, include = FALSE}
# The vignette refuses to present smoke-mode output as calibration evidence,
# and the shipped artifact must meet the release-mode replication floor.
stopifnot(identical(manifest$mode, "release"))
stopifnot(manifest$reps_analytic >= 2500L, manifest$reps_bootstrap >= 500L)
stopifnot(all(c("method", "fixture", "mode", "n_requested",
                "n_evaluated_null", "n_refused_null", "n_boundary_null",
                "type_I_error", "type_I_lower", "type_I_upper",
                "coverage_at_alt", "coverage_lower", "coverage_upper",
                "reason") %in% names(simulation)))
stopifnot(all(simulation$mode == "release"))
ran <- simulation[simulation$n_requested > 0, ]
stopifnot(nrow(ran) > 0L)
# Accounting closes: evaluated + refused + failed + fit_failed == requested.
stopifnot(all(ran$n_evaluated_null + ran$n_refused_null +
                ran$n_failed_null + ran$n_fit_failed_null == ran$n_requested))
```

The plot shows each route's empirical rates with their Wilson intervals.
P-value routes contribute Type I error and power; interval routes contribute
coverage. Rates are conditional on a successful evaluation; the CSV also
carries unconditional versions (refusals and failures counted against the
route) and the full replicate accounting. Routes that did not run carry an
explicit `reason` — "excluded in this mode" is machine-distinguishable from
"not wired".

```{r inference-simulation-plot, eval = requireNamespace("ggplot2", quietly = TRUE), fig.width = 8, fig.height = 6, fig.alt = "Inference calibration: Type I error, power, and coverage with Wilson intervals, by method and fixture."}
library(ggplot2)

metric_frame <- function(d, value, lower, upper, label) {
  out <- d[, c("method", "fixture", value, lower, upper)]
  names(out) <- c("method", "fixture", "value", "lower", "upper")
  out$metric <- label
  out
}
simulation_long <- rbind(
  metric_frame(ran, "type_I_error", "type_I_lower", "type_I_upper",
               "Type I error"),
  metric_frame(ran, "power_at_alt", "power_lower", "power_upper",
               "Power at small alternative"),
  metric_frame(ran, "coverage_at_alt", "coverage_lower", "coverage_upper",
               "Coverage at small alternative")
)
simulation_long <- subset(simulation_long, is.finite(value))
simulation_long$metric <- factor(
  simulation_long$metric,
  levels = c("Type I error", "Power at small alternative",
             "Coverage at small alternative")
)
nominal <- data.frame(
  metric = factor(c("Type I error", "Coverage at small alternative"),
                  levels = levels(simulation_long$metric)),
  y = c(0.05, 0.95)
)

ggplot(simulation_long, aes(method, value)) +
  geom_hline(data = nominal, aes(yintercept = y), linetype = 2,
             linewidth = 0.3, colour = "grey40") +
  geom_pointrange(aes(ymin = lower, ymax = upper), linewidth = 0.4,
                  fatten = 1.5) +
  facet_grid(metric ~ fixture) +
  coord_flip(ylim = c(0, 1)) +
  labs(x = NULL, y = "empirical proportion (Wilson 95% interval)") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())
```

## How do you read the result?

For a local speed claim, read the summary CSV and compute ratios from elapsed
seconds. Values above 1 mean the baseline route took longer than the `mixeff`
route.

```{r speedup-example, eval = FALSE}
summary <- read.csv("benchmarks/bootstrap-inference/bootstrap-inference-summary.csv")
mixeff_ci <- subset(summary, route == "mixeff_confint_bootstrap")
lme4_boot <- subset(summary, route == "lme4_bootMer_fixef_distribution")

data.frame(
  comparison = "lme4 bootMer / mixeff bootstrap CI",
  speedup = lme4_boot$median_sec / mixeff_ci$median_sec
)
```

Treat that ratio as local evidence, not a universal constant. It depends on the
formula, optimizer, random-effects structure, `nsim`, and machine. The point of
the benchmark scripts is to make the claim reproducible for the model class you
are actually fitting.

## What should go in a report?

When you report benchmark results, include the model formula, number of rows,
number of grouping levels, random-effects structure, `nsim`, repetitions, and
the package versions. The raw CSV has the timing rows; the summary CSV is for
tables and figures.

```{r reporting-fields}
data.frame(
  field = c("formula", "nobs", "groups", "random_terms", "nsim", "reps", "versions"),
  reason = c(
    "defines the fixed and random structure being refit",
    "controls the amount of data processed per fit",
    "controls the random-effect dimension",
    "controls covariance structure and optimizer work",
    "controls bootstrap cost",
    "shows timing stability",
    "makes the run reproducible"
  )
)
```
