---
title: "Reproducing the Loo Aphantasia GLMMs"
output: albersdown::albers_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the Loo Aphantasia GLMMs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
run_live <- identical(tolower(Sys.getenv("MIXEFF_RUN_APHANTASIA_VIGNETTE")),
                      "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r load-package}
library(mixeff)
```

The fixture bundled with `mixeff` is an anonymized copy of the trial-level data
used by the revision 3 Loo aphantasia manuscript analysis. It is intended as a
real GLMM reproduction target: large enough to expose optimizer and naming
drift, but small enough to ship as package test data.

```{r load-fixture}
fixture_candidates <- c(
  system.file("extdata", "aphantasia", package = "mixeff"),
  file.path("..", "..", "inst", "extdata", "aphantasia"),
  file.path("inst", "extdata", "aphantasia")
)
fixture_dir <- fixture_candidates[
  dir.exists(fixture_candidates) & nzchar(fixture_candidates)
][1L]
trials <- readRDS(file.path(fixture_dir, "trials.rds"))
metadata <- readRDS(file.path(fixture_dir, "metadata.rds"))
reference <- jsonlite::fromJSON(file.path(fixture_dir, "reference.json"))

c(
  trials = nrow(trials),
  participants = length(unique(trials$participant)),
  metadata_rows = nrow(metadata)
)
```

The primary analysis uses the occluded trials, excludes the four intermediate
VVIQ controls, and models accuracy with crossed participant and item effects.

```{r prepare-data}
excluded <- unlist(reference$excluded_participants, use.names = FALSE)

primary <- subset(
  trials,
  bubbled == "yes" & !is.na(correct) & !participant %in% excluded
)

prepare_model_data <- function(dat, stimtype = FALSE) {
  out <- transform(
    dat,
    participant = factor(participant),
    item = factor(trial_image),
    group = factor(ifelse(aphantasia == "yes", "aphant", "control"),
                   levels = c("control", "aphant")),
    mask = factor(ifelse(back_masked == "yes", "masked", "unmasked"),
                  levels = c("unmasked", "masked")),
    block = factor(block_num),
    soa_log = log(SOA)
  )
  out$soa_s <- as.numeric(scale(out$soa_log))
  if (stimtype) {
    out$stimtype <- factor(
      ifelse(out$bubbled == "yes", "occluded", "intact"),
      levels = c("intact", "occluded")
    )
  }
  out
}

primary_dat <- prepare_model_data(primary)
table(primary_dat$group, primary_dat$mask)
```

The live fits are opt-in because the primary GLMM alone takes several minutes
on a laptop. Set `MIXEFF_RUN_APHANTASIA_VIGNETTE=true` to execute the model
chunks while rendering. The test suite uses the same principle: the core
reproduction runs under `MIXEFF_RUN_APHANTASIA=true`, and the slower S1
random-effects stability variants run under `MIXEFF_RUN_APHANTASIA_STRESS=true`.

```{r primary-fit, eval = run_live}
# inference = "working_hessian": this vignette DELIBERATELY opts in to the
# uncertified working-Hessian approximation, because its purpose below is to
# compare the complete fitted workflow against the frozen glmer reference.
# The random-effect families also differ (see below), so this comparison
# does not isolate the covariance approximation. Default fits withhold Wald inference.
primary_fit <- glmm(
  correct ~ group * mask * soa_s + block +
    (1 + mask + soa_s || participant) + (1 | item),
  primary_dat,
  family = binomial(),
  inference = "working_hessian",
  control = mm_control(verbose = -1)
)

c(
  logLik = as.numeric(logLik(primary_fit)),
  AIC = AIC(primary_fit)
)
fixef(primary_fit)
```

Without live refitting, the frozen lme4 reference records the target values the
integration test compares against.

```{r primary-reference}
unlist(reference$models$primary[c("nobs", "logLik", "AIC")])
unlist(reference$models$primary$fixef)
```

The same fixture supports the manuscript sensitivity and specificity fits:
the sensitivity model assigns the four intermediate VVIQ participants to the
control group. The estimator matters for two of the models that follow.
The intact high-baseline Bernoulli model reaches near-exact lme4
fixed-effect and log-likelihood parity only under
`method = "joint_laplace"`, which is how the opt-in reproduction test fits
it (about 40 seconds per fit on a release build). On the default profiled
path its coefficients drift well past the strict tolerances, so the chunks
below do not claim parity for it. The combined model is fitted by the
default profiled path here and in the test suite; like every profiled GLMM
fit, it carries the engine's support note that a fast-PIRLS fit is not
certified as glmer joint-Laplace parity.

```{r additional-glmms, eval = run_live}
sensitivity <- subset(trials, bubbled == "yes" & !is.na(correct))
sensitivity$aphantasia[sensitivity$participant %in% excluded] <- "no"
sensitivity_dat <- prepare_model_data(sensitivity)

intact <- subset(
  trials,
  bubbled == "no" & !is.na(correct) & !participant %in% excluded
)
intact_dat <- prepare_model_data(intact)

combined <- subset(trials, !is.na(correct) & !participant %in% excluded)
combined_dat <- prepare_model_data(combined, stimtype = TRUE)

sensitivity_fit <- glmm(
  correct ~ group * mask * soa_s + block +
    (1 + mask + soa_s || participant) + (1 | item),
  sensitivity_dat,
  family = binomial(),
  control = mm_control(verbose = -1)
)

intact_fit <- glmm(
  correct ~ group * mask * soa_s + block +
    (1 + mask + soa_s || participant) + (1 | item),
  intact_dat,
  family = binomial(),
  control = mm_control(verbose = -1)
)

combined_fit <- glmm(
  correct ~ group * mask * soa_s * stimtype + block +
    (1 + mask + soa_s || participant) + (1 | item),
  combined_dat,
  family = binomial(),
  control = mm_control(verbose = -1)
)

rbind(
  sensitivity = c(logLik = as.numeric(logLik(sensitivity_fit)),
                  AIC = AIC(sensitivity_fit)),
  intact = c(logLik = as.numeric(logLik(intact_fit)), AIC = AIC(intact_fit)),
  combined = c(logLik = as.numeric(logLik(combined_fit)),
               AIC = AIC(combined_fit))
)
```

The RT sensitivity is a Gaussian LMM over correct trials with finite positive
reaction times. Re-standardize log SOA within that subset, as in the manuscript.

```{r rt-lmm, eval = run_live}
rt_dat <- subset(primary_dat, correct == 1 & is.finite(rt) & rt > 0)
rt_dat$log_rt <- log(rt_dat$rt)
rt_dat$soa_s <- as.numeric(scale(rt_dat$soa_log))

rt_fit <- lmm(
  log_rt ~ group * mask * soa_s + block +
    (1 | participant) + (1 | item),
  rt_dat,
  REML = FALSE,
  control = mm_control(verbose = -1)
)

c(logLik = as.numeric(logLik(rt_fit)), AIC = AIC(rt_fit))
fixef(rt_fit)
```

## What the manuscript reports, computed three ways

Since the engine began serializing the GLMM fixed-effect covariance
artifact, the quantities the manuscript actually reports are available
through three `mixeff` functions. The chunks
below run only when live fitting is enabled
(`MIXEFF_RUN_APHANTASIA_VIGNETTE=true`); otherwise the same calls
remain valid against any locally-built fit.

This comparison changes more than the covariance approximation. With a factor
inside `||`, mixeff fits independent treatment-coded random coefficients;
lme4's expansion retains a correlated factor block. Thus the reference and
the fit above use different random-effect families. An isolated assessment of
the working Hessian requires an explicitly aligned family and estimator.

This section illustrates the working-Hessian opt-in in that broader workflow.
The fit above passes `inference = "working_hessian"` — the explicit
opt-in that unlocks the UNCERTIFIED profiled working-Hessian approximation.
Without it, a default profiled fit withholds Wald standard errors, tests, and
intervals. Request `method = "joint_laplace"` and inspect its covariance and
convergence evidence before using Wald inference, or use
`confint(fit, method = "bootstrap")` for refit-based intervals.
With the opt-in, `summary(fit, tests = "coefficients")` returns a Wald-z
table in which every row is labelled `wald_z_working_hessian`:

```{r primary-summary, eval = run_live}
sm <- summary(primary_fit, tests = "coefficients")
round(sm$coefficients[, c("Estimate", "Std. Error", "z value", "Pr(>|z|)")], 3)
sm$vcov_status
```

The rows carry `reliability = "moderate"` and status
`available_noninferential`. The grade is a rubric label, not a claim of
calibrated inference. The parity ledger records SEs about 11% smaller than
the reference in a historical run, with a 15% tolerance. That is a difference
between complete workflows, not evidence that the working-Hessian
approximation alone causes an 11% shortfall. The historical decomposition and
its engine pin are described in the caveats below.

A separate comparison at engine pin `977ef11` aligned all 17,280 primary rows,
Bernoulli/logit likelihood, `nAGQ = 1`, factor coding, and the explicit six-theta
random-effect family in both packages:

```{r aligned-family, eval = FALSE}
correct ~ group * mask * soa_s + block +
  (1 | participant) + (0 + mask | participant) +
  (0 + soa_s | participant) + (1 | item)
```

With `method = "joint_laplace"`, the largest fixed-effect difference from a
fresh `lme4::glmer()` fit was 0.000055; the relative log-likelihood difference
was 0.000000080. Both used 15 model parameters. These are point-estimate
comparisons: the `mixeff` result was `not_optimized`, with stationarity residual
1.083 against a tolerance of 0.02. Wald SEs and tests were withheld. This
expanded formula also contains a covariance redundancy: its two mask dummy
columns already span the participant intercept. Their unrestricted covariance
and a separate intercept variance cannot be uniquely separated. A full
positive-definite Hessian is therefore not a suitable expectation for this
parameterization; this run cannot isolate a Hessian-computation defect.

A second comparison replaced only the random mask term with
`(0 + mask_masked | participant)`, where
`mask_masked <- as.numeric(mask == "masked")`. This gives both packages the
identifiable four-variance family used by the native diagonal model. The
largest coefficient difference was 0.000058. Across the nine coefficients,
joint-Laplace SEs differed from `glmer` by at most 1.06%, and the profiled
working-Hessian SEs by at most 0.13%. The joint covariance passed its own
positive-definiteness check, but optimization was still `not_optimized`
(stationarity residual 0.119 against 0.02). These comparisons separate model
families and expose the remaining diagnostic limits; they do not establish
calibration or recompute the historical SE decomposition.

The manuscript's primary estimand --- the difference-in-differences
contrast at the centered SOA and at the focal 25 ms SOA --- is a
linear combination of fixed effects, which `mm_lincomb()` computes:

```{r primary-lincomb, eval = run_live}
soa_s_25 <- (log(0.025) - mean(primary_dat$soa_log)) /
              sd(primary_dat$soa_log)

dd_center <- mm_lincomb(
  primary_fit,
  c("groupaphant:maskmasked" = 1)
)
dd_25 <- mm_lincomb(
  primary_fit,
  c("groupaphant:maskmasked"       = 1,
    "groupaphant:maskmasked:soa_s" = soa_s_25)
)
list(centered_soa = dd_center, ms25 = dd_25)
```

Compare directly against the `lme4` reference frozen in the fixture:

```{r dd-vs-reference}
as.data.frame(reference$inference$primary_dd)
```

Both contrasts come out negative (larger masking cost in aphantasia), as
in the manuscript. Be precise about what is pinned. The opt-in test suite
enforces agreement with the frozen reference at ±0.02 on the estimates
and ±0.01 on the SEs; at the worst corner of those bands the centered-SOA
p-value — .039 in the reference — could reach .106, so the significance
verdict there is not guaranteed by the tests. The historical ledger reports
smaller SEs, but neither those tolerance bands nor the opt-in's reliability
grade establish calibration. The sign and approximate size are the targets
of this reference comparison.

`emmeans` works on `mm_glmm` via `emm_basis.mm_glmm`. Population-level
cell means at the centered SOA, on the response (probability) scale,
recover the same group × mask pattern the manuscript reports:

```{r primary-emm, eval = run_live && requireNamespace("emmeans", quietly = TRUE)}
em <- emmeans::emmeans(
  primary_fit, ~ mask | group,
  at   = list(soa_s = 0),
  type = "response"
)
as.data.frame(summary(em))
```

The reverse-pairwise contrast inside each group is the
group-conditional masking cost on the log-odds scale:

```{r primary-emm-contrast, eval = run_live && requireNamespace("emmeans", quietly = TRUE)}
emmeans::contrast(
  emmeans::emmeans(primary_fit, ~ mask | group,
                   at = list(soa_s = 0)),
  method = "revpairwise"
)
```

## Out of scope here

The full sensitivity, intact-stimulus, combined-stimtype, log-RT, S1
random-effects-spec stability, S7 age-covariate, and S9 folder-based
age-matched analyses are reproducible against the same fixture and
reference via `tests/testthat/test-aphantasia-reproduction.R` when
`MIXEFF_RUN_APHANTASIA=true` is set. The S3 leave-one-participant-out
sweep, S4 specification curve, and S5 rstanarm posterior are
intentionally left out of the regular reproduction run: the first
two are heavy opt-in jobs, and `mixeff` is not a Bayesian engine.

## Caveats

* The Wald numbers in this article come from the PIRLS/Laplace
  working-Hessian covariance via the explicit `inference =
  "working_hessian"` opt-in. Default profiled fits refuse Wald inference;
  joint Laplace assesses covariance separately, and parametric bootstrap
  produces intervals by simulation and refitting.
  The historical SE gap is about 11% (ledger tolerance 15%). A
  hold-the-point decomposition at engine pin `6731062` attributed roughly
  73% of that gap to the random-effect family, 26% to working-Hessian scaling,
  and 1% to optimizer drift. These are historical measurements, not results
  regenerated at the current engine pin. They are recorded in
  `inst/extdata/expected-mismatches.json`, case `aphantasia_primary`, field
  `inference.primary_dd.SE`. The smaller reference-relative SEs alone do not
  measure Type I error or coverage.
* Population-level GLMM prediction (`re.form = NA`, `type = "link"` or
  `"response"`) is supported and matches `predict(glmer, re.form = NA)`
  on joint-Laplace fits; `emmeans(..., type = "response")` remains the
  route for averaged marginal probabilities.
* Coefficient names are lme4-identical since 0.2.0; see
  `vignette("lme4-migration")`.

## Citation

If you use the bundled fixture or reproduce these analyses in
downstream work, please cite:

> Loo, C., & Buchsbaum, B. R. (2026). Fragile recurrent processing in
> aphantasia: Evidence from visual pattern completion. *Manuscript
> submitted for publication.*
